### 2025-05-28 paper/main.tex
Add paper in greek

\documentclass[12pt]{extarticle}

\input{preamble}

% Basic Document Info

% \title{Διάγνωση ασθενειών με χρήση \en{ASP}}
\title{Αυτόματη διάγνωση ασθενειών με χρήση Μεγάλων Γλωσσικών Μοντέλων και Answer Set Programming}
% \date{\today}
\author{Ιωάννα Γέμου\thanks{Τμήμα Ηλεκτρολόγων Μηχανικών και Τεχνολογίας Υπολογιστών, Πανεπιστήμιο Πατρών} \\ \src{ioannagemou@upnet.gr} \and Ευάγγελος Λάμπρου\footnotemark[1] \\ \src{e.lamprou@upnet.gr}}
% \institute{Πανεπιστήμιο Πατρών \\ Τμήμα Ηλεκτρολόγων Μηχανικών και % Τεχνολογίας Υπολογιστών} % Institute or school name

\begin{document}

\maketitle
\tableofcontents
\newpage
\section{Περίληψη}
Η ανάπτυξη αυτόματων συστημάτων πρόβλεψης ασθενειών έχει κρίσιμη σημασία για τη βελτίωση της ανθρώπινης υγείας και την υποστήριξη των ιατρών στην ακριβή διάγνωση και αντιμετώπιση των ασθενειών. Η ακριβής διάγνωση είναι κρίσιμη για την έγκαιρη αντιμετώπιση και αποτελεσματική θεραπεία, ενώ μπορεί επίσης να συμβάλει στην αποφυγή περαιτέρω επιπλοκών και επιβαρυντικών ιατρικών διαδικασιών. 
Η συμβολική τεχνητή νοημοσύνη έχει ξαναεφαρμοστεί στο πεδίο της ιατρικής \cite{Alviano_2020}. Ωστόσο,
εφαρμογή πάντα ήταν περιορισμένη λόγω των σύνθετων βάσεων γνώσης που θα έπρεπε να αναπτυχθούν για τη λειτουργία 
αυτών των συστημάτων.
Η προτεινόμενη μέθοδος, η οποία συνδυάζει τα Large Language Models και το Answer Set Programming (ASP), αποτελεί ένα ισχυρό εργαλείο, καθώς παρέχει ένα αξιόπιστο και γρήγορο μέσο δημιουργίας μιας βάσης γνώσης η οποία 
στη συνέχεια μπορεί να χρησιμοποιηθεί για την πρόβλεψη ασθενειών.

% Η αυτόματη πρόβλεψη ασθενειών βασίζεται στην επεξεργασία της ιατρικής βιβλιογραφίας, που περιέχει μεγάλο όγκο πληροφοριών για διάφορες ασθένειες και τα συμπτώματά τους, καθώς και τις βιολογικές ενδείξεις που σχετίζονται με αυτές. Μέσω της χρήσης των Large Language Models, καταφέρνουμε να μετατρέψουμε αυτήν τη γνώση σε κώδικα, χρησιμοποιώντας τη γλώσσα προγραμματισμού Clingo. Ουσιαστικά, τα μοντέλα γλώσσας μετατρέπουν τις πληροφορίες αυτές σε μια μορφή που μπορεί να επεξεργαστεί από το πρόγραμμα Answer Set Programming.

% Το πρόγραμμά μας λειτουργεί με τον εξής τρόπο: δέχεται ως είσοδο συγκεκριμένα συμπτώματα τα οποία ένας ασθενής μπορεί να παρουσιάζει και αναλύει αυτά τα συμπτώματα με βάση την κωδικοποιημένη ιατρική γνώση. Μέσω του Answer Set Programming, το πρόγραμμα επιλέγει τις ασθένειες που είναι πιθανό να προκαλούν τα συμπτώματα αυτά και τις επιστρέφει ως αποτέλεσμα.

Η καινοτομία της προτεινόμενης μεθόδου έγκειται στον συνδυασμό των δύο μεθοδολογιών. Τα Large Language Models μπορούν να κωδικοποιήσουν την ιατρική βιβλιογραφία 
σε κώδικα ASP με αξιοσημείωτη ακρίβεια και σύνθετη κατανόηση, ενώ το ASP παρέχει τον συλλογισμό και την επεξηγηματικότητά του. Ο συνδυασμός αυτών των δύο προσεγγίσεων επιτρέπει την ανάπτυξη ενός ισχυρού συστήματος πρόβλεψης ασθενειών που συνδυάζει την \textbf{ακρίβεια και την ευφυΐα των γλωσσικών μοντέλων} με την \textbf{επεξηγηματικότητα και την αποδοτικότητα του Answer Set Programming}.

Η μέθοδός μας φαίνεται να έχει υποσχόμενα αποτελέσματα πάνω στο σύνολο των δεδομένων ασθενών, στο οποίο
την εφαρμόσαμε.

% \section{Σκοπός της εργασίας}
% Στην συγκεκριμένη εργασία προτείνουμε ένα νέο τρόπο αυτόματης πρόβλεψης ασθενειών με χρήση Large Language Models και Answer Set Programming.

% H αυτόματη πρόβλεψη βασίζεται στην κωδικοποίηση της ιατρικής βιβλιογραφίας, δηλαδή την αντιστοίχηση των ασθενειών με συγκεκριμένα ιατρικά συμπτώματα και βιολογικές ενδείξεις.

% Mε την χρήση των Μεγάλων Γλωσσικών Μοντέλων καταφέρνουμε να κωδικοποιήσουμε 
% την αντιστοίχηση αυτή σε κώδικα \textit{Clingo} με αποτέλεσμα το πρόγραμμά μας, έχοντας ως είσοδο συγκεκριμένα συμπτώματα, να προβλέπει την ασθένεια πίσω από αυτά. Ο συνδυασμός των δύο αυτών μεθοδολογιών αξιοποιεί τις ικανότητες \say{μεταμόρφωσης} των μεγάλων 
% γλωσσικών μοντέλων σε συνδυασμό με την επεξηγηματικότητα της συμβολικής τεχνητής νοημοσύνης.

\section{Ανασκόπηση Βιβλιογραφίας}

\subsection{Large Language Models}

Τα μεγάλα γλωσσικά μοντέλα \cite{zhao2023survey} είναι προηγμένα υπολογιστικά μοντέλα που έχουν σχεδιαστεί για να κατανοούν και να δημιουργούν ανθρώπινη γλώσσα. Αυτά τα μοντέλα, που αναπτύχθηκαν χρησιμοποιώντας εξελιγμένες τεχνικές μηχανικής μάθησης, έχουν τη δυνατότητα να επεξεργάζονται και να δημιουργούν κείμενο με τρόπο που μοιάζει πολύ με τα πρότυπα ανθρώπινης γλώσσας.

Μεγάλα γλωσσικά μοντέλα, όπως η αρχιτεκτονική \textit{GPT-4} \cite{openai2023gpt4}, εκπαιδεύονται σε τεράστιες ποσότητες δεδομένων κειμένου, γεγονός που τους επιτρέπει να αποκτούν γνώση σε διάφορους τομείς και θέματα. Αυτά τα μοντέλα χρησιμοποιούν αλγόριθμους βαθιάς μάθησης (deep learning) \cite{Sarker2021} και ειδικά δίκτυα μετασχηματιστών (transformer networks) \cite{Dosovitskiy2020}, για την ανάλυση και την κατανόηση των σχέσεων μεταξύ λέξεων, φράσεων και προτάσεων. Καταγράφοντας μοτίβα και δομές στα δεδομένα εκπαίδευσης, αυτά τα μοντέλα μπορούν να δημιουργήσουν συνεκτικές και συναφείς
απαντήσεις με βάση δοθέντα prompts.

Tα μοντέλα ονομάζονται \say{μεγάλα} επειδή έχουν δισεκατομμύρια παραμέτρους που διαμορφώνουν τις αποκρίσεις τους. Η βασική προϋπόθεση ενός γλωσσικού μοντέλου είναι η ικανότητά του να προβλέπει την επόμενη λέξη (που ονομάζεται token) με βάση το κείμενο που έχει παρατηρήσει μέχρι τώρα, δηλαδή σύμφωνα με τα συμφραζόμενα.

Τα μεγάλα γλωσσικά μοντέλα διαθέτουν την αξιοσημείωτη ικανότητα να μετασχηματίζουν κείμενο από τη μια μορφή στην άλλη, καθιστώντας τα ισχυρά εργαλεία για διάφορες εργασίες επεξεργασίας γλώσσας. 
Πάνω σε αυτό το χαρακτηριστικό τους, βασίζεται και η ιδέα αυτής της εργασίας, αφού ένα πολύ σημαντικό κομμάτι που πραγματευόμαστε είναι η κατάλληλη κωδικοποίηση της ιατρικής βιβλιογραφίας.
Παρακάτω αναφέρουμε κάποια παραδείγματα χρήσης των μεγάλων γλωσσικών μοντέλων \cite{Paranjape2023}:

\begin{itemize}
    \item Δημιουργία κειμένου που είναι συναφές και συνεκτικό με βάση τα συμφραζόμενα
    \item Μετάφραση Γλώσσας
    \item Σύνοψη / Περίληψη Κειμένου
    \item  Απάντηση σε ερωτήσεις
    \item Παραγωγή κειμένου με βάση συγκεκριμένων κανόνων που τους παρέχονται
\end{itemize}

Ο θεμέλιος λίθος των Μεγάλων Γλωσσικών μοντέλων είναι οι \textbf{transformers}.
Οι transformers \cite{Vaswani2017} επί του παρόντος θεωρούνται μία από τις πιο προηγμένες και ισχυρές κατηγορίες μοντέλων. Συνέβαλαν καθοριστικά στην επίτευξη σημαντικών ανακαλύψεων στη μηχανική μάθηση, με αποτέλεσμα ορισμένοι να τους αναφέρουν ως \say{transformers AI}. Aυτά τα μοντέλα οδηγούν σε μια θεμελιώδη αλλαγή στον τομέα της τεχνητής νοημοσύνης \cite{Bommasani2021}.
To σχήμα \ref{fig:transformers} απεικονίζει τη δομή του transformer, ο οποίος περιλαμβάνει δύο διακριτά μέρη, την κωδικοποίηση και την αποκωδικοποίηση.

\begin{figure}[htb]
    \centering
    \includegraphics[width=0.7\textwidth]{assets/transformers_architecture.png}
    \caption{Αρχιτεκτονική του μοντέλου transformer \cite{Vaswani2017}.}
    \label{fig:transformers}
\end{figure}

\subsubsection{Inputs and Input Embeddings}

Tο θεμελιώδες συστατικό του
transformer, ο μηχανισμός \say{προσοχής} είναι εμπνευσμένος από την έννοια της συνειδητής προσοχής στους ζωντανούς οργανισμούς \cite{Vaswani2017}.
Αυτός ο μηχανισμός επιτρέπει στο δίκτυο να αποδίδει μεγαλύτερη σημασία σε κρίσιμα μέρη των δεδομένων και λιγότερη σημασία σε αυτά που είναι λιγότερο σχετικά. Αν και η απλούστερη μορφή αυτής της συνάρτησης είναι ένα σταθμισμένο άθροισμα των τιμών εισόδου, ο μηχανισμός προσοχής στον μετασχηματιστή είναι πιο περίπλοκος και περιλαμβάνει πολλαπλούς μετασχηματισμούς για τον υπολογισμό των βαρών προσοχής μεταξύ διαφορετικών τμημάτων της εισόδου.

% Αν θεωρήσουμε ως είσοδο στο δίκτυο τον πίνακα $\textbf{x} \in \Re^ {d_m\times d}$ ($n$ ο αριθμός των αντικειμένων εισόδου και $d$ η διάστασή τους), οι σχέσεις που εκφράζουν τον μηχανισμό προσοχής είναι:

% \begin{equation}
%     Q(x_i) = x_i W_q 
% \end{equation}

% \begin{equation}
%     K(x_i) = x_i W_k
% \end{equation}

% \begin{equation}
%     V(x_i) = x_i W_v
% \end{equation}

% όπου: $W_q, W_k, \in \Re^{d\times d_v} $

% \begin{equation}
%     attention(Q, K, V) = softmax\frac{Q K ^ \top}{\sqrt{d_k}} V
% \end{equation}

% όπου:

% \begin{equation}
% \begin{bmatrix}
% Q(\textbf{x}_1) \\
% \vdots \\
% Q(\textbf{x}_n) \\
% \end{bmatrix}
% \in \Re ^{n \times d_q}
% \end{equation}

% \begin{equation}
% \begin{bmatrix}
% K(\textbf{x}_1) \\
% \vdots \\
% K(\textbf{x}_n) \\
% \end{bmatrix}
% \in \Re ^{n \times d_k}
% \end{equation}

% \begin{equation}
% \begin{bmatrix}
% V(\textbf{x}_1) \\
% \vdots \\
% V(\textbf{x}_n) \\
% \end{bmatrix}
% \in \Re ^{n \times d_v}
% \end{equation}

% Οι πίνακες $Q, K$ και $V$ συμβολίζονται ως ερωτήματα, κλειδιά και τιμές, αντίστοιχα, και οι εγγραφές τους λαμβάνονται πολλαπλασιάζοντας τους πίνακες $W_q, W_k$ και $W_v$ με όλα τα διανύσματα εισόδου $x_i$, όπου το $i$ κυμαίνεται από $1$ έως $n$. Κατά τη διάρκεια της εκπαίδευσης, προσδιορίζονται οι πίνακες βάρους $W_q, W_k$ και $W_v$ και τα διανύσματα εισόδου $x_i$ είναι γνωστά ως \textbf{input embeddings}.

% Κάθε είσοδος $x_i$ μετατρέπεται σε έναν πίνακα $Q(x_i)$. Στον μηχανισμό προσοχής, η συμβατότητα όλων των πινάκων $Q(x_i)$ με όλους τους βασικούς πίνακες $K(x_j)$ (όπου $j$ κυμαίνεται από 1 έως $n$ και $j$ δεν ισούται με $i$) υπολογίζεται από το εσωτερικό γινόμενο $QK ^\top$. Η προσοχή του δικτύου αυξάνεται με μεγαλύτερες τιμές του προϊόντος $QK ^\top$.

% Η σχέση 4 υπολογίζει τις κανονικοποιημένες τιμές του προϊόντος $QK ^\top$ για μια μεμονωμένη αναπαράσταση των $Q, K, V$. Η κανονικοποίηση κατά $ \sqrt{d_k}$ εφαρμόζεται για να αποτρέψει το εσωτερικό προϊόν $QK ^\top$ από το να γίνει πολύ μεγάλο και να προκαλέσει αστάθεια στην εκπαίδευση. Για να εξαχθούν περισσότερες πληροφορίες από την είσοδο, το μοντέλο δημιουργεί διαφορετικές αναπαραστάσεις των πινάκων $Q, K, V$ πολλαπλασιάζοντάς τους με $h$ διαφορετικούς πίνακες βάρους. Οι προκύπτουσες έξοδοι συνδυάζονται και διέρχονται από ένα τελικό γραμμικό στρώμα, όπως απεικονίζεται στο σχήμα και εκφράζεται με τις εξισώσεις 8 και 9. Αυτό αναφέρεται συνήθως ως "multi-head attention" μηχανισμός στην ερευνητική κοινότητα. Σε κάθε βήμα εκμάθησης, το μοντέλο ενημερώνει τις αναπαραστάσεις ερωτήματος, κλειδιού και τιμών για να βελτιώσει την απόδοσή του.

% \begin{equation}
%     head(i) = attention((W^Q_i Q^T, W^k_i K^T, W^V_i V^T) 
% \end{equation}
% \begin{equation}
%     MLH = \text{Concat}(head_1, \ldots, head_h)W^o
% \end{equation}

% \begin{figure}[htb]
%     \centering
%     \includegraphics[width=\textwidth]{assets/multihead.png}
%     \caption{Η αριστερή εικόνα δείχνει τη ροή του μηχανισμού "multi-head attention"  και η δεξιά τα λεπτομερή στάδια του κλιμακούμενου dot product. Ως $Q, K, V$ συμβολίζονται οι πίνακες που περιέχουν το ερώτημα, κλειδί, τιμή αντίστοιχα.\cite{Vaswani2017}}
%     \label{fig:multihead}
% \end{figure}

Τα τμήματα εισόδου από τον χρήστη είναι είσοδοι για τα μοντέλα μηχανικής μάθησης. Ωστόσο, τα μοντέλα κατανοούν μόνο αριθμούς, όχι κείμενο, επομένως αυτές οι είσοδοι πρέπει να μετατραπούν σε μια αριθμητική μορφή που ονομάζεται \textbf{input embeddings}. Tα input embeddings αντιπροσωπεύουν λέξεις ως αριθμούς, τους οποίους μπορούν στη συνέχεια να επεξεργαστούν τα μοντέλα μηχανικής μάθησης. Τα embeddings είναι σαν ένα λεξικό, το οποίο βοηθά το μοντέλο να κατανοήσει τη σημασία των λέξεων, μέσω της τοποθέτησής τους σε έναν μαθηματικό χώρο όπου παρόμοιες λέξεις βρίσκονται η μία κοντά στην άλλη. Κατά τη διάρκεια της εκπαίδευσης, το μοντέλο μαθαίνει πώς να δημιουργεί τα embeddings, έτσι ώστε παρόμοια διανύσματα να αντιπροσωπεύουν λέξεις με παρόμοια σημασία.

\subsubsection{Positional Encodings}

Στην επεξεργασία φυσικής γλώσσας, η σειρά των λέξεων σε μια πρόταση είναι κρίσιμη για τον προσδιορισμό του νοήματος της πρότασης. Ωστόσο, τα παραδοσιακά μοντέλα μηχανικής μάθησης, όπως τα νευρωνικά δίκτυα, δεν κατανοούν εγγενώς τη σειρά των εισόδων. Για την αντιμετώπιση αυτής της πρόκλησης, η κωδικοποίηση θέσης μπορεί να χρησιμοποιηθεί για την κωδικοποίηση της θέσης κάθε λέξης σε μία πρόταση. Η κωδικοποίηση αυτή τροφοδοτείται στο μοντέλο Transformer, μαζί με τα input embeddings. Με την ενσωμάτωση της κωδικοποίησης θέσης στην αρχιτεκτονική του Transformer, το μοντέλο μπορεί να κατανοήσει πιο αποτελεσματικά τη σειρά των λέξεων σε μια πρόταση και να δημιουργήσει γραμματικά σωστά και σημασιολογικά νόημα.
Το \say{γραμματικά σωστό} αποτέλεσμα μπορεί να σημαίνει και την παραγωγή κώδικα ο οποίος είναι 
σύμφωνος με τους κάνόνες μιας γλώσσας προγραμματισμού.

\subsubsection{Encoder}

Ο κωδικοποιητής (encoder) είναι μέρος του νευρωνικού δικτύου που επεξεργάζεται το κείμενο εισόδου και δημιουργεί μια σειρά από κρυφές καταστάσεις που καταγράφουν το νόημα του κειμένου. Ο κωδικοποιητής διαμορφώνει πρώτα το κείμενο εισόδου σε μια ακολουθία tokens, π.χ. μεμονωμένες λέξεις.

\subsubsection{Decoder}
Ο αποκωδικοποιητής (decoder) είναι μέρος του μοντέλου που δημιουργεί την ακολουθία εξόδου με βάση την κωδικοποιημένη ακολουθία εισόδου. Κατά τη διάρκεια της εκπαίδευσης, ο αποκωδικοποιητής μαθαίνει πώς να \say{μαντεύει} την επόμενη λέξη κοιτάζοντας τις λέξεις πριν από αυτήν. Για παράδειγμα, ο αποκωδικοποιητής σε \textit{GPT} δημιουργεί κείμενο φυσικής γλώσσας με βάση την ακολουθία εισόδου και το περιβάλλον που μαθαίνει ο κωδικοποιητής. Στους transformers χρησιμοποιούνται πολλαπλά στρώματα αποκωδικοποιητών.

\subsection{Δυσκολίες στην εφαρμογή των LLMs}

Τα μεγάλα γλωσσικά μοντέλα δεν είναι αλάνθαστα \cite{Raj2023}, \cite{Ruis2023}. Μπορούν περιστασιακά να παράγουν λανθασμένες ή μεροληπτικές απαντήσεις, αντανακλώντας τις προκαταλήψεις που υπάρχουν στα δεδομένα εκπαίδευσης. Καταβάλλονται προσπάθειες για τον μετριασμό αυτών των ζητημάτων μέσω τεχνικών προεπεξεργασίας δεδομένων και fine-tuning.

Συνολικά, τα μεγάλα γλωσσικά μοντέλα έχουν επιδείξει τεράστιες δυνατότητες σε ένα ευρύ φάσμα εφαρμογών, συμπεριλαμβανομένης της κατανόησης φυσικής γλώσσας, της δημιουργίας περιεχομένου και της αλληλεπίδρασης ανθρώπου-υπολογιστή. Η συνεχής έρευνα και ανάπτυξη σε αυτόν τον τομέα υπόσχεται περαιτέρω ενίσχυση αυτών των μοντέλων και των εφαρμογών τους σε διάφορους τομείς, συμβάλλοντας στην πρόοδο των τεχνολογιών γλωσσικής επεξεργασίας και επικοινωνίας.

\subsection{Answer Set Programming}
To Answer Set Programming (ASP) \cite{Eiter2009} είναι μια δηλωτική μέθοδος επίλυσης προβλημάτων που βασίζεται στον Λογικό Προγραμματισμό και τη Μη Μονοτονική Συλλογιστική.  
Στην εργασία \cite{Gelfond2000} επισημοποιήθηκε αρχικά η σημασιολογία των σταθερών μοντέλων και η βασική γλώσσα του ASP.

Ο προγραμματισμός με χρήση αυτής της προσέγγισης γίνεται σε μια οικογένεια γλωσσών που ονομάζεται ορισμένες φορές \textit{AnsProlog} \cite{Gelfond2002}.

Η ιδέα πίσω από το ASP είναι να μοντελοποιηθεί ένα πρόβλημα ως ένα σύνολο κανόνων και δεδομένων, και στη συνέχεια να χρησιμοποιηθεί ένας επίλυτης για να βρεθεί μια λύση για το πρόβλημα. Οι λύσεις παριστάνονται από σταθερά μοντέλα ή σύνολα απαντήσεων. Οι κανόνες, τα δεδομένα και οι περιορισμοί που περιγράφουν το πρόβλημα αποτελούν τα στοιχεία του προγράμματος. Στη συνέχεια, το πρόγραμμα δίνεται σε έναν επίλυτη που θα βρει μία ή περισσότερες λύσεις.

Στον παραδοσιακό προγραμματισμό, η μετάβαση από ένα πρόβλημα σε μια λύση προϋποθέτει την απόλυτη κατανόηση του δεδομένου προβλήματος από τον προγραμματιστή, και έπειτα την παραγωγή ενός προγράμματος, στο οποίο όταν δοθεί ένα instance του προβλήματος θα παράγει το σωστό αποτέλεσμα, τα οποίο στη συνέχεια θα ερμηνευτεί ως λύση.

Στο ASP, η μετάβαση από τη δήλωση προβλήματος σε ένα σύνολο λύσεων περιλαμβάνει τα ακόλουθα βήματα \cite{Gebser2013}
(εικόνα \ref{fig:asp-solving}):

\begin{figure}[htb]
    \begin{center}
\begin{tikzpicture}[node distance=4cm]
        \node [simple-rect] (problem)  at (0,0) {Problem};
        \node [simple-rect, below of=problem] (program) {Program};

        \node [simple-rect, right of=program] (grounder) {Grounder};
        \node [simple-rect, right of=grounder] (solver) {Solver};

        \node [simple-rect, right of=solver] (stable-models) {Stable Models};

        \node [simple-rect, above of=stable-models] (solution) {Solution};

        \draw [arrow] (problem) -- node[right] {\textbf{Modelling}} (program);
        \draw [arrow] (stable-models) -- node[left] {Interpreting} (solution);

        \draw [arrow-small] (program) -- node[above] {} (grounder);
        \draw [arrow-small] (grounder) -- node[above] {\footnotesize Grounding} (solver);
        \draw [arrow-small] (solver) -- node[above] {} (stable-models);

        \node[draw, dotted, fit=(grounder) (solver), inner sep=4mm, label=above:{Solving}] {};
\end{tikzpicture}
    \end{center}
    \caption{Η επίλυση ενός προβλήματος με χρήση ASP.}
    \label{fig:asp-solving}
\end{figure}

\begin{itemize}
    \item \textbf{Modeling}: Το πρόβλημα μοντελοποιείται σε γλώσσα ASP.
    \item \textbf{Grounding}: Ένας grounder (π.χ. gringo) μετατρέπει το πρόγραμμα σε ένα σύνολο βασικών κανόνων και δεδομένων.
    \item \textbf{Solving}: Ένας επιλυτής (π.χ. clasp) βρίσκει μια λύση στο πρόβλημα υπολογίζοντας το σύνολο των σταθερών μοντέλων (σύνολα απαντήσεων).
\end{itemize}

Έχουν αναπτυχθεί αρκετοί λύτες για ASP όπως το DLV \cite{Xia2020}. Σε αυτή την εργασία, θα εφαρμόσουμε το σύστημα ASP clingo \cite{Gebser2014} το οποίο συνδυάζει τον grounder gringo με τον clasp solver\cite{Holldobler2014} σε μια ενιαία εφαρμογή, παρέχοντας επίσης ένα ισχυρό Python API για την ενσωμάτωση του solver σε άλλες εφαρμογές.

\subsection{Χρήση του ASP στην διάγνωση ασθενειών}
Το Answer Set Programming (ASP) προσφέρει μια ισχυρή μέθοδο για την αναπαράσταση της ιατρικής γνώσης με δηλωτικό και λογικό τρόπο. Ακολουθούν ορισμένες μέθοδοι ASP που χρησιμοποιούνται συνήθως στην αναπαράσταση ιατρικής γνώσης \cite{Vinarti2019}:

\begin{itemize}
    \item \textbf{Αναπαράσταση βάσει κανόνων}: Το ASP επιτρέπει την αναπαράσταση της ιατρικής γνώσης ως κανόνων με τη μορφή λογικών δηλώσεων. Αυτοί οι κανόνες ορίζουν σχέσεις μεταξύ ιατρικών οντοτήτων όπως συμπτώματα, ασθένειες, θεραπείες και δεδομένα ασθενών. Για παράδειγμα, ένας κανόνας θα μπορούσε να αναφέρει ότι εάν ένας ασθενής εμφανίζει συγκεκριμένα συμπτώματα, υποδηλώνει την παρουσία ή την απουσία ορισμένων ασθενειών.

    \item \textbf{Κατασκευή Γνωσιακής Βάσης}: Το ASP παρέχει ένα πλαίσιο για την κατασκευή μιας βάσης γνώσεων που συλλαμβάνει την ιατρική γνώση. Η βάση γνώσεων αποτελείται από γεγονότα, κανόνες και περιορισμούς που καθορίζουν τις πληροφορίες και τους περιορισμούς που σχετίζονται με την ιατρική διάγνωση και θεραπεία. Μπορεί να περιλαμβάνει πληροφορίες σχετικά με τις σχέσεις ασθένειας-συμπτωμάτων, διαγνωστικά κριτήρια, κατευθυντήριες γραμμές θεραπείας και δεδομένα για τον ασθενή.

\end{itemize}

Στην εργασία \cite{Erdem2011} εισήχθησαν νέες μέθοδοι για αποτελεσματικό χειρισμό σύνθετων ερωτημάτων στο πλαίσιο βιοϊατρικών οντολογιών και βάσεων δεδομένων, με έμφαση στην εξαγωγή σχετικών πληροφοριών από αυτούς τους πόρους γνώσης. Aυτές οι μέθοδοι εφαρμόστηκαν για την αντιμετώπιση απαιτητικών ερωτημάτων στον τομέα της ανακάλυψης φαρμάκων, χρησιμοποιώντας βιολογικές βάσεις δεδομένων σε συνδυασμό με συνοπτικές εξηγήσεις για πολύπλοκα ερωτήματα που σχετίζονται με διαδικασίες ανακάλυψης φαρμάκων.

\section{Μεθοδολογία}

\subsection{Αρχιτεκτονική}

\begin{figure}[!h]
\begin{center}
\begin{tikzpicture}[node distance = 5cm, minimum width=2cm, minimum height=1cm]
    \node [doc, label=above:{}] (med-bib) {Medical Bibliography};
    \node [rect, draw, right=3cm of med-bib, fill=blue!50] (llm) {Large Language Model};
    \node [rect, draw, above= 1cm of med-bib, fill=orange!50] (prompt) {Prompt};
    \node [database, below left= 3cm of med-bib, fill=green!30] (knowledge-base) {Knowledge Base};
    \node [rect, draw, right of=knowledge-base] (solver) {ASP Solver};
    \node [rect, draw, below=1cm of solver] (patient) {\textit{Patient Data}};
    \node [rect, right of=solver] (diagnosis) {\textbf{Diagnosis}};

    \draw[thick-arrow] (med-bib) edge [] node [above] {} (llm);
    \draw[thick-arrow] (prompt) edge [in=90, out=0, looseness=1] node [above, sloped] {\textit{\say{please convert the following into ASP code...}}} (llm);
    \draw[thick-arrow] (llm) edge [in=90, out=270, looseness=1] node [below] {ASP Program} (knowledge-base);
    \draw[thick-arrow] (knowledge-base) edge [] node [above] {} (solver);
    \draw[thick-arrow] (solver) edge [] node [above] {} (diagnosis);
    \draw[thick-arrow] (patient) edge [in=270, out=180, looseness=1] node [above] {} (knowledge-base);

\end{tikzpicture}
\end{center}
    \caption{Η διαδικασία της διάγνωσης με τη χρήση της προτεινόμενης μεθόδου.
    Ιατρική βιβλιογραφία, με χρήση ενός μεγάλου γλωσσικού μοντέλου, μετατρέπεται σε ένα πρόγραμμα ASP.
    Στη συνέχεια, το πρόγραμμα αυτό εκτελείται από έναν ASP solver, ο οποίος παράγει την τελική διάγνωση
    με βάση τα δεδομένα για τον εκάστοτε ασθενή.}
    \label{fig:architecture}
\end{figure}

Σκοπός μας είναι η εξαγωγή μίας βάσης γνώμης μέσα από αδώμητο κείμενο \cite{ré2014feature}.
Η αρχιτεκτονική μας (σχήμα \ref{fig:architecture}) περιλαμβάνει τα εξής στάδια

\begin{enumerate}
    \item Συλλογή ιατρικής βιβλιογραφίας σχετική με τις εξεταζόμενες ασθένειες/παθήσεις.
    \item Κατασκευή κατάλληλου prompt για την μετατροπή ιατρικής βιβλιογραφίας σε κώδικα ASP.
    \item Προσθήκη του παραγώμενου ASP κώδικα από το μεγάλο μοντέλο γλώσσας στη βάση γνώσης.
    \item Προσθήκη σχετικών δεδομένων του ασθενή (συμπτώματα, αποτελέσματα ιατρικών εξετάσεων) στη βάση γνώσης.
    \item Παραγωγή της τελικής διάγνωσης (ή σύνολο πιθανών διαγνώσεων) ως answer set(s), αφού ο solver συλλογιστεί 
          πάνω στη τελική βάση γνώσης.
\end{enumerate}

\subsection{Κατασκευή Prompt}

Τα μεγάλα γλωσσικά μοντέλα, παρά των υψηλών δυνατοτήτων τους στην εκτέλεση οδηγιών, ακόμα 
έχουν χαμηλή ελεγξιμότητα \cite{bowman2023things}.
Έτσι, έχει προκύψει μία σχετικά νέα μεθοδολογία \say{προγραμματισμού} γνωστή ως prompt engineering \cite{liu2021pretrain, promptingguidePromptEngineering}.
Προκειμένου να μετατρέψουμε την ιατρική βιβλιογραφία σε κώδικα
κατάλληλο για την εκτέλεσή του σε επιλυτή ASP έπρεπε να κατασκευάσουμε 
ποιοτικά prompts ώστε ο παραγώμενος κώδικας να βρίσκεται στην επιθυμητή μορφή (σχήμα \ref{fig:bad-prompt}, σχήμα \ref{fig:good-prompt}).

\begin{figure}[!h]
\centering
\begin{prompt}
\textit{Medical Text...}

The above paragraph refers some common symptoms of pneumonia.
Give me a Clingo script so that based on symptoms I can diagnose pneumonia.
\end{prompt}

\begin{commandline}
\begin{verbatim}
% Symptoms of Pneumonia

symptom(cough).
symptom(cough_with_mucus).
symptom(chest_pain).
symptom(headache).
...

% Diagnosis of pneumonia
diagnosis(pneumonia) :- symptom(cough), 
                        symptom(chest_pain), 
                        symptom(headache), 
                        symptom(arthralgia),
                        ...
\end{verbatim}
\end{commandline}

\caption{Ένα απλό prompt μέσω του οποίου δεν πέρνουμε ικανοποιητικά αποτελέσματα.}
\label{fig:bad-prompt}
\end{figure}

\begin{figure}[!h]
\begin{prompt}
\textit{Medical Text...}

The above paragraph refers some common symptoms of pneumonia.
Give me a Clingo script so that based on symptoms I can diagnose pneumonia.

In the clingo script that you produced before, make some changes in diagnosis rule like:
\src{diagnosis(pneumonia) :- has(symptom(x)), has(symptom(y)) ...}

Include some alternative diagnosis for pneumonia.
Create rules based on one symptom being linked to another.
\end{prompt}

\begin{commandline}
\scriptsize
\begin{verbatim}
% Symptoms of Pneumonia
symptom(cough).
symptom(cough_with_mucus).
...

% Rules linking symptoms

% Cough with mucus can be associated with wheezing and 
% shortness of breath
linked_symptom(cough_with_mucus, wheezing).
linked_symptom(cough_with_mucus, shortness_of_breath).

...

% Rule to infer linked symptoms
has(symptom(Symptom2)) :- has(symptom(Symptom1)), 
              linked_symptom(Symptom1, Symptom2).

% Diagnosis of Pneumonia
diagnosis(pneumonia) :- has(symptom(cough)),
                        has(symptom(cough_with_mucus)),
...

% Alternative Diagnoses for Pneumonia
diagnosis(pneumonia) :- has(symptom(cough_with_mucus)),
                        has(symptom(chest_pain)),
...
\end{verbatim}
\end{commandline}
\caption{Δίνοντας αναλυτικότερες πληροφορίες μπορούμε να καθοδηγήσουμε το μεγάλο γλωσσικό μοντέλο στο να μετατρέπει την εισαγώμενη ιατρική βιβλιογραφία στην μορφή που θέλουμε.}
\label{fig:good-prompt}
\end{figure}

\subsection{Κατασκευή Συλλογιστικού Προγράμματος}

Η βάση γνώσης αρχικά αποτελείται από ένα σύνολο συμπτωμάτων και ιατρικών ενδείξεων της μορφής:

\begin{align}
    symptom(s). & \; s \in \{ \text{cough, chest pain, rash}, \dots \} \\
    indicator(i). & \; i \in \{ \text{low MCV, high TIBC}, \dots \}
\end{align}

Αποτυπώνουμε έναν ασθενή ο οποίος έχει κάποιο σύμπτωμα ή βγήκε θετικός σε κάποια εξέταση ως:

\begin{equation}
    has(x). \; x \in S \cup I
\end{equation}

Οι λογικές προτάσεις με βάση τις οποίες το σύστημά μας θα βγάζει συμπεράσματα είναι της μορφής:

\begin{align}
    diagnosis(d_1) & \longleftarrow has(x_1) \land has(x_2) \land \dots \\
    diagnosis(d_1) & \longleftarrow has(x_1) \land has(x_3) \land \dots \\
    diagnosis(d_2) & \longleftarrow has(x_3) \land has(x_4) \land \dots
\end{align}

Ακόμα, λαμβάνουμε υπόψιν τη συσχέτιση μεταξύ ορισμένων συμπτωμάτων.
Παραδείγματος χάρη, σε ορισμένους ασθενείς έχει σημειωθεί σύνδεση  
μεταξύ δερματικών ασθενειών και πονοκεφάλων\cite{migraine-hives}.
Άλλες φορές, συμπτώματα μπορεί να οδηγήσουν στην εξέλιξη άλλων συμπτωμάτων.
Τέτοιου είδους σχέσεις κωδικοποιούνται στη βάση γνώσης μας μέσω κανόνα της μορφής.

\begin{align}
    has(symptom(S2)) \longleftarrow has(symptom(S1)) \land linked\_symptom(S1, S2).
\end{align}

Τα αντίστοιχα δεδομένα θα είναι ως εξής

\begin{align}
    linked\_symptom(grunting, chest\_retractions).\\
    linked\_symptom(nasal\_flaring, chest\_retractions).
\end{align}

Ωστόσο, είναι αναμενόμενο πως κάποιος ασθενής δεν θα κατέχει \textit{όλα} τα καταγεγραμμένα συμπτώματα.
Θέλουμε να δώσουμε έναν ελέγξιμο τρόπο το σύστημά μας να δίνει εναλλακτικές προβλέψεις
στην περίπτωση όπου το ιατρικό προσωπικό δεν έχει όλες τις πληροφορίες για την κατάσταση του ασθενή.
Έτσι, προσθέτουμε τον παρακάτω κανόνα επιλογής (choice rule), με τον οποίο προστίθεται 
ένα σύνολο πιθανών συμπτώματων για τον ασθενή στη βάση γνώσης.

\begin{equation}
    \{ add(symptom(S)) : symptom(S) \}.
\end{equation}

Ακόμα, απαιτούμε από τον επιλυτή να βρει τουλάχιστον μία διάγνωση.

\begin{equation}
    \bot \longleftarrow \sim diagnosis(\_)
\end{equation}

Τέλος, ελαχιστοποιούμε τον αριθμό \say{ψεύτικων} διαγνώσεων, έτσι ώστε ο επιλυτής 
να καταλήξει σε μία διάγνωση με όσο το δυνατόν λιγότερες υποθέσεις.

\begin{equation}
    \#minimize \{ 1, S : add(S) \}
\end{equation}

Με αυτό τον τρόπο, το σύστημά μας εγγυάται πως θα παράξει μία τουλάχιστον πρόβλεψη, η οποία 
θα βασίζεται όσο το δυνατόν περισσότερο στις έγγυρες πληροφορίες που υπάρχουν για τον ασθενή.

\subsection{Επεξηγηματικότητα Διαγνώσεων}

Η συμβολική τεχνητή νοημοσύνη έχει ένα μείζον πλεονέκτημα έναντι των υποσυμβολικών προσεγγίσεων λόγω της εξηγηματικότητάς της.
Οι σχέσεις μεταξύ των γεγονότων και των συμπερασμάτων που καταλήγει ένα σύστημα, το οποίο εκτελεί ένα συμβολικό μηχανισμό είναι σαφώς καθορισμένες.
Ωστόσο, οι επίλυτές ASP όπως το \textit{Clingo} δεν παρέχουν\footnote{Τη στιγμή της συγγραφής της εργασίας αυτής.} εύληπτα αποτυπώματα των συλλογιστικών διαδικασιών τους.
Έχει υπάρξει έρευνα στην σύνδεση λογικών προτάσεων 
με τις αιτιολογήσεις τους \cite{cabalar2014causal}.
Θεωρήστε το παρακάτω λογικό πρόγραμμα, η σύνδεση των λογικών προτάσεων με τις αιτίες τους παρουσιάζονται 
σε μορφή γράφου στο σχήμα \ref{fig:causal-g}. Αυτή είναι η δομή που μας επιτρέπει 
να εξηγήσουμε τα συμπεράσματα ενός επιλυτή συμβολικής λογικής.

\begin{equation}
\begin{aligned}[b]
    l :& punish \longleftarrow drive, drunk \\
    m :& punish \longleftarrow resist \\
    e :& prison \longleftarrow punish \\
    d :& drive \\
    k :& drunk \\
    r :& resist \\
\end{aligned}
\label{eq:example-cg}
\end{equation}

\begin{figure}
    \centering
    \includegraphics{assets/causal_g.png}
    \caption{Αιτιολογικός γράφος λογικού ποργράμματος (εξ. \ref{eq:example-cg}).}
    \label{fig:causal-g}
\end{figure}

Το εργαλείο \textit{xclingo}\footnote{\url{https://github.com/bramucas/xclingo2}} \cite{Cabalar_2020} επιτρέπει την παρουσίαση αιτιολογήσεων σχετικά με τα αποτελέσματα που δίνονται από τον επιλυτή \textit{Clingo}.
Αυτό επιτυγχάνεται προσθέτοντας ένα \src{trace} που αντιστοιχεί στους κανόνες του λογικού προγράμματός μας.
Η παροχή μιας σαφούς εξηγήσεως για το πώς το σύστημα κατέληξε σε συγκεκριμένα συμπεράσματα, μπορεί να βοηθήσει στον εντοπισμό πιθανών σφαλμάτων κατά την δημιουργία της βάσης γνώσης και να παρέχει στον χρήστη του συστήματος μια αιτιολόγηση για το τελικό αποτέλεσμα. Αυτό είναι κάτι πολύ σημαντικό στην περίπτωση που η εφαρμογή χρησιμοποιείται σε ιατρικά πλαίσια.

\section{Αποτελέσματα}

Για την αξιολόγηση του συστήματός μας χρησιμοποίησαμε δημοσίως διαθέσιμα δεδομένα ασθενών\footnote{\url{https://www.kaggle.com/datasets/itachi9604/disease-symptom-description-dataset?select=dataset.csv}}.
Στο σύστημά μπήκε σαν είσοδος το σύνολο των συμπτωμάτων του κάθε ασθενή.
Σαν βάση γνώσης έχουμε τον συνδυασμό όλων των κανόνων που έχουν προκύψει από την κωδικοποίηση την ιατρικής βιβλιογραφίας.  
Μία επιτυχημένη πρόβλεψη ορίζεται όταν η πραγματική ασθένεια του ασθενή συμπίπτει με την τελική
πρόβλεψη του επιλυτή ASP.

\begin{table}[h]
    \centering
    \begin{tabular}{lcc}
    \hline
    Ασθένεις & Μέγεθος Βάσης Γνώσης & Ακρίβεια \\
    \hline
    \hline
    Chicken Pox      &  66  & 95\% \\
    Pneumonia        &  75  & 100\% \\
    Common Cold    &   44 & 100\% \\
    \hline
    \end{tabular}
    \caption{Ακρίβεια της μεθόδου για μερικές ενδεικτικές ασθένειες.}
\end{table}

Η ακρίβεια του μοντέλου μας μπορεί να φτάσει πολύ υψηλά ποσοστά ακρίβειας αλλά απαιτεί την εισαγωγή ενός μεγάλου πλήθους κανόνων στη βάση γνώσης.
Το μέγεθος της βάσης γνώσης το ορίζουμε ως το πλήθος των λογικών όρων μέσα 
στο ASP πρόγραμμα που αντιστοιχεί σε κάθε μία από τις ασθένειες.

Στο πρόγραμμα που κατασκευάσαμε προσθέσαμε και τη δυνατότητα 
επεξήγησης της διάγνωσης, κατά την οποία αξιοποιούμε 
το εργαλείο \src{xclingo} για τη δημιουργία ενός γραφήματος
το οποίο αποτυπώνει τους συνειρμούς που ακολούθησε το σύστημά 
μας για τις προβλέψεις του.

\begin{lstlisting}[caption={Μέρος της επεξήγησης για τη διάγνωση της ασθένειας chickenpox. Μέσα στο διάγραμμα δέντρο φαίνονται οι συσχετίσεις μεταξύ των διαφορετικών συμπτωμάτων 
και το πώς το σύστημά μας καταλήγει στα συμπεράσματά του.}, label=lst:terms]{Name}
  *
  |__diagnosis(chickenpox)
  |  |__has(symptom(itching));has(symptom(itching))
  |  |__has(symptom(fatigue));has(symptom(fatigue));has(symptom(fatigue));has(symptom(fatigue));has(symptom(fatigue))
  |  |__has(symptom(lethargy))
  |  |  |__has(symptom(fatigue));has(symptom(fatigue));has(symptom(fatigue));has(symptom(fatigue));has(symptom(fatigue))
  |  |  |__linked_symptom(fatigue,lethargy)
  |  |__has(symptom(high_fever))
  |  |  |__has(symptom(mild_fever));has(symptom(mild_fever))
  |  |  |  |__has(symptom(loss_of_appetite));has(symptom(loss_of_appetite))
  |  |  |  |__linked_symptom(loss_of_appetite,mild_fever)
  |  |  |__linked_symptom(mild_fever,high_fever)
  |  |__has(symptom(loss_of_appetite));has(symptom(loss_of_appetite))
  |  |__has(symptom(mild_fever));has(symptom(mild_fever))
  |  |  |__has(symptom(loss_of_appetite));has(symptom(loss_of_appetite))
  |  |  |__linked_symptom(loss_of_appetite,mild_fever)
  |  |__has(symptom(swelled_lymph_nodes))
  |  |  
  ...
\end{lstlisting}

\section{Συμπεράσματα}

Στην εργασία αυτή έγινε η παρουσίαση μίας μεθοδολογίας για την αυτόματη πρόβλεψη ασθενειών 
η οποία περιλαμβάνει την κωδικοποίηση ιατρικής βιβλιογραφίας και τη μετατροπή της σε κώδικα 
ASP μέσω ενός μεγάλου γλωσσικού μοντέλου.
Έγινε παρουσίαση του τρόπου λειτουργίας των μεγάλων γλωσσικών μοντέλων και του ASP, 
ενώ έγινε συγκεκριμένη αναφορά στο ρόλο που παίζει η κάθε τεχνολογία στο τελικό σύστημα.
Ακόμα, δείξαμε την αρχιτεκτονική μας και το πώς αυτές οι τεχνολογίες θα συνδυαστούν ενώ
δώσαμε συγκεκριμένες υλοποιητικές λεπτομέρειες τις οποίες αναπτύξαμε τόσο
για την δημιουργία ποιοτικών prompts για το μεγάλο γλωσσικό μοντέλο, όσο και
στην δομή των κανόνων στην τελική ASP βάση γνώσης.
Τελικά, εκτελέσαμε προβλέψεις πάνω σε ορισμένες ασθένειες για τις οποίες δημιουργήσαμε την
βάση γνώσης με τη χρήση της μεθοδολογίας μας και παρουσιάσαμε τα αποτελέσματα.
Η μέθοδός μας φαίνεται να έχει υποσχόμενα αποτελέσματα και προσφέρει αξιόπιστες προβλέψεις υψηλής ακρίβειας
μαζί με τις επεξηγήσεις τους.

Μελλοντική δουλειά είναι η επέκταση της μεθόδου για περισσότερες ασθένειες.
Ταυτόχρονα θα ήταν χρήσιμη η δημιουργία μίας γραφικής διεπαφής μέσω 
της οποίας κάποιος να μπορεί να εισάγει στοιχεία κάποιου ασθενή και να παίρνει 
πιθανές διαγνώσεις.

\newpage
\bibliographystyle{unsrt}
\bibliography{bibliography}

\end{document}

### 2025-05-28 paper/draft.tex
Use simpler template
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

--- 2025-05-28 paper/main.tex
@@
-\documentclass[12pt]{extarticle}
+\documentclass[10pt,leqno]{amsart}
+\usepackage{graphicx}
+\baselineskip=16pt
 
-\input{preamble}
+\usepackage{indentfirst,csquotes}
 
-% Basic Document Info
+\topmargin= .5cm
+\textheight= 20cm
+\textwidth= 32cc
+\baselineskip=16pt
 
-% \title{Διάγνωση ασθενειών με χρήση \en{ASP}}
+\evensidemargin= .9cm
+\oddsidemargin= .9cm
+
+\usepackage{amssymb,amsthm,amsmath}
+\usepackage{xcolor,paralist,hyperref,titlesec,fancyhdr,etoolbox}
+\usepackage{datetime}
+\usepackage{wrapfig}
+\usepackage{enumerate}
+\usepackage{float}
+\usepackage{graphicx}
+\usepackage{caption}
+\usepackage[acronym]{glossaries}
+\usepackage[nottoc]{tocbibind}
+\usepackage{siunitx}
+\usepackage{subcaption}
+\usepackage{geometry}
+\usepackage{fancyhdr}
+\usepackage{cite}
+\usepackage{subfiles}
+\usepackage{amsmath}
+\usepackage{array,tabularx}
+\usepackage{chngcntr}
+\usepackage{afterpage}
+\usepackage{ulem}
+\usepackage{polyglossia}
+\usepackage{fontspec}
+\usepackage{hyperref}
+\usepackage{dirtytalk}
+\usepackage{algorithm2e}
+\usepackage{enumitem,amssymb}
+\usepackage{pifont}
+\usepackage{amsmath}
+\usepackage{listings}
+\usepackage{xcolor}
+\usepackage{formal-grammar}
+\usepackage{varwidth}
+\usepackage{fmtcount}
+\usepackage{tikz}
+\usepackage{tikzpeople}
+\usepackage{array}
+\usepackage[framemethod=tikz]{mdframed} % Allows defining custom boxed/framed environments
+\usepackage{cleveref}
+\usepackage{epigraph}
+\usepackage{microtype}
+\usepackage{dirtytalk}
+
+\newtheorem{theorem}{Theorem}[]
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{example}[theorem]{Example}
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{proposition}[theorem]{Proposition}
+\newtheorem{corollary}[theorem]{Corollary}
+\newtheorem{conjecture}[theorem]{Conjecture}
+
+\titleformat{\section}[display]{\normalfont\huge\bfseries\centering}{\centering\chaptertitlename\thechapter}{10pt}{\Large}
+\titlespacing*{\section}{0pt}{0ex}{0ex}
+
+\hypersetup{ colorlinks=true, linkcolor=black, filecolor=black, urlcolor=black }
+\def\proof{\noindent {\it Proof. $\, $}}
+\def\endproof{\hfill $\Box$ \vskip 5 pt }
+
+\mdfdefinestyle{commandline}{
+	leftmargin=10pt,
+	rightmargin=10pt,
+	innerleftmargin=15pt,
+	middlelinecolor=black!50!white,
+	middlelinewidth=2pt,
+	frametitlerule=false,
+	backgroundcolor=black!5!white,
+	frametitle={LLM Output},
+	frametitlefont={\normalfont\sffamily\color{white}\hspace{-1em}},
+	frametitlebackgroundcolor=black!50!white,
+	nobreak,
+	singleextra={%
+    }
+}
+
+% Define a custom environment for command-line snapshots
+\newenvironment{commandline}{
+	\medskip
+	\begin{mdframed}[style=commandline]
+}{
+	\end{mdframed}
+	\medskip
+}
+
+\mdfdefinestyle{warning}{
+	topline=false, bottomline=false,
+	leftline=false, rightline=false,
+	nobreak,
+	singleextra={%
+		\draw(P-|O)++(-0.5em,0)node(tmp1){};
+		\draw(P-|O)++(0.5em,0)node(tmp2){};
+		\fill[black,rotate around={45:(P-|O)}](tmp1)rectangle(tmp2);
+		\node at(P-|O){\color{white}\scriptsize\bf ?};
+		\draw[very thick](P-|O)++(0,-1em)--(O);%--(O-|P);
+	}
+}
+
+% Define a custom environment for warning text
+\newenvironment{prompt}[1][Prompt:]{ % Set the default warning to "Warning:"
+	\medskip
+	\begin{mdframed}[style=warning]
+		\noindent{\textbf{#1}}
+}{
+	\end{mdframed}
+}
+
+\renewcommand{\lstlistingname}{Καταχώριση}
+
 \title{Αυτόματη διάγνωση ασθενειών με χρήση Μεγάλων Γλωσσικών Μοντέλων και Answer Set Programming}
-% \date{\today}
 \author{Ιωάννα Γέμου\thanks{Τμήμα Ηλεκτρολόγων Μηχανικών και Τεχνολογίας Υπολογιστών, Πανεπιστήμιο Πατρών} \\ \src{ioannagemou@upnet.gr} \and Ευάγγελος Λάμπρου\footnotemark[1] \\ \src{e.lamprou@upnet.gr}}
-% \institute{Πανεπιστήμιο Πατρών \\ Τμήμα Ηλεκτρολόγων Μηχανικών και % Τεχνολογίας Υπολογιστών} % Institute or school name
 
@@
 
-% Η αυτόματη πρόβλεψη ασθενειών βασίζεται στην επεξεργασία της ιατρικής βιβλιογραφίας, που περιέχει μεγάλο όγκο πληροφοριών για διάφορες ασθένειες και τα συμπτώματά τους, καθώς και τις βιολογικές ενδείξεις που σχετίζονται με αυτές.
-Μέσω της χρήσης των Large Language Models, καταφέρνουμε να μετατρέψουμε αυτήν τη γνώση σε κώδικα, χρησιμοποιώντας τη γλώσσα προγραμματισμού Clingo.
-Ουσιαστικά, τα μοντέλα γλώσσας μετατρέπουν τις πληροφορίες αυτές σε μια μορφή που μπορεί να επεξεργαστεί από το πρόγραμμα Answer Set Programming.
-
-% Το πρόγραμμά μας λειτουργεί με τον εξής τρόπο: δέχεται ως είσοδο συγκεκριμένα συμπτώματα τα οποία ένας ασθενής μπορεί να παρουσιάζει και αναλύει αυτά τα συμπτώματα με βάση την κωδικοποιημένη ιατρική γνώση.
-Μέσω του Answer Set Programming, το πρόγραμμα επιλέγει τις ασθένειες που είναι πιθανό να προκαλούν τα συμπτώματα αυτά και τις επιστρέφει ως αποτέλεσμα.
-
 Η καινοτομία της προτεινόμενης μεθόδου έγκειται στον συνδυασμό των δύο μεθοδολογιών. Τα Large Language Models μπορούν να κωδικοποιήσουν την ιατρική βιβλιογραφία 
@@
 την εφαρμόσαμε.
-
-% \section{Σκοπός της εργασίας}
-% Στην συγκεκριμένη εργασία προτείνουμε ένα νέο τρόπο αυτόματης πρόβλεψης ασθενειών με χρήση Large Language Models και Answer Set Programming.
-
-% H αυτόματη πρόβλεψη βασίζεται στην κωδικοποίηση της ιατρικής βιβλιογραφίας, δηλαδή την αντιστοίχηση των ασθενειών με συγκεκριμένα ιατρικά συμπτώματα και βιολογικές ενδείξεις.
-
-% Mε την χρήση των Μεγάλων Γλωσσικών Μοντέλων καταφέρνουμε να κωδικοποιήσουμε 
-% την αντιστοίχηση αυτή σε κώδικα \textit{Clingo} με αποτέλεσμα το πρόγραμμά μας, έχοντας ως είσοδο συγκεκριμένα συμπτώματα, να προβλέπει την ασθένεια πίσω από αυτά.
-Ο συνδυασμός των δύο αυτών μεθοδολογιών αξιοποιεί τις ικανότητες \say{μεταμόρφωσης} των μεγάλων
-% γλωσσικών μοντέλων σε συνδυασμό με την επεξηγηματικότητα της συμβολικής τεχνητής νοημοσύνης.
 
@@
 Αν και η απλούστερη μορφή αυτής της συνάρτησης είναι ένα σταθμισμένο άθροισμα των τιμών εισόδου, ο μηχανισμός προσοχής στον μετασχηματιστή είναι πιο περίπλοκος και περιλαμβάνει πολλαπλούς μετασχηματισμούς για τον υπολογισμό των βαρών προσοχής μεταξύ διαφορετικών τμημάτων της εισόδου.
-
-% Αν θεωρήσουμε ως είσοδο στο δίκτυο τον πίνακα $\textbf{x} \in \Re^ {d_m\times d}$ ($n$ ο αριθμός των αντικειμένων εισόδου και $d$ η διάστασή τους), οι σχέσεις που εκφράζουν τον μηχανισμό προσοχής είναι:
-
-% \begin{equation}
-%     Q(x_i) = x_i W_q 
-% \end{equation}
-
-% \begin{equation}
-%     K(x_i) = x_i W_k
-% \end{equation}
-
-% \begin{equation}
-%     V(x_i) = x_i W_v
-% \end{equation}
-
-% όπου: $W_q, W_k, \in \Re^{d\times d_v} $
-
-% \begin{equation}
-%     attention(Q, K, V) = softmax\frac{Q K ^ \top}{\sqrt{d_k}} V
-% \end{equation}
-
-% όπου:
-
-% \begin{equation}
-% \begin{bmatrix}
-% Q(\textbf{x}_1) \\
-% \vdots \\
-% Q(\textbf{x}_n) \\
-% \end{bmatrix}
-% \in \Re ^{n \times d_q}
-% \end{equation}
-
-% \begin{equation}
-% \begin{bmatrix}
-% K(\textbf{x}_1) \\
-% \vdots \\
-% K(\textbf{x}_n) \\
-% \end{bmatrix}
-% \in \Re ^{n \times d_k}
-% \end{equation}
-
-% \begin{equation}
-% \begin{bmatrix}
-% V(\textbf{x}_1) \\
-% \vdots \\
-% V(\textbf{x}_n) \\
-% \end{bmatrix}
-% \in \Re ^{n \times d_v}
-% \end{equation}
-
-% Οι πίνακες $Q, K$ και $V$ συμβολίζονται ως ερωτήματα, κλειδιά και τιμές, αντίστοιχα, και οι εγγραφές τους λαμβάνονται πολλαπλασιάζοντας τους πίνακες $W_q, W_k$ και $W_v$ με όλα τα διανύσματα εισόδου $x_i$, όπου το $i$ κυμαίνεται από $1$ έως $n$.
-Κατά τη διάρκεια της εκπαίδευσης, προσδιορίζονται οι πίνακες βάρους $W_q, W_k$ και $W_v$ και τα διανύσματα εισόδου $x_i$ είναι γνωστά ως \textbf{input embeddings}.
-
-% Κάθε είσοδος $x_i$ μετατρέπεται σε έναν πίνακα $Q(x_i)$.
-Στον μηχανισμό προσοχής, η συμβατότητα όλων των πινάκων $Q(x_i)$ με όλους τους βασικούς πίνακες $K(x_j)$ (όπου $j$ κυμαίνεται από 1 έως $n$ και $j$ δεν ισούται με $i$) υπολογίζεται από το εσωτερικό γινόμενο $QK ^\top$.
-Η προσοχή του δικτύου αυξάνεται με μεγαλύτερες τιμές του προϊόντος $QK ^\top$.
-
-% Η σχέση 4 υπολογίζει τις κανονικοποιημένες τιμές του προϊόντος $QK ^\top$ για μια μεμονωμένη αναπαράσταση των $Q, K, V$.
-Η κανονικοποίηση κατά $ \sqrt{d_k}$ εφαρμόζεται για να αποτρέψει το εσωτερικό προϊόν $QK ^\top$ από το να γίνει πολύ μεγάλο και να προκαλέσει αστάθεια στην εκπαίδευση.
-Για να εξαχθούν περισσότερες πληροφορίες από την είσοδο, το μοντέλο δημιουργεί διαφορετικές αναπαραστάσεις των πινάκων $Q, K, V$ πολλαπλασιάζοντάς τους με $h$ διαφορετικούς πίνακες βάρους.
-Οι προκύπτουσες έξοδοι συνδυάζονται και διέρχονται από ένα τελικό γραμμικό στρώμα, όπως απεικονίζεται στο σχήμα και εκφράζεται με τις εξισώσεις 8 και 9.
-Αυτό αναφέρεται συνήθως ως "multi-head attention" μηχανισμός στην ερευνητική κοινότητα.
-Σε κάθε βήμα εκμάθησης, το μοντέλο ενημερώνει τις αναπαραστάσεις ερωτήματος, κλειδιού και τιμών για να βελτιώσει την απόδοσή του.
-
-% \begin{equation}
-%     head(i) = attention((W^Q_i Q^T, W^k_i K^T, W^V_i V^T) 
-% \end{equation}
-% \begin{equation}
-%     MLH = \text{Concat}(head_1, \ldots, head_h)W^o
-% \end{equation}
-
-% \begin{figure}[htb]
-%     \centering
-%     \includegraphics[width=\textwidth]{assets/multihead.png}
-%     \caption{Η αριστερή εικόνα δείχνει τη ροή του μηχανισμού "multi-head attention"  και η δεξιά τα λεπτομερή στάδια του κλιμακούμενου dot product.
-Ως $Q, K, V$ συμβολίζονται οι πίνακες που περιέχουν το ερώτημα, κλειδί, τιμή αντίστοιχα.\cite{Vaswani2017}}
-%     \label{fig:multihead}
-% \end{figure}
 
@@
 \begin{verbatim}
-% Symptoms of Pneumonia
 
@@
 
-% Diagnosis of pneumonia
 diagnosis(pneumonia) :- symptom(cough), 
@@
 \begin{verbatim}
-% Symptoms of Pneumonia
 symptom(cough).
@@
 
-% Rules linking symptoms
-
-% Cough with mucus can be associated with wheezing and 
-% shortness of breath
 linked_symptom(cough_with_mucus, wheezing).
@@
 
-% Rule to infer linked symptoms
 has(symptom(Symptom2)) :- has(symptom(Symptom1)), 
@@
 
-% Diagnosis of Pneumonia
 diagnosis(pneumonia) :- has(symptom(cough)),
@@
 
-% Alternative Diagnoses for Pneumonia
 diagnosis(pneumonia) :- has(symptom(cough_with_mucus)),

### 2025-05-28 paper/draft.tex
Trnslate text
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

@@
 
-\title{Αυτόματη διάγνωση ασθενειών με χρήση Μεγάλων Γλωσσικών Μοντέλων και Answer Set Programming}
-\author{Ιωάννα Γέμου\thanks{Τμήμα Ηλεκτρολόγων Μηχανικών και Τεχνολογίας Υπολογιστών, Πανεπιστήμιο Πατρών} \\ \src{ioannagemou@upnet.gr} \and Ευάγγελος Λάμπρου\footnotemark[1] \\ \src{e.lamprou@upnet.gr}}
+\title{Automatic Disease Diagnosis Using Large Language Models and Answer Set Programming}
+\author{Ioanna Gemou\thanks{Department of Electrical and Computer Engineering, University of Patras} \\ \src{ioannagemou@upnet.gr} \and Evangelos Lamprou\footnotemark[1] \\ \src{e.lamprou@upnet.gr}}
 
@@
 \maketitle
-\tableofcontents
-\newpage
-\section{Περίληψη}
-Η ανάπτυξη αυτόματων συστημάτων πρόβλεψης ασθενειών έχει κρίσιμη σημασία για τη βελτίωση της ανθρώπινης υγείας και την υποστήριξη των ιατρών στην ακριβή διάγνωση και αντιμετώπιση των ασθενειών.
-Η ακριβής διάγνωση είναι κρίσιμη για την έγκαιρη αντιμετώπιση και αποτελεσματική θεραπεία, ενώ μπορεί επίσης να συμβάλει στην αποφυγή περαιτέρω επιπλοκών και επιβαρυντικών ιατρικών διαδικασιών.
-Η συμβολική τεχνητή νοημοσύνη έχει ξαναεφαρμοστεί στο πεδίο της ιατρικής \cite{Alviano_2020}. Ωστόσο,
-εφαρμογή πάντα ήταν περιορισμένη λόγω των σύνθετων βάσεων γνώσης που θα έπρεπε να αναπτυχθούν για τη λειτουργία 
-αυτών των συστημάτων.
-Η προτεινόμενη μέθοδος, η οποία συνδυάζει τα Large Language Models και το Answer Set Programming (ASP), αποτελεί ένα ισχυρό εργαλείο, καθώς παρέχει ένα αξιόπιστο και γρήγορο μέσο δημιουργίας μιας βάσης γνώσης η οποία
-στη συνέχεια μπορεί να χρησιμοποιηθεί για την πρόβλεψη ασθενειών.
+\section{Abstract}
+The development of automatic disease prediction systems is critically important for improving human health and supporting physicians in accurate diagnosis and treatment.
+Accurate diagnosis is essential for timely intervention and effective therapy, and it can also help prevent further complications and burdensome medical procedures.
 
-Η καινοτομία της προτεινόμενης μεθόδου έγκειται στον συνδυασμό των δύο μεθοδολογιών. Τα Large Language Models μπορούν να κωδικοποιήσουν την ιατρική βιβλιογραφία 
-σε κώδικα ASP με αξιοσημείωτη ακρίβεια και σύνθετη κατανόηση, ενώ το ASP παρέχει τον συλλογισμό και την επεξηγηματικότητά του.
-Ο συνδυασμός αυτών των δύο προσεγγίσεων επιτρέπει την ανάπτυξη ενός ισχυρού συστήματος πρόβλεψης ασθενειών που συνδυάζει την \textbf{ακρίβεια και την ευφυΐα των γλωσσικών μοντέλων} με την \textbf{επεξηγηματικότητα και την αποδοτικότητα του Answer Set Programming}.
+Symbolic artificial intelligence has previously been applied in the medical domain \cite{Alviano_2020}.
+However, its adoption has been limited due to the complex knowledge bases required for such systems to operate.
 
-Η μέθοδός μας φαίνεται να έχει υποσχόμενα αποτελέσματα πάνω στο σύνολο των δεδομένων ασθενών, στο οποίο
-την εφαρμόσαμε.
+The proposed method, which combines Large Language Models (LLMs) with Answer Set Programming (ASP), offers a powerful approach by providing a reliable and efficient means of generating a knowledge base that can then be used for disease prediction.
 
-\section{Ανασκόπηση Βιβλιογραφίας}
+The novelty of our method lies in the combination of these two paradigms.
+LLMs are capable of encoding medical literature into ASP code with remarkable accuracy and deep comprehension, while ASP contributes structured reasoning and explainability.
+This combination enables the development of a robust disease prediction system that integrates the \textbf{accuracy and intelligence of language models} with the \textbf{explainability and efficiency of Answer Set Programming}.
+
+Our method demonstrates promising results on the patient dataset to which it was applied.
+
+\section{Literature Review}
 
@@
 
-Τα μεγάλα γλωσσικά μοντέλα \cite{zhao2023survey} είναι προηγμένα υπολογιστικά μοντέλα που έχουν σχεδιαστεί για να κατανοούν και να δημιουργούν ανθρώπινη γλώσσα.
-Αυτά τα μοντέλα, που αναπτύχθηκαν χρησιμοποιώντας εξελιγμένες τεχνικές μηχανικής μάθησης, έχουν τη δυνατότητα να επεξεργάζονται και να δημιουργούν κείμενο με τρόπο που μοιάζει πολύ με τα πρότυπα ανθρώπινης γλώσσας.
+Large language models \cite{zhao2023survey} are advanced computational models designed to understand and generate human language.
+These models, developed using sophisticated machine learning techniques, have the ability to process and generate text in a way that closely resembles human language patterns.
 
-Μεγάλα γλωσσικά μοντέλα, όπως η αρχιτεκτονική \textit{GPT-4} \cite{openai2023gpt4}, εκπαιδεύονται σε τεράστιες ποσότητες δεδομένων κειμένου, γεγονός που τους επιτρέπει να αποκτούν γνώση σε διάφορους τομείς και θέματα.
-Αυτά τα μοντέλα χρησιμοποιούν αλγόριθμους βαθιάς μάθησης (deep learning) \cite{Sarker2021} και ειδικά δίκτυα μετασχηματιστών (transformer networks) \cite{Dosovitskiy2020}, για την ανάλυση και την κατανόηση των σχέσεων μεταξύ λέξεων, φράσεων και προτάσεων.
-Καταγράφοντας μοτίβα και δομές στα δεδομένα εκπαίδευσης, αυτά τα μοντέλα μπορούν να δημιουργήσουν συνεκτικές και συναφείς
-απαντήσεις με βάση δοθέντα prompts.
+Large language models, such as the \textit{GPT-4} architecture \cite{openai2023gpt4}, are trained on vast amounts of text data, allowing them to acquire knowledge across various fields and topics.
+These models use deep learning algorithms \cite{Sarker2021} and specifically transformer networks \cite{Dosovitskiy2020} to analyze and understand the relationships between words, phrases, and sentences.
+By capturing patterns and structures in the training data, these models can generate coherent and relevant responses based on given prompts.
 
-Tα μοντέλα ονομάζονται \say{μεγάλα} επειδή έχουν δισεκατομμύρια παραμέτρους που διαμορφώνουν τις αποκρίσεις τους.
-Η βασική προϋπόθεση ενός γλωσσικού μοντέλου είναι η ικανότητά του να προβλέπει την επόμενη λέξη (που ονομάζεται token) με βάση το κείμενο που έχει παρατηρήσει μέχρι τώρα, δηλαδή σύμφωνα με τα συμφραζόμενα.
+The models are referred to as "large" because they have billions of parameters that shape their responses.
+The basic premise of a language model is its ability to predict the next word (referred to as a token) based on the text it has observed so far, according to the context.
 
-Τα μεγάλα γλωσσικά μοντέλα διαθέτουν την αξιοσημείωτη ικανότητα να μετασχηματίζουν κείμενο από τη μια μορφή στην άλλη, καθιστώντας τα ισχυρά εργαλεία για διάφορες εργασίες επεξεργασίας γλώσσας. 
-Πάνω σε αυτό το χαρακτηριστικό τους, βασίζεται και η ιδέα αυτής της εργασίας, αφού ένα πολύ σημαντικό κομμάτι που πραγματευόμαστε είναι η κατάλληλη κωδικοποίηση της ιατρικής βιβλιογραφίας.
-Παρακάτω αναφέρουμε κάποια παραδείγματα χρήσης των μεγάλων γλωσσικών μοντέλων \cite{Paranjape2023}:
+Large language models possess the remarkable ability to transform text from one form to another, making them powerful tools for various language processing tasks.
+This characteristic underpins the idea of this work, as a significant part we deal with is the appropriate encoding of medical literature.
+Below we list some examples of the use of large language models \cite{Paranjape2023}:
 
-\begin{itemize}
-    \item Δημιουργία κειμένου που είναι συναφές και συνεκτικό με βάση τα συμφραζόμενα
-    \item Μετάφραση Γλώσσας
-    \item Σύνοψη / Περίληψη Κειμένου
-    \item  Απάντηση σε ερωτήσεις
-    \item Παραγωγή κειμένου με βάση συγκεκριμένων κανόνων που τους παρέχονται
-\end{itemize}
+\subsection{Difficulties in Implementing LLMs}
 
-Ο θεμέλιος λίθος των Μεγάλων Γλωσσικών μοντέλων είναι οι \textbf{transformers}.
-Οι transformers \cite{Vaswani2017} επί του παρόντος θεωρούνται μία από τις πιο προηγμένες και ισχυρές κατηγορίες μοντέλων.
-Συνέβαλαν καθοριστικά στην επίτευξη σημαντικών ανακαλύψεων στη μηχανική μάθηση, με αποτέλεσμα ορισμένοι να τους αναφέρουν ως \say{transformers AI}.
-Aυτά τα μοντέλα οδηγούν σε μια θεμελιώδη αλλαγή στον τομέα της τεχνητής νοημοσύνης \cite{Bommasani2021}.
-To σχήμα \ref{fig:transformers} απεικονίζει τη δομή του transformer, ο οποίος περιλαμβάνει δύο διακριτά μέρη, την κωδικοποίηση και την αποκωδικοποίηση.
+Large language models (LLMs) are not without errors \cite{Raj2023}, \cite{Ruis2023}.
+They can occasionally produce incorrect or biased responses that reflect the biases inherent in the training data.
+Efforts are being made to mitigate these issues through data preprocessing techniques and fine-tuning.
 
-\begin{figure}[htb]
-    \centering
-    \includegraphics[width=0.7\textwidth]{assets/transformers_architecture.png}
-    \caption{Αρχιτεκτονική του μοντέλου transformer \cite{Vaswani2017}.}
-    \label{fig:transformers}
-\end{figure}
-
-\subsubsection{Inputs and Input Embeddings}
-
-Tο θεμελιώδες συστατικό του
-transformer, ο μηχανισμός \say{προσοχής} είναι εμπνευσμένος από την έννοια της συνειδητής προσοχής στους ζωντανούς οργανισμούς \cite{Vaswani2017}.
-Αυτός ο μηχανισμός επιτρέπει στο δίκτυο να αποδίδει μεγαλύτερη σημασία σε κρίσιμα μέρη των δεδομένων και λιγότερη σημασία σε αυτά που είναι λιγότερο σχετικά.
-Αν και η απλούστερη μορφή αυτής της συνάρτησης είναι ένα σταθμισμένο άθροισμα των τιμών εισόδου, ο μηχανισμός προσοχής στον μετασχηματιστή είναι πιο περίπλοκος και περιλαμβάνει πολλαπλούς μετασχηματισμούς για τον υπολογισμό των βαρών προσοχής μεταξύ διαφορετικών τμημάτων της εισόδου.
-
-Τα τμήματα εισόδου από τον χρήστη είναι είσοδοι για τα μοντέλα μηχανικής μάθησης.
-Ωστόσο, τα μοντέλα κατανοούν μόνο αριθμούς, όχι κείμενο, επομένως αυτές οι είσοδοι πρέπει να μετατραπούν σε μια αριθμητική μορφή που ονομάζεται \textbf{input embeddings}.
-Tα input embeddings αντιπροσωπεύουν λέξεις ως αριθμούς, τους οποίους μπορούν στη συνέχεια να επεξεργαστούν τα μοντέλα μηχανικής μάθησης.
-Τα embeddings είναι σαν ένα λεξικό, το οποίο βοηθά το μοντέλο να κατανοήσει τη σημασία των λέξεων, μέσω της τοποθέτησής τους σε έναν μαθηματικό χώρο όπου παρόμοιες λέξεις βρίσκονται η μία κοντά στην άλλη.
-Κατά τη διάρκεια της εκπαίδευσης, το μοντέλο μαθαίνει πώς να δημιουργεί τα embeddings, έτσι ώστε παρόμοια διανύσματα να αντιπροσωπεύουν λέξεις με παρόμοια σημασία.
-
-\subsubsection{Positional Encodings}
-
-Στην επεξεργασία φυσικής γλώσσας, η σειρά των λέξεων σε μια πρόταση είναι κρίσιμη για τον προσδιορισμό του νοήματος της πρότασης.
-Ωστόσο, τα παραδοσιακά μοντέλα μηχανικής μάθησης, όπως τα νευρωνικά δίκτυα, δεν κατανοούν εγγενώς τη σειρά των εισόδων.
-Για την αντιμετώπιση αυτής της πρόκλησης, η κωδικοποίηση θέσης μπορεί να χρησιμοποιηθεί για την κωδικοποίηση της θέσης κάθε λέξης σε μία πρόταση.
-Η κωδικοποίηση αυτή τροφοδοτείται στο μοντέλο Transformer, μαζί με τα input embeddings.
-Με την ενσωμάτωση της κωδικοποίησης θέσης στην αρχιτεκτονική του Transformer, το μοντέλο μπορεί να κατανοήσει πιο αποτελεσματικά τη σειρά των λέξεων σε μια πρόταση και να δημιουργήσει γραμματικά σωστά και σημασιολογικά νόημα.
-Το \say{γραμματικά σωστό} αποτέλεσμα μπορεί να σημαίνει και την παραγωγή κώδικα ο οποίος είναι 
-σύμφωνος με τους κάνόνες μιας γλώσσας προγραμματισμού.
-
-\subsubsection{Encoder}
-
-Ο κωδικοποιητής (encoder) είναι μέρος του νευρωνικού δικτύου που επεξεργάζεται το κείμενο εισόδου και δημιουργεί μια σειρά από κρυφές καταστάσεις που καταγράφουν το νόημα του κειμένου.
-Ο κωδικοποιητής διαμορφώνει πρώτα το κείμενο εισόδου σε μια ακολουθία tokens, π.χ.
-μεμονωμένες λέξεις.
-
-\subsubsection{Decoder}
-Ο αποκωδικοποιητής (decoder) είναι μέρος του μοντέλου που δημιουργεί την ακολουθία εξόδου με βάση την κωδικοποιημένη ακολουθία εισόδου.
-Κατά τη διάρκεια της εκπαίδευσης, ο αποκωδικοποιητής μαθαίνει πώς να \say{μαντεύει} την επόμενη λέξη κοιτάζοντας τις λέξεις πριν από αυτήν.
-Για παράδειγμα, ο αποκωδικοποιητής σε \textit{GPT} δημιουργεί κείμενο φυσικής γλώσσας με βάση την ακολουθία εισόδου και το περιβάλλον που μαθαίνει ο κωδικοποιητής.
-Στους transformers χρησιμοποιούνται πολλαπλά στρώματα αποκωδικοποιητών.
-
-\subsection{Δυσκολίες στην εφαρμογή των LLMs}
-
-Τα μεγάλα γλωσσικά μοντέλα δεν είναι αλάνθαστα \cite{Raj2023}, \cite{Ruis2023}.
-Μπορούν περιστασιακά να παράγουν λανθασμένες ή μεροληπτικές απαντήσεις, αντανακλώντας τις προκαταλήψεις που υπάρχουν στα δεδομένα εκπαίδευσης.
-Καταβάλλονται προσπάθειες για τον μετριασμό αυτών των ζητημάτων μέσω τεχνικών προεπεξεργασίας δεδομένων και fine-tuning.
-
-Συνολικά, τα μεγάλα γλωσσικά μοντέλα έχουν επιδείξει τεράστιες δυνατότητες σε ένα ευρύ φάσμα εφαρμογών, συμπεριλαμβανομένης της κατανόησης φυσικής γλώσσας, της δημιουργίας περιεχομένου και της αλληλεπίδρασης ανθρώπου-υπολογιστή.
-Η συνεχής έρευνα και ανάπτυξη σε αυτόν τον τομέα υπόσχεται περαιτέρω ενίσχυση αυτών των μοντέλων και των εφαρμογών τους σε διάφορους τομείς, συμβάλλοντας στην πρόοδο των τεχνολογιών γλωσσικής επεξεργασίας και επικοινωνίας.
+Overall, large language models have demonstrated tremendous potential across a wide range of applications, including natural language understanding, content creation, and human-computer interaction.
+Ongoing research and development in this field promise further enhancement of these models and their applications across various domains, contributing to the advancement of language processing and communication technologies.
 
 \subsection{Answer Set Programming}
-To Answer Set Programming (ASP) \cite{Eiter2009} είναι μια δηλωτική μέθοδος επίλυσης προβλημάτων που βασίζεται στον Λογικό Προγραμματισμό και τη Μη Μονοτονική Συλλογιστική.  
-Στην εργασία \cite{Gelfond2000} επισημοποιήθηκε αρχικά η σημασιολογία των σταθερών μοντέλων και η βασική γλώσσα του ASP.
 
-Ο προγραμματισμός με χρήση αυτής της προσέγγισης γίνεται σε μια οικογένεια γλωσσών που ονομάζεται ορισμένες φορές \textit{AnsProlog} \cite{Gelfond2002}.
+Answer Set Programming (ASP) \cite{Eiter2009} is a declarative problem-solving approach based on Logic Programming and Non-Monotonic Reasoning.
+The semantics of stable models and the basic language of ASP were initially formalized in the work \cite{Gelfond2000}.
 
-Η ιδέα πίσω από το ASP είναι να μοντελοποιηθεί ένα πρόβλημα ως ένα σύνολο κανόνων και δεδομένων, και στη συνέχεια να χρησιμοποιηθεί ένας επίλυτης για να βρεθεί μια λύση για το πρόβλημα.
-Οι λύσεις παριστάνονται από σταθερά μοντέλα ή σύνολα απαντήσεων.
-Οι κανόνες, τα δεδομένα και οι περιορισμοί που περιγράφουν το πρόβλημα αποτελούν τα στοιχεία του προγράμματος.
-Στη συνέχεια, το πρόγραμμα δίνεται σε έναν επίλυτη που θα βρει μία ή περισσότερες λύσεις.
+Programming with this approach is done in a family of languages sometimes referred to as \textit{AnsProlog} \cite{Gelfond2002}.
 
-Στον παραδοσιακό προγραμματισμό, η μετάβαση από ένα πρόβλημα σε μια λύση προϋποθέτει την απόλυτη κατανόηση του δεδομένου προβλήματος από τον προγραμματιστή, και έπειτα την παραγωγή ενός προγράμματος, στο οποίο όταν δοθεί ένα instance του προβλήματος θα παράγει το σωστό αποτέλεσμα, τα οποίο στη συνέχεια θα ερμηνευτεί ως λύση.
+The idea behind ASP is to model a problem as a set of rules and data, and then use a solver to find a solution to the problem.
+Solutions are represented by stable models or answer sets.
+The rules, data, and constraints that describe the problem are the elements of the program.
+Subsequently, the program is given to a solver that will find one or more solutions.
 
-Στο ASP, η μετάβαση από τη δήλωση προβλήματος σε ένα σύνολο λύσεων περιλαμβάνει τα ακόλουθα βήματα \cite{Gebser2013}
-(εικόνα \ref{fig:asp-solving}):
+In traditional programming, transitioning from a problem to a solution requires the programmer's thorough understanding of the given problem, followed by the creation of a program that will produce the correct output when provided with an instance of the problem, which will then be interpreted as the solution.
+
+In ASP, the transition from problem statement to a set of solutions involves the following steps \cite{Gebser2013}~(\cref{fig:asp-solving}):
 
@@
     \end{center}
-    \caption{Η επίλυση ενός προβλήματος με χρήση ASP.}
+		\caption{Solving a problem using ASP.}
     \label{fig:asp-solving}
@@
 \begin{itemize}
-    \item \textbf{Modeling}: Το πρόβλημα μοντελοποιείται σε γλώσσα ASP.
-    \item \textbf{Grounding}: Ένας grounder (π.χ. gringo) μετατρέπει το πρόγραμμα σε ένα σύνολο βασικών κανόνων και δεδομένων.
-    \item \textbf{Solving}: Ένας επιλυτής (π.χ. clasp) βρίσκει μια λύση στο πρόβλημα υπολογίζοντας το σύνολο των σταθερών μοντέλων (σύνολα απαντήσεων).
+    \item \textbf{Modeling}: The problem is modeled using ASP (Answer Set Programming) language.
+    \item \textbf{Grounding}: A grounder (e.g., gringo) converts the program into a set of basic rules and facts.
+    \item \textbf{Solving}: A solver (e.g., clasp) finds a solution to the problem by computing the set of stable models (answer sets).
 \end{itemize}
 
-Έχουν αναπτυχθεί αρκετοί λύτες για ASP όπως το DLV \cite{Xia2020}.
-Σε αυτή την εργασία, θα εφαρμόσουμε το σύστημα ASP clingo \cite{Gebser2014} το οποίο συνδυάζει τον grounder gringo με τον clasp solver\cite{Holldobler2014} σε μια ενιαία εφαρμογή, παρέχοντας επίσης ένα ισχυρό Python API για την ενσωμάτωση του solver σε άλλες εφαρμογές.
+Several solvers for ASP have been developed, such as DLV \cite{Xia2020}.
+In this work, we will apply the ASP system clingo \cite{Gebser2014}, which combines the grounder gringo with the clasp solver \cite{Holldobler2014} into a single application, also providing a powerful Python API for integrating the solver into other applications.
 
-\subsection{Χρήση του ASP στην διάγνωση ασθενειών}
-Το Answer Set Programming (ASP) προσφέρει μια ισχυρή μέθοδο για την αναπαράσταση της ιατρικής γνώσης με δηλωτικό και λογικό τρόπο.
-Ακολουθούν ορισμένες μέθοδοι ASP που χρησιμοποιούνται συνήθως στην αναπαράσταση ιατρικής γνώσης \cite{Vinarti2019}:
+\subsection{Use of ASP in Disease Diagnosis}
+Answer Set Programming (ASP) offers a powerful method for representing medical knowledge in a declarative and logical manner.
+Below are some common ASP methods used in the representation of medical knowledge \cite{Vinarti2019}:
 
 \begin{itemize}
-    \item \textbf{Αναπαράσταση βάσει κανόνων}: Το ASP επιτρέπει την αναπαράσταση της ιατρικής γνώσης ως κανόνων με τη μορφή λογικών δηλώσεων.
-Αυτοί οι κανόνες ορίζουν σχέσεις μεταξύ ιατρικών οντοτήτων όπως συμπτώματα, ασθένειες, θεραπείες και δεδομένα ασθενών.
-Για παράδειγμα, ένας κανόνας θα μπορούσε να αναφέρει ότι εάν ένας ασθενής εμφανίζει συγκεκριμένα συμπτώματα, υποδηλώνει την παρουσία ή την απουσία ορισμένων ασθενειών.
-
-    \item \textbf{Κατασκευή Γνωσιακής Βάσης}: Το ASP παρέχει ένα πλαίσιο για την κατασκευή μιας βάσης γνώσεων που συλλαμβάνει την ιατρική γνώση.
-Η βάση γνώσεων αποτελείται από γεγονότα, κανόνες και περιορισμούς που καθορίζουν τις πληροφορίες και τους περιορισμούς που σχετίζονται με την ιατρική διάγνωση και θεραπεία.
-Μπορεί να περιλαμβάνει πληροφορίες σχετικά με τις σχέσεις ασθένειας-συμπτωμάτων, διαγνωστικά κριτήρια, κατευθυντήριες γραμμές θεραπείας και δεδομένα για τον ασθενή.
-
+    \item \textbf{Rule-Based Representation}: ASP allows for the representation of medical knowledge as rules in the form of logical statements.
+These rules define relationships between medical entities such as symptoms, diseases, treatments, and patient data.
+For example, a rule could state that if a patient exhibits specific symptoms, it indicates the presence or absence of certain diseases.
+    \item \textbf{Knowledge Base Construction}: ASP provides a framework for constructing a knowledge base that captures medical knowledge.
+The knowledge base consists of facts, rules, and constraints that define the information and limitations related to medical diagnosis and treatment.
+It may include information about disease-symptom relationships, diagnostic criteria, treatment guidelines, and patient data.
 \end{itemize}
 
-Στην εργασία \cite{Erdem2011} εισήχθησαν νέες μέθοδοι για αποτελεσματικό χειρισμό σύνθετων ερωτημάτων στο πλαίσιο βιοϊατρικών οντολογιών και βάσεων δεδομένων, με έμφαση στην εξαγωγή σχετικών πληροφοριών από αυτούς τους πόρους γνώσης.
-Aυτές οι μέθοδοι εφαρμόστηκαν για την αντιμετώπιση απαιτητικών ερωτημάτων στον τομέα της ανακάλυψης φαρμάκων, χρησιμοποιώντας βιολογικές βάσεις δεδομένων σε συνδυασμό με συνοπτικές εξηγήσεις για πολύπλοκα ερωτήματα που σχετίζονται με διαδικασίες ανακάλυψης φαρμάκων.
+In the work \cite{Erdem2011}, new methods were introduced for effectively handling complex queries within the context of biomedical ontologies and databases, with an emphasis on extracting relevant information from these knowledge resources.
+These methods were applied to address demanding queries in the field of drug discovery, using biological databases in conjunction with concise explanations for complex queries related to drug discovery processes.
 
-\section{Μεθοδολογία}
+\section{Methodology}
 
-\subsection{Αρχιτεκτονική}
+\subsection{Architecture}
 
@@
 \end{center}
-    \caption{Η διαδικασία της διάγνωσης με τη χρήση της προτεινόμενης μεθόδου.
-    Ιατρική βιβλιογραφία, με χρήση ενός μεγάλου γλωσσικού μοντέλου, μετατρέπεται σε ένα πρόγραμμα ASP.
-    Στη συνέχεια, το πρόγραμμα αυτό εκτελείται από έναν ASP solver, ο οποίος παράγει την τελική διάγνωση
-    με βάση τα δεδομένα για τον εκάστοτε ασθενή.}
+\caption{The diagnostic process using the proposed method. Medical literature,
+	through the use of a large language model, is transformed into an ASP
+	program. Subsequently, this program is executed by an ASP solver, which
+	produces the final diagnosis based on the data of each patient.}
     \label{fig:architecture}
@@
 
-Σκοπός μας είναι η εξαγωγή μίας βάσης γνώμης μέσα από αδώμητο κείμενο \cite{ré2014feature}.
-Η αρχιτεκτονική μας (σχήμα \ref{fig:architecture}) περιλαμβάνει τα εξής στάδια
+Our goal is to extract a knowledge base from unstructured text \cite{ré2014feature}. Our architecture (Figure \ref{fig:architecture}) includes the following stages:
 
-\begin{enumerate}
-    \item Συλλογή ιατρικής βιβλιογραφίας σχετική με τις εξεταζόμενες ασθένειες/παθήσεις.
-    \item Κατασκευή κατάλληλου prompt για την μετατροπή ιατρικής βιβλιογραφίας σε κώδικα ASP.
-    \item Προσθήκη του παραγώμενου ASP κώδικα από το μεγάλο μοντέλο γλώσσας στη βάση γνώσης.
-    \item Προσθήκη σχετικών δεδομένων του ασθενή (συμπτώματα, αποτελέσματα ιατρικών εξετάσεων) στη βάση γνώσης.
-    \item Παραγωγή της τελικής διάγνωσης (ή σύνολο πιθανών διαγνώσεων) ως answer set(s), αφού ο solver συλλογιστεί 
-          πάνω στη τελική βάση γνώσης.
-\end{enumerate}
+\begin{itemize}
+	\item Collecting medical literature related to the diseases/conditions under examination.
+	\item Creating an appropriate prompt for converting medical literature into ASP code.
+	\item Adding the ASP code generated by the large language model to the knowledge base.
+	\item Adding relevant patient data (symptoms, medical examination results) to the knowledge base.
+	\item Producing the final diagnosis (or set of possible diagnoses) as answer set(s), after the solver reasons over the final knowledge base.
+\end{itemize}
 
-\subsection{Κατασκευή Prompt}
+\subsection{Prompt Construction}
 
-Τα μεγάλα γλωσσικά μοντέλα, παρά των υψηλών δυνατοτήτων τους στην εκτέλεση οδηγιών, ακόμα 
-έχουν χαμηλή ελεγξιμότητα \cite{bowman2023things}.
-Έτσι, έχει προκύψει μία σχετικά νέα μεθοδολογία \say{προγραμματισμού} γνωστή ως prompt engineering \cite{liu2021pretrain, promptingguidePromptEngineering}.
-Προκειμένου να μετατρέψουμε την ιατρική βιβλιογραφία σε κώδικα
-κατάλληλο για την εκτέλεσή του σε επιλυτή ASP έπρεπε να κατασκευάσουμε 
-ποιοτικά prompts ώστε ο παραγώμενος κώδικας να βρίσκεται στην επιθυμητή μορφή (σχήμα \ref{fig:bad-prompt}, σχήμα \ref{fig:good-prompt}).
+Despite their high capabilities in executing instructions, large language
+models still have low reliability \cite{bowman2023things}. Thus, a relatively
+new programming methodology known as prompt engineering has emerged
+\cite{liu2021pretrain, promptingguidePromptEngineering}. In order to convert
+medical literature into code suitable for execution in an ASP solver, we had to
+construct quality prompts so that the generated code is in the desired form
+(Figure \ref{fig:bad-prompt}, Figure \ref{fig:good-prompt}).
 
@@
 
-\caption{Ένα απλό prompt μέσω του οποίου δεν πέρνουμε ικανοποιητικά αποτελέσματα.}
+\caption{A simple prompt that does not yield satisfactory results.}
 \label{fig:bad-prompt}
@@
 \end{commandline}
-\caption{Δίνοντας αναλυτικότερες πληροφορίες μπορούμε να καθοδηγήσουμε το μεγάλο γλωσσικό μοντέλο στο να μετατρέπει την εισαγώμενη ιατρική βιβλιογραφία στην μορφή που θέλουμε.}
+\caption{A good prompt that yields satisfactory results.}
 \label{fig:good-prompt}
@@
 
-\subsection{Κατασκευή Συλλογιστικού Προγράμματος}
+\subsection{Construction of a Logical Program}
 
-Η βάση γνώσης αρχικά αποτελείται από ένα σύνολο συμπτωμάτων και ιατρικών ενδείξεων της μορφής:
+The knowledge base initially consists of a set of symptoms and medical indicators in the form:
 
@@
 
-Αποτυπώνουμε έναν ασθενή ο οποίος έχει κάποιο σύμπτωμα ή βγήκε θετικός σε κάποια εξέταση ως:
+We represent a patient who has a certain symptom or tested positive for a certain test as:
 
@@
 
-Οι λογικές προτάσεις με βάση τις οποίες το σύστημά μας θα βγάζει συμπεράσματα είναι της μορφής:
+The logical propositions based on which our system will draw conclusions are in the form:
 
@@
 
-Ακόμα, λαμβάνουμε υπόψιν τη συσχέτιση μεταξύ ορισμένων συμπτωμάτων.
-Παραδείγματος χάρη, σε ορισμένους ασθενείς έχει σημειωθεί σύνδεση  
-μεταξύ δερματικών ασθενειών και πονοκεφάλων\cite{migraine-hives}.
-Άλλες φορές, συμπτώματα μπορεί να οδηγήσουν στην εξέλιξη άλλων συμπτωμάτων.
-Τέτοιου είδους σχέσεις κωδικοποιούνται στη βάση γνώσης μας μέσω κανόνα της μορφής.
+We also take into account the correlation between certain symptoms.
+For example, there has been a noted connection between skin diseases and headaches in certain patients\cite{migraine-hives}.
+Sometimes, symptoms may lead to the development of other symptoms.
+Such relationships are encoded in our knowledge base through a rule of the form:
 
@@
 
-Τα αντίστοιχα δεδομένα θα είναι ως εξής
+The corresponding data will be as follows:
 
@@
 
-Ωστόσο, είναι αναμενόμενο πως κάποιος ασθενής δεν θα κατέχει \textit{όλα} τα καταγεγραμμένα συμπτώματα.
-Θέλουμε να δώσουμε έναν ελέγξιμο τρόπο το σύστημά μας να δίνει εναλλακτικές προβλέψεις
-στην περίπτωση όπου το ιατρικό προσωπικό δεν έχει όλες τις πληροφορίες για την κατάσταση του ασθενή.
-Έτσι, προσθέτουμε τον παρακάτω κανόνα επιλογής (choice rule), με τον οποίο προστίθεται 
-ένα σύνολο πιθανών συμπτώματων για τον ασθενή στη βάση γνώσης.
+However, it is expected that a patient will not exhibit \textit{all} recorded symptoms.
+We want to provide a controllable way for our system to make alternative predictions in cases where the medical staff does not have all the information about the patient's condition.
+Thus, we add the following choice rule, which adds a set of possible symptoms for the patient to the knowledge base.
 
@@
 
-Ακόμα, απαιτούμε από τον επιλυτή να βρει τουλάχιστον μία διάγνωση.
+Additionally, we require the solver to find at least one diagnosis.
 
@@
 
-Τέλος, ελαχιστοποιούμε τον αριθμό \say{ψεύτικων} διαγνώσεων, έτσι ώστε ο επιλυτής 
-να καταλήξει σε μία διάγνωση με όσο το δυνατόν λιγότερες υποθέσεις.
+Finally, we minimize the number of "false" diagnoses so that the solver arrives at a diagnosis with as few assumptions as possible.
 
@@
 
-Με αυτό τον τρόπο, το σύστημά μας εγγυάται πως θα παράξει μία τουλάχιστον πρόβλεψη, η οποία 
-θα βασίζεται όσο το δυνατόν περισσότερο στις έγγυρες πληροφορίες που υπάρχουν για τον ασθενή.
+This way, our system guarantees that it will produce at least one prediction, which will be based as much as possible on valid information available about the patient.
 
-\subsection{Επεξηγηματικότητα Διαγνώσεων}
+\subsection{Explainability of Diagnoses}
 
-Η συμβολική τεχνητή νοημοσύνη έχει ένα μείζον πλεονέκτημα έναντι των υποσυμβολικών προσεγγίσεων λόγω της εξηγηματικότητάς της.
-Οι σχέσεις μεταξύ των γεγονότων και των συμπερασμάτων που καταλήγει ένα σύστημα, το οποίο εκτελεί ένα συμβολικό μηχανισμό είναι σαφώς καθορισμένες.
-Ωστόσο, οι επίλυτές ASP όπως το \textit{Clingo} δεν παρέχουν\footnote{Τη στιγμή της συγγραφής της εργασίας αυτής.} εύληπτα αποτυπώματα των συλλογιστικών διαδικασιών τους.
-Έχει υπάρξει έρευνα στην σύνδεση λογικών προτάσεων 
-με τις αιτιολογήσεις τους \cite{cabalar2014causal}.
-Θεωρήστε το παρακάτω λογικό πρόγραμμα, η σύνδεση των λογικών προτάσεων με τις αιτίες τους παρουσιάζονται 
-σε μορφή γράφου στο σχήμα \ref{fig:causal-g}. Αυτή είναι η δομή που μας επιτρέπει 
-να εξηγήσουμε τα συμπεράσματα ενός επιλυτή συμβολικής λογικής.
+Symbolic artificial intelligence has a major advantage over subsymbolic approaches due to its explainability.
+The relationships between the facts and conclusions that a system, which executes a symbolic mechanism, reaches are clearly defined.
+However, ASP solvers like \textit{Clingo} do not provide\footnote{At the time of writing this paper.} intuitive traces of their reasoning processes.
+There has been research on linking logical propositions with their justifications \cite{cabalar2014causal}.
+Consider the following logic program; the connection of the logical propositions with their causes is presented in the form of a graph in Figure \ref{fig:causal-g}.
+This structure allows us to explain the conclusions of a symbolic logic solver.
 
@@
     \includegraphics{assets/causal_g.png}
-    \caption{Αιτιολογικός γράφος λογικού ποργράμματος (εξ. \ref{eq:example-cg}).}
+    \caption{Causal graph of logic program (eq. \ref{eq:example-cg}).}
     \label{fig:causal-g}
@@
 
-Το εργαλείο \textit{xclingo}\footnote{\url{https://github.com/bramucas/xclingo2}} \cite{Cabalar_2020} επιτρέπει την παρουσίαση αιτιολογήσεων σχετικά με τα αποτελέσματα που δίνονται από τον επιλυτή \textit{Clingo}.
-Αυτό επιτυγχάνεται προσθέτοντας ένα \src{trace} που αντιστοιχεί στους κανόνες του λογικού προγράμματός μας.
-Η παροχή μιας σαφούς εξηγήσεως για το πώς το σύστημα κατέληξε σε συγκεκριμένα συμπεράσματα, μπορεί να βοηθήσει στον εντοπισμό πιθανών σφαλμάτων κατά την δημιουργία της βάσης γνώσης και να παρέχει στον χρήστη του συστήματος μια αιτιολόγηση για το τελικό αποτέλεσμα.
-Αυτό είναι κάτι πολύ σημαντικό στην περίπτωση που η εφαρμογή χρησιμοποιείται σε ιατρικά πλαίσια.
+The tool \textit{xclingo}\footnote{\url{https://github.com/bramucas/xclingo2}} \cite{Cabalar_2020} allows the presentation of justifications regarding the results given by the solver \textit{Clingo}.
+This is achieved by adding a \src{trace} corresponding to the rules of our logic program.
+Providing a clear explanation of how the system reached specific conclusions can help identify potential errors in the creation of the knowledge base and provide the system user with a rationale for the final outcome.
+This is particularly important when the application is used in medical contexts.
 
-\section{Αποτελέσματα}
+\section{Results}
 
-Για την αξιολόγηση του συστήματός μας χρησιμοποίησαμε δημοσίως διαθέσιμα δεδομένα ασθενών\footnote{\url{https://www.kaggle.com/datasets/itachi9604/disease-symptom-description-dataset?select=dataset.csv}}.
-Στο σύστημά μπήκε σαν είσοδος το σύνολο των συμπτωμάτων του κάθε ασθενή.
-Σαν βάση γνώσης έχουμε τον συνδυασμό όλων των κανόνων που έχουν προκύψει από την κωδικοποίηση την ιατρικής βιβλιογραφίας.  
-Μία επιτυχημένη πρόβλεψη ορίζεται όταν η πραγματική ασθένεια του ασθενή συμπίπτει με την τελική
-πρόβλεψη του επιλυτή ASP.
+To evaluate our system, we used publicly available patient data\footnote{\url{https://www.kaggle.com/datasets/itachi9604/disease-symptom-description-dataset?select=dataset.csv}}.
+The symptoms of each patient were fed into the system as input.
+Our knowledge base consists of the combination of all rules derived from encoding the medical literature.
+A successful prediction is defined when the actual disease of the patient matches the final prediction of the ASP solver.
 
@@
     \hline
-    Ασθένεις & Μέγεθος Βάσης Γνώσης & Ακρίβεια \\
+    Diseases & Size of Knowledge Base & Accuracy \\
     \hline
@@
     \end{tabular}
-    \caption{Ακρίβεια της μεθόδου για μερικές ενδεικτικές ασθένειες.}
+    \caption{Accuracy of the method for some indicative diseases.}
 \end{table}
 
-Η ακρίβεια του μοντέλου μας μπορεί να φτάσει πολύ υψηλά ποσοστά ακρίβειας αλλά απαιτεί την εισαγωγή ενός μεγάλου πλήθους κανόνων στη βάση γνώσης.
-Το μέγεθος της βάσης γνώσης το ορίζουμε ως το πλήθος των λογικών όρων μέσα 
-στο ASP πρόγραμμα που αντιστοιχεί σε κάθε μία από τις ασθένειες.
+The accuracy of our model can reach very high rates but requires the input of a large number of rules into the knowledge base.
+We define the size of the knowledge base as the number of logical terms within the ASP program corresponding to each of the diseases.
 
-Στο πρόγραμμα που κατασκευάσαμε προσθέσαμε και τη δυνατότητα 
-επεξήγησης της διάγνωσης, κατά την οποία αξιοποιούμε 
-το εργαλείο \src{xclingo} για τη δημιουργία ενός γραφήματος
-το οποίο αποτυπώνει τους συνειρμούς που ακολούθησε το σύστημά 
-μας για τις προβλέψεις του.
+In the program we developed, we also added the capability to explain the diagnosis, where we utilize the \src{xclingo} tool to create a graph that depicts the associations followed by our system for its predictions.
 
-\begin{lstlisting}[caption={Μέρος της επεξήγησης για τη διάγνωση της ασθένειας chickenpox. Μέσα στο διάγραμμα δέντρο φαίνονται οι συσχετίσεις μεταξύ των διαφορετικών συμπτωμάτων 
-και το πώς το σύστημά μας καταλήγει στα συμπεράσματά του.}, label=lst:terms]{Name}
+\begin{lstlisting}[caption={Part of the explanation for the diagnosis of the
+disease chickenpox. The decision tree diagram shows the associations between
+different symptoms and how our system arrives at its conclusions.},
+label=lst:terms]{Name}
   *
@@
 
-\section{Συμπεράσματα}
+\section{Conclusions}
 
-Στην εργασία αυτή έγινε η παρουσίαση μίας μεθοδολογίας για την αυτόματη πρόβλεψη ασθενειών 
-η οποία περιλαμβάνει την κωδικοποίηση ιατρικής βιβλιογραφίας και τη μετατροπή της σε κώδικα 
-ASP μέσω ενός μεγάλου γλωσσικού μοντέλου.
-Έγινε παρουσίαση του τρόπου λειτουργίας των μεγάλων γλωσσικών μοντέλων και του ASP, 
-ενώ έγινε συγκεκριμένη αναφορά στο ρόλο που παίζει η κάθε τεχνολογία στο τελικό σύστημα.
-Ακόμα, δείξαμε την αρχιτεκτονική μας και το πώς αυτές οι τεχνολογίες θα συνδυαστούν ενώ
-δώσαμε συγκεκριμένες υλοποιητικές λεπτομέρειες τις οποίες αναπτύξαμε τόσο
-για την δημιουργία ποιοτικών prompts για το μεγάλο γλωσσικό μοντέλο, όσο και
-στην δομή των κανόνων στην τελική ASP βάση γνώσης.
-Τελικά, εκτελέσαμε προβλέψεις πάνω σε ορισμένες ασθένειες για τις οποίες δημιουργήσαμε την
-βάση γνώσης με τη χρήση της μεθοδολογίας μας και παρουσιάσαμε τα αποτελέσματα.
-Η μέθοδός μας φαίνεται να έχει υποσχόμενα αποτελέσματα και προσφέρει αξιόπιστες προβλέψεις υψηλής ακρίβειας
-μαζί με τις επεξηγήσεις τους.
+In this paper, a methodology for the automatic prediction of diseases was
+presented, which includes the encoding of medical literature and its
+transformation into ASP code through a large language model. The operation of
+large language models and ASP was presented, with specific reference to the
+role each technology plays in the final system. We also showed our architecture
+and how these technologies will be combined, while providing specific
+implementation details that we developed for both the creation of quality
+prompts for the large language model and the structure of the rules in the
+final ASP knowledge base. Finally, we executed predictions on certain diseases
+for which we created the knowledge base using our methodology and presented the
+results. Our method appears to have promising outcomes and offers reliable
+high-accuracy predictions along with their explanations.
 
-Μελλοντική δουλειά είναι η επέκταση της μεθόδου για περισσότερες ασθένειες.
-Ταυτόχρονα θα ήταν χρήσιμη η δημιουργία μίας γραφικής διεπαφής μέσω 
-της οποίας κάποιος να μπορεί να εισάγει στοιχεία κάποιου ασθενή και να παίρνει 
-πιθανές διαγνώσεις.
-
+Future work includes extending the method to cover more diseases.
+Simultaneously, it would be useful to create a graphical interface through
+which someone could input patient data and receive possible diagnoses.
 \newpage

### 2025-05-28 paper/draft.tex
Make doc compilable
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

@@
 \baselineskip=16pt
-
 \usepackage{indentfirst,csquotes}
@@
 
-\usepackage{amssymb,amsthm,amsmath}
-\usepackage{xcolor,paralist,hyperref,titlesec,fancyhdr,etoolbox}
 \usepackage{datetime}
@@
 \usepackage{enumerate}
-\usepackage{float}
 \usepackage{graphicx}
 \usepackage{caption}
-\usepackage[acronym]{glossaries}
-\usepackage[nottoc]{tocbibind}
 \usepackage{siunitx}
 \usepackage{subcaption}
-\usepackage{geometry}
-\usepackage{fancyhdr}
-\usepackage{cite}
-\usepackage{subfiles}
 \usepackage{amsmath}
@@
 \usepackage{ulem}
-\usepackage{polyglossia}
-\usepackage{fontspec}
 \usepackage{hyperref}
@@
 \usepackage{tikz}
-\usepackage{tikzpeople}
 \usepackage{array}
@@
 \usepackage{cleveref}
-\usepackage{epigraph}
 \usepackage{microtype}
 \usepackage{dirtytalk}
+
+\usetikzlibrary{trees}
+\usetikzlibrary{fit}
+\usetikzlibrary{shapes}
+\usetikzlibrary{arrows.meta, positioning, shadows}
+\usetikzlibrary{chains,shapes.multipart}
+\usetikzlibrary{external}
+\usetikzlibrary{matrix}
+
+\tikzstyle{rect} = [rectangle,fill=white, text centered]
+\tikzstyle{database} = [draw, cylinder, shape border rotate = 90, aspect = 0.2]
+\tikzstyle{thick-arrow} = [->, thick]
+\tikzstyle{arrow} = [thick,->,>=stealth]
+\tikzstyle{arrow-small} = [->,>=stealth]
+\tikzstyle{simple-rect} = [rectangle, text centered, draw = black, inner sep=4mm]
+
+\tikzset{
+    doc/.style={draw, minimum height=4em, minimum width=3em, 
+                fill=white, 
+                double copy shadow={shadow xshift=4pt, 
+                             shadow yshift=4pt, fill=white, draw}}
+}
+
+\tikzset{
+    dcs/.style = {double copy shadow},
+}
 
@@
 
-\titleformat{\section}[display]{\normalfont\huge\bfseries\centering}{\centering\chaptertitlename\thechapter}{10pt}{\Large}
-\titlespacing*{\section}{0pt}{0ex}{0ex}
+\hypersetup{
+    colorlinks=true,
+    filecolor=magenta,      
+    urlcolor=blue,
+}
 
-\hypersetup{ colorlinks=true, linkcolor=black, filecolor=black, urlcolor=black }
-\def\proof{\noindent {\it Proof. $\, $}}
-\def\endproof{\hfill $\Box$ \vskip 5 pt }
+\hypersetup{linkcolor=black}
+
+\newcommand{\src}[1]{\texttt{#1}}
+
+\lstset{basicstyle=\footnotesize\ttfamily,breaklines=true, captionpos=b}
 
@@
 
-\renewcommand{\lstlistingname}{Καταχώριση}
-
 \title{Automatic Disease Diagnosis Using Large Language Models and Answer Set Programming}
-\author{Ioanna Gemou\thanks{Department of Electrical and Computer Engineering, University of Patras} \\ \src{ioannagemou@upnet.gr} \and Evangelos Lamprou\footnotemark[1] \\ \src{e.lamprou@upnet.gr}}
+\author{
+	Ioanna Gemou
+	\and
+	Evangelos Lamprou
+}
 
@@
 \maketitle
-\section{Abstract}
+\begin{abstract}
 The development of automatic disease prediction systems is critically important for improving human health and supporting physicians in accurate diagnosis and treatment.
 Accurate diagnosis is essential for timely intervention and effective therapy, and it can also help prevent further complications and burdensome medical procedures.
-
 Symbolic artificial intelligence has previously been applied in the medical domain \cite{Alviano_2020}.
@@
 The proposed method, which combines Large Language Models (LLMs) with Answer Set Programming (ASP), offers a powerful approach by providing a reliable and efficient means of generating a knowledge base that can then be used for disease prediction.
-
 The novelty of our method lies in the combination of these two paradigms.
@@
 This combination enables the development of a robust disease prediction system that integrates the \textbf{accuracy and intelligence of language models} with the \textbf{explainability and efficiency of Answer Set Programming}.
-
 Our method demonstrates promising results on the patient dataset to which it was applied.
+\end{abstract}
 
@@
 \bibliographystyle{unsrt}
-\bibliography{bibliography}
+\bibliography{bib}
 

### 2025-06-01 paper/draft.tex
Full pass over the paper

@@
 
-\section{Literature Review}
+\section{Related work}
 
 \subsection{Large Language Models}
+Large language models (LLMs) \cite{zhao2023survey} are advanced computational systems 
+designed to understand and generate human language. 
+Developed using cutting-edge machine learning techniques, 
+these models can produce text that closely mirrors human linguistic patterns.
+Modern architectures such as \textit{GPT-4} \cite{openai2023gpt4}, 
+are trained on massive corpora spanning diverse domains, 
+enabling them to acquire broad, cross-disciplinary knowledge. 
+LLMs use deep learning algorithms \cite{Sarker2021} 
+and specifically transformer networks \cite{Dosovitskiy2020} 
+to analyze and understand the relationships between words, phrases, and sentences. 
+By capturing patterns and structures in the training data, 
+they can generate coherent and relevant responses based on given prompts.
 
-Large language models \cite{zhao2023survey} are advanced computational models designed to understand and generate human language.
-These models, developed using sophisticated machine learning techniques, have the ability to process and generate text in a way that closely resembles human language patterns.
+The term “large” refers to the billions of parameters 
+that allow these models to generate context-aware responses. 
+The basic premise of a language model is its ability to predict 
+the next word (referred to as a token) based on the text it has observed so far, 
+according to the context.
 
-Large language models, such as the \textit{GPT-4} architecture \cite{openai2023gpt4}, are trained on vast amounts of text data, allowing them to acquire knowledge across various fields and topics.
-These models use deep learning algorithms \cite{Sarker2021} and specifically transformer networks \cite{Dosovitskiy2020} to analyze and understand the relationships between words, phrases, and sentences.
-By capturing patterns and structures in the training data, these models can generate coherent and relevant responses based on given prompts.
+% The following examples illustrate how LLMs have supported various applications \cite{Paranjape2023}.
 
-The models are referred to as "large" because they have billions of parameters that shape their responses.
-The basic premise of a language model is its ability to predict the next word (referred to as a token) based on the text it has observed so far, according to the context.
+LLMs have demonstrated strong performance across a variety of natural language tasks, 
+including question answering \cite{Brown2020}, summarization \cite{Zhang2020}, 
+dialogue generation \cite{Thoppilan2022lamda}, and code synthesis \cite{Chen2021}. 
+In specialized domains such as biomedicine, domain-adapted LLMs have been employed 
+for clinical decision support \cite{Singhal2023} and 
+biomedical knowledge retrieval \cite{Luo2022}.
+Among their capabilities, LLMs excel at transforming text from one representation to another, 
+making them powerful tools for a wide range of natural language processing tasks. 
+This characteristic is particurarly relevant to the present work, 
+which focuses on the effective encoding of medical literature.
 
-Large language models possess the remarkable ability to transform text from one form to another, making them powerful tools for various language processing tasks.
-This characteristic underpins the idea of this work, as a significant part we deal with is the appropriate encoding of medical literature.
-Below we list some examples of the use of large language models \cite{Paranjape2023}:
+\subsection{Challenges in applying LLMs}
 
-\subsection{Difficulties in Implementing LLMs}
+LLMs are not without errors \cite{Raj2023, Ruis2023}. 
+They can occasionally produce incorrect or biased responses 
+that reflect the biases inherent in the training data. 
+There have been efforts to mitigate these issues 
+through data preprocessing techniques and fine-tuning \cite{Dodge2021}.
 
-Large language models (LLMs) are not without errors \cite{Raj2023}, \cite{Ruis2023}.
-They can occasionally produce incorrect or biased responses that reflect the biases inherent in the training data.
-Efforts are being made to mitigate these issues through data preprocessing techniques and fine-tuning.
-
-Overall, large language models have demonstrated tremendous potential across a wide range of applications, including natural language understanding, content creation, and human-computer interaction.
-Ongoing research and development in this field promise further enhancement of these models and their applications across various domains, contributing to the advancement of language processing and communication technologies.
+Ongoing research seeks to improve the factual accuracy, safety, and interpretability 
+of these models, enabling their deployment in high-stakes environments \cite{Ganguli2022}.
 
@@
 
-Answer Set Programming (ASP) \cite{Eiter2009} is a declarative problem-solving approach based on Logic Programming and Non-Monotonic Reasoning.
-The semantics of stable models and the basic language of ASP were initially formalized in the work \cite{Gelfond2000}.
+Answer Set Programming (ASP) \cite{Eiter2009} is a declarative problem-solving paradigm 
+rooted in logic programming and non-monotonic reasoning \cite{Brewka2011}. 
+The formal semantics of stable models and the foundational ASP language 
+were first introduced by Gelfond and Lifschitz \cite{Gelfond2000, gel88}.
 
-Programming with this approach is done in a family of languages sometimes referred to as \textit{AnsProlog} \cite{Gelfond2002}.
+Programming with this approach is done in a family of languages 
+sometimes referred to as \textit{AnsProlog} \cite{Gelfond2002}.
 
-The idea behind ASP is to model a problem as a set of rules and data, and then use a solver to find a solution to the problem.
-Solutions are represented by stable models or answer sets.
-The rules, data, and constraints that describe the problem are the elements of the program.
-Subsequently, the program is given to a solver that will find one or more solutions.
+The central idea of Answer Set Programming is to represent a problem 
+declaratively using logical rules, facts, and constraints. 
+These elements collectively form a program that describes the conditions any solution must satisfy. 
+Instead of writing an algorithm to solve the problem, 
+the programmer encodes the specification of valid solutions. 
+An ASP solver then processes this program to compute one or more stable models (also known as answer sets),
+each of which corresponds to a solution that satisfies all the given rules and constraints.
 
-In traditional programming, transitioning from a problem to a solution requires the programmer's thorough understanding of the given problem, followed by the creation of a program that will produce the correct output when provided with an instance of the problem, which will then be interpreted as the solution.
-
-In ASP, the transition from problem statement to a set of solutions involves the following steps \cite{Gebser2013}~(\cref{fig:asp-solving}):
+In traditional programming, transitioning from a problem to a solution 
+requires the programmer's thorough understanding of the given problem, 
+followed by the creation of a program that will produce the correct output 
+when provided with an instance of the problem, which will then be interpreted as the solution. 
+In Answer Set Programming, the process of deriving solutions from a problem specification 
+typically follows a sequence of well-defined steps \cite{Gebser2013}, 
+as illustrated in \cref{fig:asp-solving}.
 
@@
 Several solvers for ASP have been developed, such as DLV \cite{Xia2020}. 
-In this work, we will apply the ASP system clingo \cite{Gebser2014}, which combines the grounder gringo with the clasp solver \cite{Holldobler2014} into a single application, also providing a powerful Python API for integrating the solver into other applications.
+This work uses the ASP system clingo \cite{Gebser2014}, 
+which combines the grounder gringo with the clasp solver \cite{Holldobler2014} 
+into a single application, also providing a powerful Python API for integrating the solver into other applications.
 
-\subsection{Use of ASP in Disease Diagnosis}
-Answer Set Programming (ASP) offers a powerful method for representing medical knowledge in a declarative and logical manner.
-Below are some common ASP methods used in the representation of medical knowledge \cite{Vinarti2019}:
+\subsection{ASP for disease diagnosis}
+ASP offers a powerful method for representing medical knowledge in a declarative and logical manner. 
+Some common ASP methods used in the representation of medical knowledge \cite{Vinarti2019}:
 
@@
     \item \textbf{Knowledge Base Construction}: ASP provides a framework for constructing a knowledge base that captures medical knowledge.
-The knowledge base consists of facts, rules, and constraints that define the information and limitations related to medical diagnosis and treatment.
-It may include information about disease-symptom relationships, diagnostic criteria, treatment guidelines, and patient data.
+The knowledge base consists of facts, rules, and constraints and may capture a wide range of medical information, including symptom-disease associations, diagnostic criteria, treatment protocols, and patient-specific data.
 \end{itemize}
 
-In the work \cite{Erdem2011}, new methods were introduced for effectively handling complex queries within the context of biomedical ontologies and databases, with an emphasis on extracting relevant information from these knowledge resources.
-These methods were applied to address demanding queries in the field of drug discovery, using biological databases in conjunction with concise explanations for complex queries related to drug discovery processes.
+Earlier methods address complex queries over biomedical ontologies 
+and databases by extracting relevant information from structured 
+knowledge sources \cite{Erdem2011}. Applied in drug discovery, 
+these techniques integrate biological databases with concise, 
+interpretable explanations for complex queries related to drug development.
 
@@
 \end{center}
-\caption{The diagnostic process using the proposed method. Medical literature,
-	through the use of a large language model, is transformed into an ASP
-	program. Subsequently, this program is executed by an ASP solver, which
-	produces the final diagnosis based on the data of each patient.}
+\caption{The diagnostic process begins by transforming medical literature 
+    into an ASP program using a LLM. 
+    An ASP solver then executes the program and produces a final diagnosis 
+    based on each patient’s data.}
     \label{fig:architecture}
@@
 
-Our goal is to extract a knowledge base from unstructured text \cite{ré2014feature}. Our architecture (Figure \ref{fig:architecture}) includes the following stages:
+The goal is to extract a structured knowledge base from unstructured medical text \cite{ré2014feature}.
+\Cref{fig:architecture} illustrates the proposed architecture, which follows the key steps below:
 
 \begin{itemize}
-	\item Collecting medical literature related to the diseases/conditions under examination.
-	\item Creating an appropriate prompt for converting medical literature into ASP code.
-	\item Adding the ASP code generated by the large language model to the knowledge base.
-	\item Adding relevant patient data (symptoms, medical examination results) to the knowledge base.
-	\item Producing the final diagnosis (or set of possible diagnoses) as answer set(s), after the solver reasons over the final knowledge base.
+    \item Collect medical literature relevant to the diseases or conditions of interest.
+    \item Design a suitable prompt to convert the medical literature into ASP code.
+    \item Incorporate the ASP code generated by the LLM into the knowledge base.
+    \item Add relevant patient data—such as symptoms and examination results—to the knowledge base.
+    \item Run an ASP solver to reason over the final knowledge base and produce the diagnosis or set of possible diagnoses as answer sets.
 \end{itemize}
@@
 
-Despite their high capabilities in executing instructions, large language
-models still have low reliability \cite{bowman2023things}. Thus, a relatively
-new programming methodology known as prompt engineering has emerged
-\cite{liu2021pretrain, promptingguidePromptEngineering}. In order to convert
-medical literature into code suitable for execution in an ASP solver, we had to
-construct quality prompts so that the generated code is in the desired form
-(Figure \ref{fig:bad-prompt}, Figure \ref{fig:good-prompt}).
+While large language models demonstrate strong instruction-following capabilities, 
+their reliability remains limited in practice \cite{bowman2023things}. 
+To address this, a relatively recent methodology known as prompt engineering 
+has emerged \cite{liu2021pretrain, promptingguidePromptEngineering}. 
+In this work, prompts are carefully crafted to guide the model in 
+converting medical literature into ASP code suitable for execution 
+by a solver (\cref{fig:bad-prompt}, \cref{fig:good-prompt}).
+\footnote{The example of a poorly constructed prompt is included for educational purposes, to highlight common pitfalls in prompt design.}
 
+% Ioanna: I commented out the bad prompt
 \begin{figure}[!h]
@@
 
-The above paragraph refers some common symptoms of pneumonia.
-Give me a Clingo script so that based on symptoms I can diagnose pneumonia.
+The paragraph above lists common symptoms of pneumonia.
+Write a Clingo script that diagnoses pneumonia based on these symptoms.
 \end{prompt}
@@
 
-The above paragraph refers some common symptoms of pneumonia.
-Give me a Clingo script so that based on symptoms I can diagnose pneumonia.
+The paragraph above lists common symptoms of pneumonia.
+Write a Clingo script that diagnoses pneumonia based on these symptoms.
 
-In the clingo script that you produced before, make some changes in diagnosis rule like:
+In the diagnosis rule, use a structure like:
+
 \src{diagnosis(pneumonia) :- has(symptom(x)), has(symptom(y)) ...}
 
-Include some alternative diagnosis for pneumonia.
-Create rules based on one symptom being linked to another.
+Include alternative diagnoses that share overlapping symptoms with pneumonia.
+Add rules that link one symptom to another (e.g., symptom propagation or dependency).
+
 \end{prompt}
@@
 \end{commandline}
-\caption{A good prompt that yields satisfactory results.}
+\caption{An effective prompt that produces accurate and well-structured output.}
 \label{fig:good-prompt}
@@
 
-The knowledge base initially consists of a set of symptoms and medical indicators in the form:
-
+The knowledge base is initialized with a set of symptoms and medical indicators represented as:
 \begin{align}
@@
 
-We represent a patient who has a certain symptom or tested positive for a certain test as:
-
+A patient exhibiting a specific symptom or testing positive for a particular indicator is represented as:
 \begin{equation}
@@
 
-The logical propositions based on which our system will draw conclusions are in the form:
-
+Logical rules used for inferring diagnoses follow this general form:
 \begin{align}
@@
 
-We also take into account the correlation between certain symptoms.
-For example, there has been a noted connection between skin diseases and headaches in certain patients\cite{migraine-hives}.
-Sometimes, symptoms may lead to the development of other symptoms.
-Such relationships are encoded in our knowledge base through a rule of the form:
-
+Correlations between certain symptoms are also encoded in the knowledge base. 
+For example, clinical observations have reported links between skin conditions and headaches \cite{migraine-hives}. 
+In some cases, the presence of one symptom may lead to the appearance of another. 
+Such relationships are captured using rules of the form:
 \begin{align}
@@
 
-The corresponding data will be as follows:
-
+Example entries for such relationships include:
 \begin{align}
@@
 
-However, it is expected that a patient will not exhibit \textit{all} recorded symptoms.
-We want to provide a controllable way for our system to make alternative predictions in cases where the medical staff does not have all the information about the patient's condition.
-Thus, we add the following choice rule, which adds a set of possible symptoms for the patient to the knowledge base.
-
+Since patients may not exhibit \textit{all} known symptoms, 
+the system must support reasoning under partial information. 
+To allow flexible reasoning, a choice rule introduces possible symptoms 
+into the knowledge base when patient data is incomplete:
 \begin{equation}
@@
 
-Additionally, we require the solver to find at least one diagnosis.
-
+To ensure that the solver identifies at least one diagnosis, 
+the following integrity constraint is imposed:
 \begin{equation}
@@
 
-Finally, we minimize the number of "false" diagnoses so that the solver arrives at a diagnosis with as few assumptions as possible.
-
+Finally, the solver minimizes assumptions by penalizing added symptoms, 
+favoring diagnoses based on known patient information:
 \begin{equation}
@@
 
-This way, our system guarantees that it will produce at least one prediction, which will be based as much as possible on valid information available about the patient.
+This formulation guarantees that at least one diagnosis is produced, 
+relying as much as possible on available clinical data rather than hypothetical assumptions.
 
@@
 
-Symbolic artificial intelligence has a major advantage over subsymbolic approaches due to its explainability.
-The relationships between the facts and conclusions that a system, which executes a symbolic mechanism, reaches are clearly defined.
-However, ASP solvers like \textit{Clingo} do not provide\footnote{At the time of writing this paper.} intuitive traces of their reasoning processes.
-There has been research on linking logical propositions with their justifications \cite{cabalar2014causal}.
-Consider the following logic program; the connection of the logical propositions with their causes is presented in the form of a graph in Figure \ref{fig:causal-g}.
-This structure allows us to explain the conclusions of a symbolic logic solver.
-
+Symbolic artificial intelligence offers a significant advantage 
+over subsymbolic methods in terms of explainability. 
+In symbolic systems, the relationships between facts and conclusions are explicitly defined, 
+enabling transparent reasoning. However, ASP solvers such as \textit{Clingo} do not, 
+at the time of writing,\footnote{At the time of writing this paper.} 
+provide intuitive traces of their reasoning processes. 
+Prior work has explored methods for associating logical propositions 
+with their justifications \cite{cabalar2014causal}. For example, 
+the logic program shown below is accompanied by a causal graph 
+in \cref{fig:causal-g}, which illustrates the connections 
+between rules and their inferred outcomes. 
+This structure supports the interpretability of conclusions derived by a symbolic logic solver.
 \begin{equation}
@@
     \includegraphics{assets/causal_g.png}
-    \caption{Causal graph of logic program (eq. \ref{eq:example-cg}).}
+    \caption{Causal graph of logic program (\cref{eq:example-cg}).}
     \label{fig:causal-g}
@@
 
-The tool \textit{xclingo}\footnote{\url{https://github.com/bramucas/xclingo2}} \cite{Cabalar_2020} allows the presentation of justifications regarding the results given by the solver \textit{Clingo}.
+The tool \textit{xclingo}\footnote{\url{https://github.com/bramucas/xclingo2}} \cite{Cabalar_2020} 
+allows the presentation of justifications regarding the results given by the solver \textit{Clingo}. 
 This is achieved by adding a \src{trace} corresponding to the rules of our logic program. 
-Providing a clear explanation of how the system reached specific conclusions can help identify potential errors in the creation of the knowledge base and provide the system user with a rationale for the final outcome.
+roviding a clear explanation of how the system reached specific conclusions can help identify potential 
+errors in the creation of the knowledge base and provide the system user with a rationale for the final outcome. 
 This is particularly important when the application is used in medical contexts.
@@
 
-To evaluate our system, we used publicly available patient data\footnote{\url{https://www.kaggle.com/datasets/itachi9604/disease-symptom-description-dataset?select=dataset.csv}}.
-The symptoms of each patient were fed into the system as input.
-Our knowledge base consists of the combination of all rules derived from encoding the medical literature.
-A successful prediction is defined when the actual disease of the patient matches the final prediction of the ASP solver.
-
+Evaluation was conducted using publicly available patient data.\footnote{\url{https://www.kaggle.com/datasets/itachi9604/disease-symptom-description-dataset?select=dataset.csv}} 
+Each patient's symptoms were provided as input to the system. 
+The knowledge base combines all rules derived from the encoded medical literature. 
+A prediction is considered successful when the disease identified by the ASP solver matches the patient’s actual diagnosis.
 \begin{table}[h]
@@
     \end{tabular}
-    \caption{Accuracy of the method for some indicative diseases.}
+    \caption{Accuracy of the method on selected diseases.}
 \end{table}
 
-The accuracy of our model can reach very high rates but requires the input of a large number of rules into the knowledge base.
-We define the size of the knowledge base as the number of logical terms within the ASP program corresponding to each of the diseases.
+The model achieves high accuracy when a sufficiently 
+large set of rulesis included in the knowledge base. 
+The size of the knowledge base is defined as the number 
+of logical terms in the ASP program associated with each disease.
 
-In the program we developed, we also added the capability to explain the diagnosis, where we utilize the \src{xclingo} tool to create a graph that depicts the associations followed by our system for its predictions.
+The system also incorporates an explanation mechanism using the \src{xclingo} tool,
+which generates a graph visualizing the logical associations used to derive the diagnosis.
 
-\begin{lstlisting}[caption={Part of the explanation for the diagnosis of the
-disease chickenpox. The decision tree diagram shows the associations between
-different symptoms and how our system arrives at its conclusions.},
+\begin{lstlisting}[caption={A part of the explanation for the chickenpox diagnosis. 
+    The decision tree illustrates the symptom associations and outlines 
+    the reasoning process used by the system to reach its conclusion.},
 label=lst:terms]{Name}
@@
 
-In this paper, a methodology for the automatic prediction of diseases was
-presented, which includes the encoding of medical literature and its
-transformation into ASP code through a large language model. The operation of
-large language models and ASP was presented, with specific reference to the
-role each technology plays in the final system. We also showed our architecture
-and how these technologies will be combined, while providing specific
-implementation details that we developed for both the creation of quality
-prompts for the large language model and the structure of the rules in the
-final ASP knowledge base. Finally, we executed predictions on certain diseases
-for which we created the knowledge base using our methodology and presented the
-results. Our method appears to have promising outcomes and offers reliable
-high-accuracy predictions along with their explanations.
+This work presents a method for automatic disease prediction 
+that encodes medical literature and transforms it into ASP code using an LLM. 
+It describes the roles of both LLMs and ASP, emphasizing how each contributes to the overall system. 
+The system architecture is outlined, along with implementation details for constructing effective prompts 
+and structuring rules within the ASP knowledge base.
+The method includes predictions for a set of diseases based on the constructed knowledge bases. 
+Results show that this approach achieves high-accuracy predictions and provides interpretable explanations.
 
-Future work includes extending the method to cover more diseases.
-Simultaneously, it would be useful to create a graphical interface through
-which someone could input patient data and receive possible diagnoses.
+Future work includes integrating retrieval-augmented generation (RAG) to improve the factual accuracy 
+and reliability of the ASP code, and fine-tuning the LLM on clinical reasoning tasks to enhance rule quality. 
+Expanding the approach to cover additional diseases is also a key direction.
 \newpage

### 2025-06-07 paper/draft.tex
Add intro, update abstract and add related works

@@
 
+\newcommand{\sys}{\textsc{Sys}\xspace}
+
 \title{Automatic Disease Diagnosis Using Large Language Models and Answer Set Programming}
@@
 \begin{abstract}
-The development of automatic disease prediction systems is critically important for improving human health and supporting physicians in accurate diagnosis and treatment.
-Accurate diagnosis is essential for timely intervention and effective therapy, and it can also help prevent further complications and burdensome medical procedures.
-Symbolic artificial intelligence has previously been applied in the medical domain \cite{Alviano_2020}.
-However, its adoption has been limited due to the complex knowledge bases required for such systems to operate.
-
-The proposed method, which combines Large Language Models (LLMs) with Answer Set Programming (ASP), offers a powerful approach by providing a reliable and efficient means of generating a knowledge base that can then be used for disease prediction.
-The novelty of our method lies in the combination of these two paradigms.
-LLMs are capable of encoding medical literature into ASP code with remarkable accuracy and deep comprehension, while ASP contributes structured reasoning and explainability.
-This combination enables the development of a robust disease prediction system that integrates the \textbf{accuracy and intelligence of language models} with the \textbf{explainability and efficiency of Answer Set Programming}.
-Our method demonstrates promising results on the patient dataset to which it was applied.
+Accurate disease prediction is vital for timely intervention, 
+effective treatment, and reducing medical complications. 
+While symbolic AI has been applied in healthcare, 
+its adoption remains limited due to the complexity 
+of required knowledge bases.
+This work introduces \sys, a framework that combines Large Language Models (LLMs) 
+with Answer Set Programming (ASP) to overcome this barrier. 
+LLMs extract medical knowledge from text and translate it into ASP code, 
+while ASP enables structured, explainable reasoning. 
+This integration yields a robust, interpretable prediction system 
+that leverages the strengths of both paradigms. 
+This approach shows strong performance on real patient data.
+% \sys accurately predicts diseases based on symptoms derived from medical literature.
 \end{abstract}
 
-\section{Related work}
+% TODO: Add text between sections and paragraphs
+\section{Introduction}
 
-\subsection{Large Language Models}
+The development of automatic disease prediction systems 
+is crucial for enhancing healthcare and supporting medical 
+professionals in timely and accurate diagnosis. 
+Symbolic artificial intelligence (AI), particularly Answer Set Programming (ASP), 
+has been explored in this context due to its ability 
+to support structured and explainable reasoning. 
+However, the construction of expressive knowledge bases 
+remains a significant barrier to practical adoption.
+
+This paper introduces \sys, a framework 
+that combines Large Language Models (LLMs) with the 
+formal reasoning capabilities of ASP.
+\sys extracts and encodes medical knowledge from text into ASP code,
+enabling interpretable and accurate diagnosis 
+of diseases based on patient symptoms.
+
+\section{Background}
+
+\paragraph{\textbf{Large Language Models}}
 Large language models (LLMs) \cite{zhao2023survey} are advanced computational systems 
@@
 they can generate coherent and relevant responses based on given prompts.
-
 The term “large” refers to the billions of parameters 
@@
 
-\subsection{Challenges in applying LLMs}
+\paragraph{\textbf{Challenges in applying LLMs}}
 
@@
 through data preprocessing techniques and fine-tuning \cite{Dodge2021}.
-
 Ongoing research seeks to improve the factual accuracy, safety, and interpretability 
@@
 
-\subsection{Answer Set Programming}
+\paragraph{\textbf{Answer Set Programming}}
 
@@
 were first introduced by Gelfond and Lifschitz \cite{Gelfond2000, gel88}.
-
 Programming with this approach is done in a family of languages 
@@
 
-\subsection{ASP for disease diagnosis}
-ASP offers a powerful method for representing medical knowledge in a declarative and logical manner. 
-Some common ASP methods used in the representation of medical knowledge \cite{Vinarti2019}:
+\section{Related work}
+% we need to add something about LLM + ASP in disease diagnosis
+\paragraph{\textbf{ASP for disease diagnosis}}
+Answer Set Programming (ASP) has been used to represent 
+and reason over medical knowledge in a declarative, logic-based manner. 
+Early work applied ASP to biomedical ontologies and structured databases 
+to answer complex diagnostic and pharmacological queries with interpretable 
+logical justifications \cite{Erdem2011}. In the clinical domain, 
+ASP is especially effective for encoding medical rules and patient-specific data, 
+enabling precise and explainable decision-making.
+Common strategies in ASP-based medical reasoning include:
 
 \begin{itemize}
-    \item \textbf{Rule-Based Representation}: ASP allows for the representation of medical knowledge as rules in the form of logical statements.
-These rules define relationships between medical entities such as symptoms, diseases, treatments, and patient data.
-For example, a rule could state that if a patient exhibits specific symptoms, it indicates the presence or absence of certain diseases.
-    \item \textbf{Knowledge Base Construction}: ASP provides a framework for constructing a knowledge base that captures medical knowledge.
-The knowledge base consists of facts, rules, and constraints and may capture a wide range of medical information, including symptom-disease associations, diagnostic criteria, treatment protocols, and patient-specific data.
+    \item \textbf{Rule-Based Representation}: ASP allows for the representation of medical 
+    knowledge as rules in the form of logical statements. 
+    These rules define relationships between medical entities such as symptoms, 
+    diseases, treatments, and patient data. For example, a rule could state that 
+    if a patient exhibits specific symptoms, it indicates the presence or absence 
+    of certain diseases.
+    \item \textbf{Knowledge Base Construction}: ASP facilitates the construction of 
+    structured knowledge bases composed of facts, rules, and constraints. These can 
+    model diagnostic criteria, symptom-disease associations, and therapeutic protocols 
+    in a consistent and extensible manner.
 \end{itemize}
 
-Earlier methods address complex queries over biomedical ontologies 
-and databases by extracting relevant information from structured 
-knowledge sources \cite{Erdem2011}. Applied in drug discovery, 
-these techniques integrate biological databases with concise, 
-interpretable explanations for complex queries related to drug development.
+\paragraph{\textbf{Integrating LLMs with ASP}}
 
+Recent work explores the synergy between LLMs and ASP 
+for enhanced diagnostic reasoning. 
+These hybrid approaches aim to combine the language understanding of LLMs 
+with the precision and explainability of symbolic reasoning. 
+A YAML-based interface has been proposed to guide LLMs 
+in extracting structured facts from natural language, 
+which are then processed by ASP for logical inference \cite{alviano2024llm2asp}. 
+Other work uses ASP to validate and trace misleading or inconsistent outputs 
+from LLMs in medical contexts \cite{Nguyen2025}. 
+A specialized lightweight LLM has also been trained to generate valid ASP code, 
+improving the automation of rule-based medical reasoning \cite{coppolillo2024llasp}.
+These systems assist in building reliable and clear reasoning pipelines 
+for disease diagnosis by combining language understanding with rule-based validation.
 \section{Methodology}
 
-\subsection{Architecture}
+\paragraph{\textbf{Architecture}}
 
@@
 
-The goal is to extract a structured knowledge base from unstructured medical text \cite{ré2014feature}.
-\Cref{fig:architecture} illustrates the proposed architecture, which follows the key steps below:
+The goal is to extract a structured knowledge base from unstructured medical
+text \cite{ré2014feature}. \Cref{fig:architecture} illustrates the proposed 
+architecture, which proceeds as follows. 
+First, medical literature relevant to the diseases or conditions of interest 
+is collected. Then, a carefully designed prompt is used to convert this literature 
+into ASP code via a large language model. 
+The generated ASP code is incorporated into a growing knowledge base, 
+which is further enriched with patient-specific data such as symptoms 
+and examination results. Finally, an ASP solver reasons over the complete 
+knowledge base to produce a diagnosis--or a set of possible diagnoses--in 
+the form of answer sets.
 
-\begin{itemize}
-    \item Collect medical literature relevant to the diseases or conditions of interest.
-    \item Design a suitable prompt to convert the medical literature into ASP code.
-    \item Incorporate the ASP code generated by the LLM into the knowledge base.
-    \item Add relevant patient data—such as symptoms and examination results—to the knowledge base.
-    \item Run an ASP solver to reason over the final knowledge base and produce the diagnosis or set of possible diagnoses as answer sets.
-\end{itemize}
-
-\subsection{Prompt Construction}
+\paragraph{\textbf{Prompt Construction}}
 
@@
 
-\subsection{Construction of a Logical Program}
+\paragraph{\textbf{Construction of a Logical Program}}
 
@@
 
-\subsection{Explainability of Diagnoses}
+\paragraph{\textbf{Diagnostic Explainability}}
 
@@
 allows the presentation of justifications regarding the results given by the solver \textit{Clingo}. 
-This is achieved by adding a \src{trace} corresponding to the rules of our logic program. 
+This is achieved by adding a \src{trace} corresponding to the logic program's rules. 
 roviding a clear explanation of how the system reached specific conclusions can help identify potential 

### 2025-06-07 paper/draft.tex
Add explanation of xclingo excerpt

@@
 while ASP enables structured, explainable reasoning. 
-This integration yields a robust, interpretable prediction system 
+This integration yields a robust, interpretable prediction framework 
 that leverages the strengths of both paradigms. 
@@
 Since patients may not exhibit \textit{all} known symptoms, 
-the system must support reasoning under partial information. 
+\sys must support reasoning under partial information. 
 To allow flexible reasoning, a choice rule introduces possible symptoms 
@@
 \end{equation}
-
+% https://docs.google.com/drawings/d/1Ing4GFH2kbniiHI5rbXzw-xcAUQAxWAJpgmg4tz1fzc/edit?usp=sharing
 \begin{figure}
     \centering
-    \includegraphics{assets/causal_g.png}
+    \includegraphics{assets/causal_g.pdf}
     \caption{Causal graph of logic program (\cref{eq:example-cg}).}
@@
 
-The tool \textit{xclingo}\footnote{\url{https://github.com/bramucas/xclingo2}} \cite{Cabalar_2020} 
-allows the presentation of justifications regarding the results given by the solver \textit{Clingo}. 
-This is achieved by adding a \src{trace} corresponding to the logic program's rules. 
-roviding a clear explanation of how the system reached specific conclusions can help identify potential 
-errors in the creation of the knowledge base and provide the system user with a rationale for the final outcome. 
-This is particularly important when the application is used in medical contexts.
+The tool \textit{xclingo}\footnote{\url{https://github.com/bramucas/xclingo2}}~\cite{Cabalar_2020} 
+enables the visualization of justifications for results produced by the \textit{Clingo} solver. 
+It does so by incorporating a \src{trace} that corresponds to the rules of the logic program. 
+Providing clear explanations of how the framework reaches specific conclusions helps users identify 
+potential errors in the construction of the knowledge base and offers transparency for the final outcome. 
+This is especially important in medical contexts, where interpretability and trust are critical.
 
@@
 Evaluation was conducted using publicly available patient data.\footnote{\url{https://www.kaggle.com/datasets/itachi9604/disease-symptom-description-dataset?select=dataset.csv}}
-Each patient's symptoms were provided as input to the system. 
-The knowledge base combines all rules derived from the encoded medical literature. 
-A prediction is considered successful when the disease identified by the ASP solver matches the patient’s actual diagnosis.
+Each patient's list of symptoms serves as input.  
+The knowledge base—comprising ASP rules automatically derived  
+from medical literature—infers potential diagnoses.  
+A prediction is considered correct when the disease inferred  
+by the ASP solver matches the actual diagnosis listed in the dataset.
+
 \begin{table}[h]
@@
     \end{tabular}
-    \caption{Accuracy of the method on selected diseases.}
+    \caption{Accuracy of the framework on selected diseases. 
+    The size column indicates the number of logical terms in the disease-specific ASP program.}
 \end{table}
 
-The model achieves high accuracy when a sufficiently 
-large set of rulesis included in the knowledge base. 
-The size of the knowledge base is defined as the number 
-of logical terms in the ASP program associated with each disease.
+High accuracy is achieved when the knowledge base includes a sufficiently 
+rich set of logical rules. The size of this base, measured as the number 
+of ASP terms associated with a disease, correlates with prediction performance.
 
-The system also incorporates an explanation mechanism using the \src{xclingo} tool,
-which generates a graph visualizing the logical associations used to derive the diagnosis.
+In addition to accurate predictions, \sys offers explainability 
+through integration with \src{xclingo}. 
+This tool visualizes the reasoning process behind each diagnosis, 
+revealing the logical path from symptoms to conclusion.
 
-\begin{lstlisting}[caption={A part of the explanation for the chickenpox diagnosis. 
-    The decision tree illustrates the symptom associations and outlines 
-    the reasoning process used by the system to reach its conclusion.},
-label=lst:terms]{Name}
+\begin{lstlisting}[caption={Excerpt from the explanation tree for the chickenpox diagnosis. The structure illustrates the reasoning path based on symptom associations.},
+    label=lst:chickenpox_explanation]{Name}
     *
-  |__diagnosis(chickenpox)
-  |  |__has(symptom(itching));has(symptom(itching))
-  |  |__has(symptom(fatigue));has(symptom(fatigue));has(symptom(fatigue));has(symptom(fatigue));has(symptom(fatigue))
-  |  |__has(symptom(lethargy))
-  |  |  |__has(symptom(fatigue));has(symptom(fatigue));has(symptom(fatigue));has(symptom(fatigue));has(symptom(fatigue))
-  |  |  |__linked_symptom(fatigue,lethargy)
-  |  |__has(symptom(high_fever))
-  |  |  |__has(symptom(mild_fever));has(symptom(mild_fever))
-  |  |  |  |__has(symptom(loss_of_appetite));has(symptom(loss_of_appetite))
-  |  |  |  |__linked_symptom(loss_of_appetite,mild_fever)
-  |  |  |__linked_symptom(mild_fever,high_fever)
-  |  |__has(symptom(loss_of_appetite));has(symptom(loss_of_appetite))
-  |  |__has(symptom(mild_fever));has(symptom(mild_fever))
-  |  |  |__has(symptom(loss_of_appetite));has(symptom(loss_of_appetite))
-  |  |  |__linked_symptom(loss_of_appetite,mild_fever)
-  |  |__has(symptom(swelled_lymph_nodes))
-  |  |  
-  ...
+    |__ diagnosis(chickenpox)
+        |__ has(symptom(itching))
+        |__ has(symptom(fatigue))
+        |__ has(symptom(lethargy))
+            |__ has(symptom(fatigue))
+            |__ linked_symptom(fatigue, lethargy)
+        |__ has(symptom(high_fever))
+            |__ has(symptom(mild_fever))
+                |__ has(symptom(loss_of_appetite))
+                |__ linked_symptom(loss_of_appetite, mild_fever)
+            |__ linked_symptom(mild_fever, high_fever)
+        |__ has(symptom(loss_of_appetite))
+        |__ has(symptom(mild_fever))
+            |__ has(symptom(loss_of_appetite))
+            |__ linked_symptom(loss_of_appetite, mild_fever)
+        |__ has(symptom(swelled_lymph_nodes))
 \end{lstlisting}
 
-\section{Conclusions}
+\Cref{lst:chickenpox_explanation} shows an excerpt from the explanation 
+tree generated for a patient diagnosed with \textit{chickenpox}. 
+The tree visualizes the reasoning steps performed by the ASP solver to reach this conclusion. 
+At the root, \sys infers \texttt{diagnosis(chickenpox)} 
+based on the presence of multiple symptoms such as \texttt{itching}, \texttt{fatigue}, \texttt{lethargy}, 
+and \texttt{high\_fever}. 
+Each symptom is further justified through observed inputs or inferred relationships. 
+For example, \texttt{lethargy} is supported by repeated instances of \texttt{fatigue} 
+and the known association \texttt{linked\_symptom(fatigue, lethargy)}. 
+Similarly, \texttt{high\_fever} is explained via the intermediate symptom \texttt{mild\_fever}, 
+which itself is supported by \texttt{loss\_of\_appetite}. 
+These explanations demonstrate how \sys combines direct observations 
+and transitive symptom relationships encoded in the knowledge base to 
+construct a transparent and interpretable diagnostic path.
+
+\section{Conclusion}
 
@@
 that encodes medical literature and transforms it into ASP code using an LLM. 
-It describes the roles of both LLMs and ASP, emphasizing how each contributes to the overall system. 
-The system architecture is outlined, along with implementation details for constructing effective prompts 
+It describes the roles of both LLMs and ASP, emphasizing how each contributes to the overall framework. 
+The \sys architecture is outlined, along with implementation details for constructing effective prompts 
 and structuring rules within the ASP knowledge base.

### 2025-06-07 paper/draft.tex
Update listing's caption

@@
 
-\begin{lstlisting}[caption={Excerpt from the explanation tree for the chickenpox diagnosis. The structure illustrates the reasoning path based on symptom associations.},
+\begin{lstlisting}[caption={Excerpt from the explanation tree for the chickenpox diagnosis.
+The structure illustrates the reasoning path based on symptom associations (slightly modified for clarity).},
     label=lst:chickenpox_explanation]{Name}

### 2025-06-08 paper/draft.tex
Improve overview figure
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

@@
 \begin{center}
-\begin{tikzpicture}[node distance = 5cm, minimum width=2cm, minimum height=1cm]
+\begin{tikzpicture}[node distance = 3cm, minimum width=2cm, minimum height=1cm]
     \node [doc, label=above:{}] (med-bib) {Medical Bibliography};
-    \node [rect, draw, right=3cm of med-bib, fill=blue!50] (llm) {Large Language Model};
+    \node [rect, draw, rounded corners=2pt, right=2cm of med-bib, fill=blue!30] (llm) {Large Language Model};
     \node [rect, draw, above= 1cm of med-bib, fill=orange!50] (prompt) {Prompt};
-    \node [database, below left= 3cm of med-bib, fill=green!30] (knowledge-base) {Knowledge Base};
+    \node [database, below left= 2.5cm of med-bib, fill=green!30] (knowledge-base) {Knowledge Base};
     \node [rect, draw, right of=knowledge-base] (solver) {ASP Solver};
-    \node [rect, draw, below=1cm of solver] (patient) {\textit{Patient Data}};
+    \node [rect, draw, below=.5cm of solver] (patient) {\textit{Patient Data}};
     \node [rect, right of=solver] (diagnosis) {\textbf{Diagnosis}};
@@
     \draw[thick-arrow] (prompt) edge [in=90, out=0, looseness=1] node [above, sloped] {\textit{\say{please convert the following into ASP code...}}} (llm);
-    \draw[thick-arrow] (llm) edge [in=90, out=270, looseness=1] node [below] {ASP Program} (knowledge-base);
+    \draw[thick-arrow] (llm) edge [in=90, out=270, looseness=1] node [rect, draw] {ASP Program} (knowledge-base);
     \draw[thick-arrow] (knowledge-base) edge [] node [above] {} (solver);

### 2025-06-08 paper/draft.tex
Fix caption
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>
Set name
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>
Improve caption
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

@@
 
-\newcommand{\sys}{\textsc{Sys}\xspace}
+\newcommand{\sys}{\textsc{McCoy}\xspace}
 
@@
 \end{center}
-\caption{The diagnostic process begins by transforming medical literature 
+	\caption{\textbf{\sys overview.} The diagnostic process begins by transforming medical literature 
     into an ASP program using a LLM. 
@@
 
-\caption{A simple prompt that does not yield satisfactory results.}
+\caption{A prompt that does not yield satisfactory results. The LLM output does not decompose the disease diagnosis into discrete rules.}
 \label{fig:bad-prompt}
@@
 \end{commandline}
-\caption{An effective prompt that produces accurate and well-structured output.}
+\caption{An effective prompt that produces accurate and well-structured output.
+	The LLM output is structured into multiple rules that aid explainability and can result in partial diagnoses.}
 \label{fig:good-prompt}

### 2025-06-08 paper/draft.tex
Nits

@@
 that leverages the strengths of both paradigms. 
-This approach shows strong performance on real patient data.
+\sys shows strong performance on real patient data.
 % \sys accurately predicts diseases based on symptoms derived from medical literature.
@@
 professionals in timely and accurate diagnosis. 
-Symbolic artificial intelligence (AI), particularly Answer Set Programming (ASP), 
+Symbolic artificial intelligence (AI), particularly ASP, 
 has been explored in this context due to its ability 
@@
 
-The central idea of Answer Set Programming is to represent a problem 
+The central idea of ASP is to represent a problem 
 declaratively using logical rules, facts, and constraints. 
@@
 when provided with an instance of the problem, which will then be interpreted as the solution. 
-In Answer Set Programming, the process of deriving solutions from a problem specification 
+In ASP, the process of deriving solutions from a problem specification 
 typically follows a sequence of well-defined steps \cite{Gebser2013}, 
@@
 \begin{itemize}
-    \item \textbf{Modeling}: The problem is modeled using ASP (Answer Set Programming) language.
+    \item \textbf{Modeling}: The problem is modeled using ASP language.
     \item \textbf{Grounding}: A grounder (e.g., gringo) converts the program into a set of basic rules and facts.
@@
 \paragraph{\textbf{ASP for disease diagnosis}}
-Answer Set Programming (ASP) has been used to represent 
+ASP has been used to represent 
 and reason over medical knowledge in a declarative, logic-based manner. 

### 2025-06-14 paper/draft.tex
Improvements across
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

@@
 \usepackage{dirtytalk}
+\usepackage{booktabs}
 
@@
     filecolor=magenta,      
+		citecolor=black,
     urlcolor=blue,
@@
 
-\title{Automatic Disease Diagnosis Using Large Language Models and Answer Set Programming}
-\author{
-	Ioanna Gemou
-	\and
-	Evangelos Lamprou
+\makeatletter
+\def\@settitle{%
+  \begin{center}%
+    \normalfont\LARGE\bfseries
+    \@title\par
+  \end{center}%
 }
+\makeatother
+
+\title[\sys: Automatic Disease Diagnosis Using LLMs \& ASP]{\sys: Automatic Disease Diagnosis Using Large Language Models and Answer Set Programming}
+
+\author[Ioanna Gemou]{Ioanna Gemou}
+\author[Evangelos Lamprou]{Evangelos Lamprou}
+\thanks{Originally submitted as part of a research competition on 1st June 2023.}
 
@@
 \maketitle
+\vspace{-1.5em} % adjust as needed
+{\small\itshape \centering Originally submitted as part of a university research competition on 1st June 2023.\par}
+\vspace{0.5em} % optional: add space before body
+
 \begin{abstract}
@@
 
-% TODO: Add text between sections and paragraphs
 \section{Introduction}
@@
 This paper introduces \sys, a framework 
-that combines Large Language Models (LLMs) with the 
+that combines LLMs with the 
 formal reasoning capabilities of ASP.
@@
 \paragraph{\textbf{Large Language Models}}
-Large language models (LLMs) \cite{zhao2023survey} are advanced computational systems 
+LLMs \cite{zhao2023survey} are advanced computational systems 
 designed to understand and generate human language. 
@@
     \begin{center}
-\begin{tikzpicture}[node distance=4cm]
+\begin{tikzpicture}[node distance=2.9cm]
         \node [simple-rect] (problem)  at (0,0) {Problem};
@@
         \node [simple-rect, right of=program] (grounder) {Grounder};
-        \node [simple-rect, right of=grounder] (solver) {Solver};
+        \node [simple-rect, right=2cm of grounder] (solver) {Solver};
 
@@
 
-        \node[draw, dotted, fit=(grounder) (solver), inner sep=4mm, label=above:{Solving}] {};
+        \node[draw, dotted, fit=(grounder) (solver), inner sep=2mm, label=above:{Solving}] {};
 \end{tikzpicture}
@@
 to answer complex diagnostic and pharmacological queries with interpretable 
-logical justifications \cite{Erdem2011}. In the clinical domain, 
+logical justifications \cite{Erdem2011}. 
+In the clinical domain, 
 ASP is especially effective for encoding medical rules and patient-specific data, 
@@
     of certain diseases.
+		There have been multiple instances of rule-based medical reasoning systems~\cite{CHEN_2016,marple2017s}
     \item \textbf{Knowledge Base Construction}: ASP facilitates the construction of 
@@
     model diagnostic criteria, symptom-disease associations, and therapeutic protocols 
-    in a consistent and extensible manner.
+		in a consistent and extensible manner~\cite{chen2018automating,CHEN_2016}.
 \end{itemize}
+\sys builds upon these works and identifies a key bottleneck, which is the construction of a
+knowledge base that can cover a wide range of diseases and conditions.
 
@@
 
-Recent work explores the synergy between LLMs and ASP 
+More recent work explores the synergy between LLMs and ASP 
 for enhanced diagnostic reasoning. 
@@
 for disease diagnosis by combining language understanding with rule-based validation.
+These systems were developed after \sys and illustrate the convergent development of similar approaches, highlighting the intuitive appeal and practical value of the underlying techniques.
+
 \section{Methodology}
@@
 
-The tool \textit{xclingo}\footnote{\url{https://github.com/bramucas/xclingo2}}~\cite{Cabalar_2020} 
+The tool \textit{xclingo}~\cite{Cabalar_2020} 
 enables the visualization of justifications for results produced by the \textit{Clingo} solver. 
@@
     \centering
+    \caption{Accuracy of the framework on selected diseases. 
+    The size column indicates the number of logical terms in the disease-specific ASP program.}
     \begin{tabular}{lcc}
-    \hline
-    Diseases & Size of Knowledge Base & Accuracy \\
-    \hline
-    \hline
+    \toprule
+    Diseases & Size & Accuracy \\
+    \midrule
     Chicken Pox    &  66  & 95\% \\
@@
     Common Cold    &  44  & 100\% \\
-    \hline
+    \bottomrule
     \end{tabular}
-    \caption{Accuracy of the framework on selected diseases. 
-    The size column indicates the number of logical terms in the disease-specific ASP program.}
 \end{table}
@@
 The structure illustrates the reasoning path based on symptom associations (slightly modified for clarity).},
-    label=lst:chickenpox_explanation]{Name}
+    label=lst:chickenpox-explanation]{Name}
     *
@@
 
-\Cref{lst:chickenpox_explanation} shows an excerpt from the explanation 
+\Cref{lst:chickenpox-explanation} shows an excerpt from the explanation 
 tree generated for a patient diagnosed with \textit{chickenpox}. 

### 2025-09-12 paper/draft.tex
Update RW

@@
 these models can produce text that closely mirrors human linguistic patterns.
-Modern architectures such as \textit{GPT-4} \cite{openai2023gpt4}, 
+Modern architectures such as \textit{GPT-5} \cite{openai2023gpt4, gpt5}, 
 are trained on massive corpora spanning diverse domains, 
@@
 that allow these models to generate context-aware responses. 
-The basic premise of a language model is its ability to predict 
+The basic premise of an LLM is its ability to predict 
 the next word (referred to as a token) based on the text it has observed so far, 
@@
 This characteristic is particurarly relevant to the present work, 
-which focuses on the effective encoding of medical literature.
+which focuses on the effective encoding of medical literature. \\
 
@@
 Ongoing research seeks to improve the factual accuracy, safety, and interpretability 
-of these models, enabling their deployment in high-stakes environments \cite{Ganguli2022}.
+of these models, enabling their deployment in high-stakes environments \cite{Ganguli2022}. \\
 
@@
 as illustrated in \cref{fig:asp-solving}. 
+First, the problem is modeled using the ASP language. 
+Next, a grounder (e.g., gringo \cite{Gebser2014}) transforms the program into a set of basic rules and facts. 
+Finally, a solver (e.g., clasp \cite{Holldobler2014}) computes the stable models (answer sets), 
+which correspond to the solutions of the original problem.
 
@@
 
-\begin{itemize}
-    \item \textbf{Modeling}: The problem is modeled using ASP language.
-    \item \textbf{Grounding}: A grounder (e.g., gringo) converts the program into a set of basic rules and facts.
-    \item \textbf{Solving}: A solver (e.g., clasp) finds a solution to the problem by computing the set of stable models (answer sets).
-\end{itemize}
-
 Several solvers for ASP have been developed, such as DLV \cite{Xia2020}. 
-This work uses the ASP system clingo \cite{Gebser2014}, 
-which combines the grounder gringo with the clasp solver \cite{Holldobler2014} 
-into a single application, also providing a powerful Python API for integrating the solver into other applications.
+This work uses the ASP system clingo, 
+which integrates the gringo grounder \cite{Gebser2014} with the clasp solver \cite{Holldobler2014} into a single system. 
+Clingo also offers a powerful Python API, enabling the solver's integration into other applications.
 
 \section{Related work}
-% we need to add something about LLM + ASP in disease diagnosis
 \paragraph{\textbf{ASP for disease diagnosis}}
-ASP has been used to represent 
-and reason over medical knowledge in a declarative, logic-based manner. 
-Early work applied ASP to biomedical ontologies and structured databases 
-to answer complex diagnostic and pharmacological queries with interpretable 
-logical justifications \cite{Erdem2011}. 
-In the clinical domain, 
-ASP is especially effective for encoding medical rules and patient-specific data, 
+ASP has been used to represent and reason over medical knowledge in a declarative, logic-based manner. 
+Early work applied ASP to biomedical ontologies and structured databases to answer complex diagnostic 
+and pharmacological queries with interpretable logical justifications \cite{Erdem2011}. 
+In the clinical domain, ASP is especially effective for encoding medical rules and patient-specific data, 
 enabling precise and explainable decision-making.
-Common strategies in ASP-based medical reasoning include:
 
-\begin{itemize}
-    \item \textbf{Rule-Based Representation}: ASP allows for the representation of medical 
-    knowledge as rules in the form of logical statements. 
-    These rules define relationships between medical entities such as symptoms, 
-    diseases, treatments, and patient data. For example, a rule could state that 
-    if a patient exhibits specific symptoms, it indicates the presence or absence 
-    of certain diseases.
-		There have been multiple instances of rule-based medical reasoning systems~\cite{CHEN_2016,marple2017s}
-    \item \textbf{Knowledge Base Construction}: ASP facilitates the construction of 
-    structured knowledge bases composed of facts, rules, and constraints. These can 
-    model diagnostic criteria, symptom-disease associations, and therapeutic protocols 
-		in a consistent and extensible manner~\cite{chen2018automating,CHEN_2016}.
-\end{itemize}
-\sys builds upon these works and identifies a key bottleneck, which is the construction of a
-knowledge base that can cover a wide range of diseases and conditions.
+Two common strategies illustrate the role of ASP in medical reasoning. 
+The first is \textit{rule-based representation}, where medical knowledge is expressed as logical statements 
+that capture relationships among symptoms, diseases, treatments, and patient data. 
+For instance, a rule might specify that if a patient exhibits certain symptoms, 
+this supports or excludes the presence of particular diseases. 
+The second is \textit{knowledge base construction}, where facts, rules, and constraints 
+are combined into structured repositories that consistently model diagnostic criteria, 
+symptom–disease associations, and therapeutic protocols in an extensible way. 
+Together, these approaches demonstrate how ASP provides a flexible yet rigorous framework 
+for supporting medical diagnosis, offering not only accurate predictions 
+but also transparent reasoning paths that can be trusted in high-stakes settings.\\
 
@@
 improving the automation of rule-based medical reasoning \cite{coppolillo2024llasp}.
-These systems assist in building reliable and clear reasoning pipelines 
+Collectively, these systems assist in building reliable and clear reasoning pipelines 
 for disease diagnosis by combining language understanding with rule-based validation.
-These systems were developed after \sys and illustrate the convergent development of similar approaches, highlighting the intuitive appeal and practical value of the underlying techniques.
-
 \section{Methodology}
@@
 
+\Cref{fig:architecture} illustrates the \sys architecture, 
+which transforms unstructured medical text into a structured knowledge base \cite{ré2014feature}. 
+The process begins with the collection of medical literature relevant to the diseases or conditions of interest. 
+A carefully designed prompt is then used to guide an LLM in translating this literature into ASP code, 
+which is incorporated into a growing knowledge base. 
+This knowledge base is further enriched with patient-specific information such as symptoms and examination results. 
+An ASP solver subsequently reasons over the combined knowledge base and produces a diagnosis
+--or a set of candidate diagnoses--in the form of answer sets.
+
 \begin{figure}[!h]
-\begin{center}
-\begin{tikzpicture}[node distance = 3cm, minimum width=2cm, minimum height=1cm]
-    \node [doc, label=above:{}] (med-bib) {Medical Bibliography};
-    \node [rect, draw, rounded corners=2pt, right=2cm of med-bib, fill=blue!30] (llm) {Large Language Model};
-    \node [rect, draw, above= 1cm of med-bib, fill=orange!50] (prompt) {Prompt};
-    \node [database, below left= 2.5cm of med-bib, fill=green!30] (knowledge-base) {Knowledge Base};
-    \node [rect, draw, right of=knowledge-base] (solver) {ASP Solver};
-    \node [rect, draw, below=.5cm of solver] (patient) {\textit{Patient Data}};
-    \node [rect, right of=solver] (diagnosis) {\textbf{Diagnosis}};
-
-    \draw[thick-arrow] (med-bib) edge [] node [above] {} (llm);
-    \draw[thick-arrow] (prompt) edge [in=90, out=0, looseness=1] node [above, sloped] {\textit{\say{please convert the following into ASP code...}}} (llm);
-    \draw[thick-arrow] (llm) edge [in=90, out=270, looseness=1] node [rect, draw] {ASP Program} (knowledge-base);
-    \draw[thick-arrow] (knowledge-base) edge [] node [above] {} (solver);
-    \draw[thick-arrow] (solver) edge [] node [above] {} (diagnosis);
-    \draw[thick-arrow] (patient) edge [in=270, out=180, looseness=1] node [above] {} (knowledge-base);
-
-\end{tikzpicture}
-\end{center}
+    % https://docs.google.com/drawings/d/1cr2lTIjSUxfQH54CZui8vGqLS2keOfW41i4g4uGfwhw/edit?usp=sharing
+    \includegraphics[width=\textwidth]{assets/mccoy.pdf}
     \caption{\textbf{\sys overview.} The diagnostic process begins by transforming medical literature 
@@
 
-The goal is to extract a structured knowledge base from unstructured medical
-text \cite{ré2014feature}. \Cref{fig:architecture} illustrates the proposed 
-architecture, which proceeds as follows. 
-First, medical literature relevant to the diseases or conditions of interest 
-is collected. Then, a carefully designed prompt is used to convert this literature 
-into ASP code via a large language model. 
-The generated ASP code is incorporated into a growing knowledge base, 
-which is further enriched with patient-specific data such as symptoms 
-and examination results. Finally, an ASP solver reasons over the complete 
-knowledge base to produce a diagnosis--or a set of possible diagnoses--in 
-the form of answer sets.
-
 \paragraph{\textbf{Prompt Construction}}
-
-While large language models demonstrate strong instruction-following capabilities, 
-their reliability remains limited in practice \cite{bowman2023things}. 
-To address this, a relatively recent methodology known as prompt engineering 
-has emerged \cite{liu2021pretrain, promptingguidePromptEngineering}. 
-In this work, prompts are carefully crafted to guide the model in 
-converting medical literature into ASP code suitable for execution 
-by a solver (\cref{fig:bad-prompt}, \cref{fig:good-prompt}).
+While large language models exhibit strong instruction-following capabilities, 
+their reliability in high-stakes domains remains limited \cite{bowman2023things}. 
+To improve robustness, a wide range of prompting strategies have been developed 
+and refined over the past several years \cite{liu2021pretrain, promptingguidePromptEngineering}. 
+In \sys, carefully designed prompts are employed to guide the model in translating medical literature 
+into ASP code suitable for execution by a solver (\cref{fig:bad-prompt}, \cref{fig:good-prompt}).
 \footnote{The example of a poorly constructed prompt is included for educational purposes, to highlight common pitfalls in prompt design.}
 
-% Ioanna: I commented out the bad prompt
 \begin{figure}[!h]
@@
 \begin{commandline}
+\footnotesize
 \begin{verbatim}
@@
 \begin{commandline}
-\scriptsize
+\footnotesize
 \begin{verbatim}
@@
 
-\paragraph{\textbf{Construction of a Logical Program}}
+\paragraph{\textbf{Logical Program Construction}}
 
@@
 relying as much as possible on available clinical data rather than hypothetical assumptions.
+\\
 
@@
 
-Symbolic artificial intelligence offers a significant advantage 
+Symbolic Artifical Intelligence offers a significant advantage 
 over subsymbolic methods in terms of explainability. 
@@
 This structure supports the interpretability of conclusions derived by a symbolic logic solver.
+% \begin{equation}
+% \begin{aligned}[b]
+%     l :& punish \longleftarrow drive, drunk \\
+%     m :& punish \longleftarrow resist \\
+%     e :& prison \longleftarrow punish \\
+%     d :& drive \\
+%     k :& drunk \\
+%     r :& resist \\
+% \end{aligned}
+% \label{eq:example-cg}
+% \end{equation}
+% % https://docs.google.com/drawings/d/1Ing4GFH2kbniiHI5rbXzw-xcAUQAxWAJpgmg4tz1fzc/edit?usp=sharing
+% \begin{figure}
+%     \centering
+%     \includegraphics{assets/causal_g.pdf}
+%     \caption{Causal graph of logic program (\cref{eq:example-cg}).}
+%     \label{fig:causal-g}
+% \end{figure}
+\begin{figure}[!h]
+    \centering
+    \begin{minipage}{0.48\textwidth}
     \begin{equation}
@@
     \end{equation}
-% https://docs.google.com/drawings/d/1Ing4GFH2kbniiHI5rbXzw-xcAUQAxWAJpgmg4tz1fzc/edit?usp=sharing
-\begin{figure}
+    \end{minipage}
+    \hfill
+    \begin{minipage}{0.48\textwidth}
     \centering
-    \includegraphics{assets/causal_g.pdf}
+    \includegraphics[width=\textwidth]{assets/causal_g.pdf}
     \caption{Causal graph of logic program (\cref{eq:example-cg}).}
     \label{fig:causal-g}
+    \end{minipage}
     \end{figure}
@@
 potential errors in the construction of the knowledge base and offers transparency for the final outcome.
-This is especially important in medical contexts, where interpretability and trust are critical.
 
@@
 and transitive symptom relationships encoded in the knowledge base to 
-construct a transparent and interpretable diagnostic path.
+construct a transparent and explainable diagnostic path.
 
@@
 
-This work presents a method for automatic disease prediction 
+This work presents \sys, a framework for automatic disease prediction 
 that encodes medical literature and transforms it into ASP code using an LLM. 
@@
 
-Future work includes integrating retrieval-augmented generation (RAG) to improve the factual accuracy 
-and reliability of the ASP code, and fine-tuning the LLM on clinical reasoning tasks to enhance rule quality. 
-Expanding the approach to cover additional diseases is also a key direction.
+Future work will focus on integrating retrieval-augmented generation (RAG) 
+to ground ASP code in reliable medical sources, fine-tuning LLMs on clinical reasoning tasks 
+to improve the quality and consistency of generated rules, 
+and extending the framework to additional diseases to demonstrate scalability in diverse diagnostic settings.
+
 \newpage

### 2025-09-12 paper/draft.tex
Update intro

@@
-\documentclass[10pt,leqno]{amsart}
+\documentclass[11pt,leqno]{amsart}
 \usepackage{graphicx}
@@
 \usepackage{indentfirst,csquotes}
+\usepackage[margin=1in]{geometry}
 
-\topmargin= .5cm
-\textheight= 20cm
-\textwidth= 32cc
-\baselineskip=16pt
+% \topmargin= .5cm
+% \textheight= 20cm
+% \textwidth= 32cc
+% \baselineskip=16pt
 
-\evensidemargin= .9cm
-\oddsidemargin= .9cm
+% \evensidemargin= .9cm
+% \oddsidemargin= .9cm
 
@@
 
-The development of automatic disease prediction systems 
-is crucial for enhancing healthcare and supporting medical 
-professionals in timely and accurate diagnosis. 
-Symbolic artificial intelligence (AI), particularly ASP, 
-has been explored in this context due to its ability 
-to support structured and explainable reasoning. 
-However, the construction of expressive knowledge bases 
-remains a significant barrier to practical adoption.
+Accurate and timely diagnosis is critical for effective treatment and patient outcomes, 
+but building automated systems that can deliver this reliably remains challenging. 
+Symbolic AI---and ASP in particular---offers transparent reasoning, 
+but adoption has been hindered by the high cost of 
+authoring and maintaining rich, domain-specific knowledge bases at scale. 
+Meanwhile, LLMs show strong performance on medical text 
+but often produce unreliable or unstructured outputs, limiting their use in 
+high-stakes clinical settings.
 
-This paper introduces \sys, a framework 
-that combines LLMs with the 
-formal reasoning capabilities of ASP.
-\sys extracts and encodes medical knowledge from text into ASP code,
-enabling interpretable and accurate diagnosis 
-of diseases based on patient symptoms.
+\sys addresses this gap by coupling LLMs with ASP to automatically construct 
+explainable diagnostic programs from medical literature. 
+\sys uses prompting to translate disease-specific literature into ASP rules, 
+augments these rules with patient symptoms and indicators, 
+and invokes an ASP solver to produce candidate diagnoses together with explicit logical justifications.
 
-\section{Background}
+Concretely, this work:
+\begin{itemize}
+    \item Builds a framework that turns medical text into an executable ASP knowledge base.
+    \item Uses prompts to produce modular rules (including alternatives and symptom links) for better explainability and partial inference.
+    \item Adds solver-based explanations to show clear symptom $\rightarrow$ rule $\rightarrow$ diagnosis chains.
+    \item Demonstrates accurate results on real patient data while keeping the reasoning transparent.
+\end{itemize}
+
+On publicly available patient records, \sys achieves 
+95–-100\% accuracy across selected diseases, while producing transparent explanations for each diagnosis.
+
+This paper begins with background on LLMs and ASP (\cref{sec:background}), 
+then presents the \sys methodology (\cref{sec:methodology}), 
+evaluates the framework on patient data (\cref{sec:results}),
+and concludes with limitations and future directions (\cref{sec:conclusion}).
+
+\section{Background}\label{sec:background}
 
@@
     \end{center}
-		\caption{Solving a problem using ASP.}
+		\caption{\textbf{Solving a problem with ASP}.
+A problem is first modeled as a program, then transformed into ground rules by a grounder.
+A solver computes the stable models, which are interpreted as solutions.}
     \label{fig:asp-solving}
@@
 for disease diagnosis by combining language understanding with rule-based validation.
-\section{Methodology}
+
+\section{Methodology} \label{sec:methodology}
 
@@
 
-\section{Results}
+\section{Results}\label{sec:results}
 
@@
 
-\section{Conclusion}
+\section{Conclusion}\label{sec:conclusion}
 

### 2025-09-18 paper/draft.tex
Update intro

--- 2025-09-12 paper/draft.tex
@@
 \section{Introduction}
+Getting an accurate diagnosis quickly can make all the difference in patient outcomes, 
+but building automated systems that practitioners can actually trust is still a major challenge. 
+Symbolic AI, and in particular Answer Set Programming (ASP), provides clear logical reasoning \cite{Brewka2011,Gelfond_Kahl_2014}. 
+However, its adoption has been limited because constructing and maintaining large, 
+domain-specific knowledge bases is very labor-intensive.
+On the other hand, Large language models (LLMs) achieve strong results on tasks
+such as medical question answering and clinical note summarization \cite{Singhal2022LargeLM,Singhal2023,Zheng2024LLMmed}, 
+but their tendency to produce inconsistent or opaque outputs makes them risky for clinical use \cite{aljohani2025comprehensivesurveytrustworthinesslarge, rudin2019stopexplainingblackbox}.
 
-The development of automatic disease prediction systems 
-is crucial for enhancing healthcare and supporting medical 
-professionals in timely and accurate diagnosis. 
-Symbolic artificial intelligence (AI), particularly ASP, 
-has been explored in this context due to its ability 
-to support structured and explainable reasoning. 
-However, the construction of expressive knowledge bases 
-remains a significant barrier to practical adoption.
+In clinical decision support, the real goal is to get the best of both worlds: 
+systems that reason accurately while still being easy for practitioners to trust and use. 
+\sys takes a step in that direction by combining the strengths of LLMs and ASP. 
+Instead of requiring experts to hand-craft large knowledge bases, 
+\sys automatically converts medical literature into ASP rules. 
+These rules are then enriched with patient symptoms and clinical indicators, 
+and passed through an ASP solver to produce diagnostic suggestions together with logical justifications.
 
-This paper introduces \sys, a framework 
-that combines LLMs with the 
-formal reasoning capabilities of ASP.
-\sys extracts and encodes medical knowledge from text into ASP code,
-enabling interpretable and accurate diagnosis 
-of diseases based on patient symptoms.
+Concretely, this work:
+\begin{itemize}
+    \item Introduces a framework that turns medical text into an executable ASP knowledge base.
+    \item Uses prompts to generate modular rules (including alternatives and symptom links) that the solver applies for reasoning.
+    \item Provides solver-based explanations showing how symptoms lead to rules and then to diagnoses.
+    \item Demonstrates accurate results on real patient data while keeping the reasoning fully transparent.
+\end{itemize}
 
-\section{Background}
+On public patient records, \sys reaches 95–100\% predictive accuracy on selected diseases, 
+while giving transparent explanations for each diagnosis.
+
+The paper begins with background on LLMs and ASP (\cref{sec:background}), 
+then details the \sys methodology (\cref{sec:methodology}), 
+evaluates the approach on patient data (\cref{sec:results}), 
+and concludes with a discussion of limitations and future directions (\cref{sec:conclusion}).
+
+\section{Background}\label{sec:background}
 

### 2025-09-18 paper/draft.tex
Updates in intro

@@
 
-This work presents \sys, a framework for automatic disease prediction 
-that encodes medical literature and transforms it into ASP code using an LLM. 
-It describes the roles of both LLMs and ASP, emphasizing how each contributes to the overall framework. 
-The \sys architecture is outlined, along with implementation details for constructing effective prompts 
-and structuring rules within the ASP knowledge base.
-The method includes predictions for a set of diseases based on the constructed knowledge bases. 
-Results show that this approach achieves high-accuracy predictions and provides interpretable explanations.
+This paper presents \sys, a framework for disease prediction using LLMs and ASP. 
+The system uses LLMs to extract knowledge from medical texts and translate it into ASP rules, 
+which then drive predictions and provide interpretable explanations. 
+The architecture, prompt design, and rule structuring in the knowledge base are described.
+Experiments show that \sys can predict diseases accurately 
+while keeping the reasoning process transparent.
 
-Future work will focus on integrating retrieval-augmented generation (RAG) 
-to ground ASP code in reliable medical sources, fine-tuning LLMs on clinical reasoning tasks 
-to improve the quality and consistency of generated rules, 
-and extending the framework to additional diseases to demonstrate scalability in diverse diagnostic settings.
-
+Future work will focus on grounding ASP rules in reliable sources 
+through retrieval-augmented generation, fine-tuning LLMs for clinical reasoning, 
+and scaling the framework to cover a wider range of diseases.
 \newpage

### 2025-09-19 paper/draft.tex
Small tweaks
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

@@
 and passed through an ASP solver to produce diagnostic suggestions together with logical justifications.
-
-Concretely, this work:
-\begin{itemize}
-    \item Introduces a framework that turns medical text into an executable ASP knowledge base.
-    \item Uses prompts to generate modular rules (including alternatives and symptom links) that the solver applies for reasoning.
-    \item Provides solver-based explanations showing how symptoms lead to rules and then to diagnoses.
-    \item Demonstrates accurate results on real patient data while keeping the reasoning fully transparent.
-\end{itemize}
-
 On public patient records, \sys reaches 95–100\% predictive accuracy on selected diseases, 
@@
 
+% Concretely, this work:
+% \begin{itemize}
+%     \item Introduces a framework that combines the flexibility of LLMs with the explanability of ASP for disease diagnosis.
+%     \item Provides solver-based explanations showing how symptoms lead to rules and then to diagnoses.
+%     \item Demonstrates accurate results on real patient data while keeping the reasoning fully transparent.
+% \end{itemize}
+
 The paper begins with background on LLMs and ASP (\cref{sec:background}), 
-then details the \sys methodology (\cref{sec:methodology}), 
+then details the \sys framework (\cref{sec:methodology}), 
 evaluates the approach on patient data (\cref{sec:results}), 
@@
 
-\begin{figure}[htb]
+\begin{figure}[t]
     \begin{center}
@@
 
-\begin{figure}[!h]
+\begin{figure}[t]
     % https://docs.google.com/drawings/d/1cr2lTIjSUxfQH54CZui8vGqLS2keOfW41i4g4uGfwhw/edit?usp=sharing
@@
 % \end{figure}
-\begin{figure}[!h]
+\begin{figure}[t]
     \centering
@@
     \hfill
-    \begin{minipage}{0.48\textwidth}
+    \begin{minipage}{0.45\textwidth}
     \centering
-    \includegraphics[width=\textwidth]{assets/causal_g.pdf}
+    \includegraphics[width=.5\textwidth]{assets/causal_g.pdf}
     \caption{Causal graph of logic program (\cref{eq:example-cg}).}
@@
 and scaling the framework to cover a wider range of diseases.
-\newpage
 \bibliographystyle{unsrt}

### 2025-12-30 paper/draft.tex
Fixes
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

@@
 \usepackage{enumerate}
-\usepackage{graphicx}
 \usepackage{caption}
@@
 \usepackage{subcaption}
-\usepackage{amsmath}
 \usepackage{array,tabularx}
@@
 \usepackage{tikz}
-\usepackage{array}
 \usepackage[framemethod=tikz]{mdframed} % Allows defining custom boxed/framed environments
@@
 \usepackage{microtype}
-\usepackage{dirtytalk}
 \usepackage{booktabs}
@@
 \usetikzlibrary{matrix}
+
+\newcommand{\eg}{{e.g.}, }
+\newcommand{\ie}{{i.e.}, }
+\newcommand{\etc}{{etc.}\xspace}
+\newcommand{\vs}{{vs.} }
+\newcommand{\cf}[1]{(\emph{Cf}.\S\ref{#1})}
+\newcommand{\sx}[1]{(\cref{#1})}
 
@@
 		\fill[black,rotate around={45:(P-|O)}](tmp1)rectangle(tmp2);
-		\node at(P-|O){\color{white}\scriptsize\bf ?};
+		\node at(P-|O){\color{white}\scriptsize\bf !};
 		\draw[very thick](P-|O)++(0,-1em)--(O);%--(O-|P);
@@
 
-\title[\sys: Automatic Disease Diagnosis Using LLMs \& ASP]{\sys: Automatic Disease Diagnosis Using Large Language Models and Answer Set Programming}
+\title[\sys: A PoC for Explainable Disease Diagnosis Using LLMs \& ASP]{\sys: A Proof-of-Concept for Explainable Disease Diagnosis Using Large Language Models and Answer Set Programming}
 
@@
 \author[Evangelos Lamprou]{Evangelos Lamprou}
-\thanks{Originally submitted as part of a research competition on 1st June 2023.}
+\thanks{Originally submitted as part of a university research competition on 1st June 2023.}
 
@@
 \maketitle
-\vspace{-1.5em} % adjust as needed
-{\small\itshape \centering Originally submitted as part of a university research competition on 1st June 2023.\par}
-\vspace{0.5em} % optional: add space before body
 
@@
 that leverages the strengths of both paradigms. 
-\sys shows strong performance on real patient data.
+\sys shows strong performance on public datasets.
 % \sys accurately predicts diseases based on symptoms derived from medical literature.
@@
 domain-specific knowledge bases is very labor-intensive.
-On the other hand, Large language models (LLMs) achieve strong results on tasks
+On the other hand, Large Language Models (LLMs) achieve strong results on tasks
 such as medical question answering and clinical note summarization \cite{Singhal2022LargeLM,Singhal2023,Zheng2024LLMmed}, 
-but their tendency to produce inconsistent or opaque outputs makes them risky for clinical use \cite{aljohani2025comprehensivesurveytrustworthinesslarge, rudin2019stopexplainingblackbox}.
+but their tendency to produce inconsistent or opaque outputs makes them risky for clinical use~\cite{aljohani2025comprehensivesurveytrustworthinesslarge, rudin2019stopexplainingblackbox}.
 
@@
 and passed through an ASP solver to produce diagnostic suggestions together with logical justifications.
-On public patient records, \sys reaches 95–100\% predictive accuracy on selected diseases, 
+On public datasets, \sys reaches 95–100\% predictive accuracy on selected diseases, 
 while giving transparent explanations for each diagnosis.
@@
 these models can produce text that closely mirrors human linguistic patterns.
-Modern architectures such as \textit{GPT-5} \cite{openai2023gpt4, gpt5}, 
+Modern architectures such as \textit{GPT-5}~\cite{gpt5}
 are trained on massive corpora spanning diverse domains, 
@@
 LLMs use deep learning algorithms \cite{Sarker2021} 
-and specifically transformer networks \cite{Dosovitskiy2020} 
+and specifically transformer networks \cite{Vaswani2017} 
 to analyze and understand the relationships between words, phrases, and sentences. 
@@
 they can generate coherent and relevant responses based on given prompts.
-The term “large” refers to the billions of parameters 
+The term ``large" often refers to models with billions of parameters 
 that allow these models to generate context-aware responses. 
@@
 making them powerful tools for a wide range of natural language processing tasks. 
-This characteristic is particurarly relevant to the present work, 
+This characteristic is particularly relevant to the present work, 
 which focuses on the effective encoding of medical literature. \\
@@
 were first introduced by Gelfond and Lifschitz \cite{Gelfond2000, gel88}.
-Programming with this approach is done in a family of languages 
-sometimes referred to as \textit{AnsProlog} \cite{Gelfond2002}.
+These are implemented in a family of languages commonly known as \textit{AnsProlog} \cite{Gelfond2002}.
 
@@
 First, the problem is modeled using the ASP language. 
-Next, a grounder (e.g., gringo \cite{Gebser2014}) transforms the program into a set of basic rules and facts. 
-Finally, a solver (e.g., clasp \cite{Holldobler2014}) computes the stable models (answer sets), 
+Next, a grounder (\eg gringo \cite{Gebser2014}) transforms the program into a set of basic rules and facts. 
+Finally, a solver (\eg clasp \cite{Holldobler2014}) computes the stable models (answer sets), 
 which correspond to the solutions of the original problem.
@@
 
-        \draw [arrow] (problem) -- node[right] {\textbf{Modelling}} (program);
+        \draw [arrow] (problem) -- node[right] {\textbf{Modeling}} (program);
         \draw [arrow] (stable-models) -- node[left] {Interpreting} (solution);
@@
 Several solvers for ASP have been developed, such as DLV \cite{Xia2020}. 
-This work uses the ASP system clingo, 
+This work uses the ASP system \src{clingo}, 
 which integrates the gringo grounder \cite{Gebser2014} with the clasp solver \cite{Holldobler2014} into a single system. 
-Clingo also offers a powerful Python API, enabling the solver's integration into other applications.
+The Clingo system also offers a powerful Python API, enabling integration into other applications.
 
@@
 This knowledge base is further enriched with patient-specific information such as symptoms and examination results. 
-An ASP solver subsequently reasons over the combined knowledge base and produces a diagnosis
---or a set of candidate diagnoses--in the form of answer sets.
+An ASP solver subsequently reasons over the combined knowledge base and produces a diagnosis, or a set of plausible diagnoses.
 
@@
 \paragraph{\textbf{Prompt Construction}}
-While large language models exhibit strong instruction-following capabilities, 
+While Large Language Models exhibit strong instruction-following capabilities, 
 their reliability in high-stakes domains remains limited \cite{bowman2023things}. 
@@
 The paragraph above lists common symptoms of pneumonia.
-Write a Clingo script that diagnoses pneumonia based on these symptoms.
+Write a clingo script that diagnoses pneumonia based on these symptoms.
 \end{prompt}
@@
 The paragraph above lists common symptoms of pneumonia.
-Write a Clingo script that diagnoses pneumonia based on these symptoms.
-
+Write a clingo script that diagnoses pneumonia based on these symptoms.
 In the diagnosis rule, use a structure like:
@@
 Include alternative diagnoses that share overlapping symptoms with pneumonia.
-Add rules that link one symptom to another (e.g., symptom propagation or dependency).
+Add rules that link one symptom to another (\eg symptom propagation or dependency).
 
@@
 \end{align}
+Let $S$ denote the set of symptoms and $I$ the set of indicators.
 
-A patient exhibiting a specific symptom or testing positive for a particular indicator is represented as:
+A patient exhibiting a specific symptom or testing positive for an indicator is represented as:
 \begin{equation}
@@
 
-To ensure that the solver identifies at least one diagnosis, 
-the following integrity constraint is imposed:
+The following integrity constraint imposes at least one diagnosis:
 \begin{equation}
@@
 \begin{equation}
-    \#minimize \{ 1, S : add(S) \}
+    \#minimize \{ 1, S : add(symptom(S)) \}
 \end{equation}
@@
 
-Symbolic Artifical Intelligence offers a significant advantage 
+Symbolic Artificial Intelligence offers a significant advantage 
 over subsymbolic methods in terms of explainability. 
 In symbolic systems, the relationships between facts and conclusions are explicitly defined, 
-enabling transparent reasoning. However, ASP solvers such as \textit{Clingo} do not, 
-at the time of writing,\footnote{At the time of writing this paper.} 
-provide intuitive traces of their reasoning processes. 
-Prior work has explored methods for associating logical propositions 
+enabling transparent reasoning. Prior work has explored methods for associating logical propositions 
 with their justifications \cite{cabalar2014causal}. For example, 
@@
     \includegraphics[width=.5\textwidth]{assets/causal_g.pdf}
-    \caption{Causal graph of logic program (\cref{eq:example-cg}).}
+    \caption{Causal graph of logic program in \cref{eq:example-cg}.}
     \label{fig:causal-g}
@@
 
-The tool \textit{xclingo}~\cite{Cabalar_2020} 
-enables the visualization of justifications for results produced by the \textit{Clingo} solver. 
-It does so by incorporating a \src{trace} that corresponds to the rules of the logic program. 
-Providing clear explanations of how the framework reaches specific conclusions helps users identify 
-potential errors in the construction of the knowledge base and offers transparency for the final outcome.
+The tool \src{xclingo}~\cite{Cabalar_2020} 
+enables the visualization of justifications for results produced by the \src{clingo} solver. 
+It does so by incorporating a trace that corresponds to the rules of the logic program. 
+This can help identify potential flaws in \sys's knowledge base and reasoning process.
 
@@
 
-Evaluation was conducted using publicly available patient data.\footnote{\url{https://www.kaggle.com/datasets/itachi9604/disease-symptom-description-dataset?select=dataset.csv}}
-Each patient's list of symptoms serves as input.  
+Evaluation was conducted using a publicly available dataset.\footnote{\url{https://www.kaggle.com/datasets/itachi9604/disease-symptom-description-dataset?select=dataset.csv}}
+Each record's list of symptoms serves as input.  
 The knowledge base—comprising ASP rules automatically derived  
@@
     \midrule
-    Chicken Pox    &  66  & 95\% \\
+    Chickenpox     &  66  & 95\% \\
     Pneumonia      &  75  & 100\% \\
@@
 and scaling the framework to cover a wider range of diseases.
+\newpage
+
 \bibliographystyle{unsrt}

### 2025-12-30 paper/draft.tex
Update title
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

@@
 
-\title[\sys: A PoC for Explainable Disease Diagnosis Using LLMs \& ASP]{\sys: A Proof-of-Concept for Explainable Disease Diagnosis Using Large Language Models and Answer Set Programming}
+\title[A PoC for Explainable Disease Diagnosis Using LLMs \& ASP]{A Proof-of-Concept for Explainable Disease Diagnosis\\ Using Large Language Models and\\ Answer Set Programming}
 

### 2025-12-30 paper/draft.tex
Prepare for arxiv
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

@@
 \author[Evangelos Lamprou]{Evangelos Lamprou}
-\thanks{Originally submitted as part of a university research competition on 1st June 2023.}
+\thanks{Originally submitted as part of a university research competition on 1st June 2023. Both authors were students at the University of Patras, Greece, at the time.}
 
@@
 While symbolic AI has been applied in healthcare, 
-its adoption remains limited due to the complexity 
-of required knowledge bases.
-This work introduces \sys, a framework that combines Large Language Models (LLMs) 
-with Answer Set Programming (ASP) to overcome this barrier. 
-LLMs extract medical knowledge from text and translate it into ASP code, 
-while ASP enables structured, explainable reasoning. 
-This integration yields a robust, interpretable prediction framework 
-that leverages the strengths of both paradigms. 
-\sys shows strong performance on public datasets.
+its adoption remains limited due to the effort required for constructing high-quality knowledge bases.
+This work introduces \sys, a framework that combines Large Language Models (LLMs) with Answer Set Programming (ASP) to overcome this barrier. 
+\sys orchestrates an LLM to translate medical literature into ASP code, combines it with patient data, and processes it using an ASP solver to arrive at the final diagnosis.
+This integration yields a robust, interpretable prediction framework that leverages the strengths of both paradigms. 
+Preliminary results show \sys has strong performance on small-scale disease diagnosis tasks.
 % \sys accurately predicts diseases based on symptoms derived from medical literature.
@@
 Getting an accurate diagnosis quickly can make all the difference in patient outcomes, 
-but building automated systems that practitioners can actually trust is still a major challenge. 
-Symbolic AI, and in particular Answer Set Programming (ASP), provides clear logical reasoning \cite{Brewka2011,Gelfond_Kahl_2014}. 
+but building automated systems that practitioners can trust is still a major challenge~\cite{Vinarti2019}.
+Symbolic AI, and in particular Answer Set Programming (ASP), provides clear logical reasoning~\cite{Brewka2011,Gelfond_Kahl_2014}
+and has seen success in the medical domain~\cite{Alviano_2020}.
 However, its adoption has been limited because constructing and maintaining large, 
 domain-specific knowledge bases is very labor-intensive.
-On the other hand, Large Language Models (LLMs) achieve strong results on tasks
+On the other hand, Large Language Models (LLMs) have shown strong results on tasks
 such as medical question answering and clinical note summarization \cite{Singhal2022LargeLM,Singhal2023,Zheng2024LLMmed}, 
-but their tendency to produce inconsistent or opaque outputs makes them risky for clinical use~\cite{aljohani2025comprehensivesurveytrustworthinesslarge, rudin2019stopexplainingblackbox}.
+but their tendency to produce inconsistent or opaque outputs makes them risky for clinical use~\cite{aljohani2025comprehensivesurveytrustworthinesslarge,rudin2019stopexplainingblackbox,Bommasani2021}.
 
-In clinical decision support, the real goal is to get the best of both worlds: 
-systems that reason accurately while still being easy for practitioners to trust and use. 
+There is an acute need for clinical support systems that are both interpretable and widely applicable.
 \sys takes a step in that direction by combining the strengths of LLMs and ASP. 
@@
 \paragraph{\textbf{Large Language Models}}
-LLMs \cite{zhao2023survey} are advanced computational systems 
-designed to understand and generate human language. 
-Developed using cutting-edge machine learning techniques, 
-these models can produce text that closely mirrors human linguistic patterns.
-Modern architectures such as \textit{GPT-5}~\cite{gpt5}
-are trained on massive corpora spanning diverse domains, 
+LLMs are advanced computational systems 
+designed to understand and generate human language~\cite{zhao2023survey}. 
+These models can produce text that closely mirrors human linguistic patterns and have emergent reasoning capabilities~\cite{Paranjape2023}.
+Recent models such as \textit{GPT-5}~\cite{gpt5}
+have been trained on massive corpora spanning diverse domains, 
 enabling them to acquire broad, cross-disciplinary knowledge. 
-LLMs use deep learning algorithms \cite{Sarker2021} 
-and specifically transformer networks \cite{Vaswani2017} 
-to analyze and understand the relationships between words, phrases, and sentences. 
+To give an intuition of the way they operate, LLMs are deep neural networks that have converged to transformer networks~\cite{Vaswani2017},
+which are trained using self-supervised learning~\cite{Gui2024surveySSL}.
+During training, these models refine an internal representation that enables them to analyze and understand the relationships between words, phrases, and sentences. 
 By capturing patterns and structures in the training data, 
-they can generate coherent and relevant responses based on given prompts.
-The term ``large" often refers to models with billions of parameters 
-that allow these models to generate context-aware responses. 
-The basic premise of an LLM is its ability to predict 
-the next word (referred to as a token) based on the text it has observed so far, 
-according to the context.
+they can then be probed to generate coherent and relevant responses based on given prompts.
+The term ``large" often refers to models with a large number of parameters (often in the billions)
+that allow these models to build high-quality representations of language.
+At a very basic level, an LLM is able to predict the next word (referred to as a token) based on its preceding context.
 
@@
 including question answering \cite{Brown2020}, summarization \cite{Zhang2020}, 
-dialogue generation \cite{Thoppilan2022lamda}, and code synthesis \cite{Chen2021}. 
+dialogue generation \cite{Thoppilan2022lamda}, and program synthesis \cite{Chen2021}. 
 In specialized domains such as biomedicine, domain-adapted LLMs have been employed 
@@
 biomedical knowledge retrieval \cite{Luo2022}.
-Among their capabilities, LLMs excel at transforming text from one representation to another, 
-making them powerful tools for a wide range of natural language processing tasks. 
+Among their capabilities, LLMs excel at transforming text from one representation to another~\cite{Vaswani2017,Li2024LLMDataProcessing}.
 This characteristic is particularly relevant to the present work, 
@@
 
-LLMs are not without errors \cite{Raj2023, Ruis2023}. 
-They can occasionally produce incorrect or biased responses 
-that reflect the biases inherent in the training data. 
+However, LLMs are not without errors~\cite{Raj2023, Ruis2023}. 
+They can occasionally produce incorrect or nonsensical responses 
+\cite{Huang2025SurveyHallucinationLLM}.
 There have been efforts to mitigate these issues 
@@
 Ongoing research seeks to improve the factual accuracy, safety, and interpretability 
-of these models, enabling their deployment in high-stakes environments \cite{Ganguli2022}. \\
+of these models, enabling their deployment in high-stakes environments \cite{Ganguli2022}.
+This work specifically tackles this issue by grounding LLM outputs on real medical knowledge, and having an ASP solver provide further validation and explainability.
+\\
 
@@
 The formal semantics of stable models and the foundational ASP language 
-were first introduced by Gelfond and Lifschitz \cite{Gelfond2000, gel88}.
+were first introduced by Gelfond and Lifschitz \cite{gel88}.
 These are implemented in a family of languages commonly known as \textit{AnsProlog} \cite{Gelfond2002}.
@@
 These elements collectively form a program that describes the conditions any solution must satisfy. 
-Instead of writing an algorithm to solve the problem, 
-the programmer encodes the specification of valid solutions. 
-An ASP solver then processes this program to compute one or more stable models (also known as answer sets),
+% Instead of writing an algorithm to solve the problem, 
+% the programmer encodes the specification of valid solutions. 
+Given a valid program, an ASP solver processes it to compute one or more stable models (also known as answer sets),
 each of which corresponds to a solution that satisfies all the given rules and constraints.
@@
 requires the programmer's thorough understanding of the given problem, 
-followed by the creation of a program that will produce the correct output 
-when provided with an instance of the problem, which will then be interpreted as the solution. 
+followed by the creation of a program that will take as input an instance of the problem,
+and produce output which will then be interpreted as the solution. 
 In ASP, the process of deriving solutions from a problem specification 
@@
 
-Several solvers for ASP have been developed, such as DLV \cite{Xia2020}. 
-This work uses the ASP system \src{clingo}, 
-which integrates the gringo grounder \cite{Gebser2014} with the clasp solver \cite{Holldobler2014} into a single system. 
+Several solvers for ASP have been developed, with some seeing industrial adoption~\cite{Xia2020,Gebser2014}.
+This work uses the Clingo ASP system,
+which integrates the \src{gringo} grounder \cite{Gebser2014} and the \src{clasp} solver \cite{Holldobler2014} into a single pipeline.
 The Clingo system also offers a powerful Python API, enabling integration into other applications.
@@
 \paragraph{\textbf{ASP for disease diagnosis}}
-ASP has been used to represent and reason over medical knowledge in a declarative, logic-based manner. 
+ASP has been used in the past to represent and reason over medical knowledge in a declarative, logic-based manner. 
 Early work applied ASP to biomedical ontologies and structured databases to answer complex diagnostic 
@@
 In the clinical domain, ASP is especially effective for encoding medical rules and patient-specific data, 
-enabling precise and explainable decision-making.
+enabling precise and explainable decision-making~\cite{Alviano_2020}
 
@@
 symptom–disease associations, and therapeutic protocols in an extensible way. 
-Together, these approaches demonstrate how ASP provides a flexible yet rigorous framework 
-for supporting medical diagnosis, offering not only accurate predictions 
-but also transparent reasoning paths that can be trusted in high-stakes settings.\\
+This work builds on these foundations and works towards closing the gap with fully automated knowledge-base construction.
+% Together, these approaches demonstrate how ASP provides a flexible yet rigorous framework 
+% for supporting medical diagnosis, offering not only accurate predictions 
+% but also transparent reasoning paths that can be trusted in high-stakes settings.
+\\
 
@@
 
-More recent work explores the synergy between LLMs and ASP 
+There has been work that explores the synergy between LLMs and ASP 
 for enhanced diagnostic reasoning. 
@@
 from LLMs in medical contexts \cite{Nguyen2025}. 
-A specialized lightweight LLM has also been trained to generate valid ASP code, 
-improving the automation of rule-based medical reasoning \cite{coppolillo2024llasp}.
-Collectively, these systems assist in building reliable and clear reasoning pipelines 
-for disease diagnosis by combining language understanding with rule-based validation.
+Another work leverages smaller-sized LLMs and uses fine-turning to enhance their performance in generating valid ASP programs~\cite{coppolillo2024llasp}.
+% Collectively, these systems assist in building reliable and clear reasoning pipelines by combining language understanding with rule-based validation.
+These works share several of the same insights and motivations as \sys.
 
@@
 
-\Cref{fig:architecture} illustrates the \sys architecture, 
-which transforms unstructured medical text into a structured knowledge base \cite{ré2014feature}. 
+\Cref{fig:architecture} illustrates \sys's architecture, 
+which transforms unstructured medical text into a structured knowledge base. 
 The process begins with the collection of medical literature relevant to the diseases or conditions of interest. 
@@
 This knowledge base is further enriched with patient-specific information such as symptoms and examination results. 
-An ASP solver subsequently reasons over the combined knowledge base and produces a diagnosis, or a set of plausible diagnoses.
+An ASP solver subsequently reasons over the combined knowledge base and produces a set of plausible diagnoses.
 
@@
 their reliability in high-stakes domains remains limited \cite{bowman2023things}. 
-To improve robustness, a wide range of prompting strategies have been developed 
-and refined over the past several years \cite{liu2021pretrain, promptingguidePromptEngineering}. 
-In \sys, carefully designed prompts are employed to guide the model in translating medical literature 
-into ASP code suitable for execution by a solver (\cref{fig:bad-prompt}, \cref{fig:good-prompt}).
-\footnote{The example of a poorly constructed prompt is included for educational purposes, to highlight common pitfalls in prompt design.}
+To improve robustness, a wide range of prompting strategies are employed by \sys~\cite{liu2021pretrain, promptingguidePromptEngineering}.
+In \sys, carefully designed prompts are employed to guide the model in translating medical literature.
+This involves a careful and accurate description of the desired output format, and specific instructions on the kind of information to encode within it.
+\Cref{fig:bad-prompt} represents an example of a poorly constructed prompt that fails to yield satisfactory results,
+and \cref{fig:good-prompt} shows the prompt that \sys ultimately uses to convert medical text into ASP code.
 
@@
 Let $S$ denote the set of symptoms and $I$ the set of indicators.
-
 A patient exhibiting a specific symptom or testing positive for an indicator is represented as:
@@
 
-Symbolic Artificial Intelligence offers a significant advantage 
-over subsymbolic methods in terms of explainability. 
+Symbolic AI offers a significant advantage 
+over sub-symbolic methods in terms of explainability. 
 In symbolic systems, the relationships between facts and conclusions are explicitly defined, 
-enabling transparent reasoning. Prior work has explored methods for associating logical propositions 
-with their justifications \cite{cabalar2014causal}. For example, 
-the logic program shown below is accompanied by a causal graph 
-in \cref{fig:causal-g}, which illustrates the connections 
+enabling transparent reasoning.
+Thus, it possible to associate logical propositions with their justifications~\cite{cabalar2014causal}.
+For example, 
+the logic program shown in \cref{eq:example-cg} is accompanied by a causal graph 
+shown in \cref{fig:causal-g}, which illustrates the connections 
 between rules and their inferred outcomes. 
 This structure supports the interpretability of conclusions derived by a symbolic logic solver.
+The tool \src{xclingo}~\cite{Cabalar_2020} 
+enables the visualization of justifications for results produced by the \src{clingo} solver. 
+It does so by incorporating a trace that corresponds to the rules of the logic program. 
+\sys leverages this produce such a trace, which can then help identify the reasoning path, as well as potential flaws in \sys's knowledge base and reasoning process.
+
 % \begin{equation}
@@
 
-The tool \src{xclingo}~\cite{Cabalar_2020} 
-enables the visualization of justifications for results produced by the \src{clingo} solver. 
-It does so by incorporating a trace that corresponds to the rules of the logic program. 
-This can help identify potential flaws in \sys's knowledge base and reasoning process.
-
 \section{Results}\label{sec:results}
 
-Evaluation was conducted using a publicly available dataset.\footnote{\url{https://www.kaggle.com/datasets/itachi9604/disease-symptom-description-dataset?select=dataset.csv}}
+Evaluation was conducted using a publicly available dataset.\footnote{\url{https://www.kaggle.com/datasets/itachi9604/disease-symptom-description-dataset}}
 Each record's list of symptoms serves as input.  
@@
 High accuracy is achieved when the knowledge base includes a sufficiently 
-rich set of logical rules. The size of this base, measured as the number 
-of ASP terms associated with a disease, correlates with prediction performance.
-
-In addition to accurate predictions, \sys offers explainability 
-through integration with \src{xclingo}. 
-This tool visualizes the reasoning process behind each diagnosis, 
-revealing the logical path from symptoms to conclusion.
+rich set of logical rules.
+% The size of this base, measured as the number of ASP terms associated with a disease, correlates with prediction performance.
+In addition to accurate predictions, \sys offers explainability and visualizes the reasoning process behind each diagnosis~(an example is shown in \cref{lst:chickenpox-explanation}).
 
@@
 
-Future work will focus on grounding ASP rules in reliable sources 
-through retrieval-augmented generation, fine-tuning LLMs for clinical reasoning, 
-and scaling the framework to cover a wider range of diseases.
+Future work will focus on scaling this process across more sources, 
+formalizing the pipeline to enable auditability and pluggability,
+and further experimentation in augmenting each step with additional guardrails, possibly including fine-tuning.
 \newpage
